### 2022-10-11 script/rembrant_description.Rmd
update script

---
title: "Rembrant"
author: "Laura Villain, Rodolphe Thiébaut, Thomas Ferte, Boris Hejblum"
output: 
  bookdown::html_document2:
    toc: yes
    toc_float: true
    number_sections: FALSE
    fig_caption: TRUE
    table_caption: TRUE
---

```{r include=FALSE}
 knitr::opts_chunk$set(cache=FALSE)

```

```{r setup, include=FALSE, eval=TRUE,message=FALSE,warning=FALSE,cache=FALSE}

library(knitr)
library(readxl)
library(survival)
library("survminer")
library(randomForestSRC)
library(dearseq)
library(limma)
library(xlsx)
library(GSA)
library(verification)
library(dynpred)
library(pec)
library(DESeq2)
library(base)
library(reshape2)
library(ggplot2)
library(stats)
library(patchwork)
library(blandr)
library(kableExtra)
library(dplyr)
library(FactoMineR)
library(scales)
library(ggrepel)
library(viridis)

```

```{r,eval=T,echo=F,message=FALSE,warning=FALSE,cache=FALSE}

datas_tot=readRDS("data/dfGenesAndClinical.rds")
datas_tot=datas_tot[,-c(1,3,5,11:34)]
datas_tot=datas_tot[,-c(2,8)]
datas_tot=datas_tot[!is.na(datas_tot[,4])&!is.na(datas_tot[,5]),]
datas_tot=datas_tot[datas_tot[,6]%in%c("ASTROCYTOMA","GBM","OLIGODENDROGLIOMA"),]

for (i in 1:length(datas_tot[,1])){
  if (datas_tot[i,6]=="ASTROCYTOMA"){
    datas_tot[i,6]="ASTRO"
  }
    if (datas_tot[i,6]=="OLIGODENDROGLIOMA"){
    datas_tot[i,6]="OLIGO"
    }

}

Age=datas_tot[,2]
Sex=datas_tot[,3]
for (i in 1:length(Age)){
  if (!is.na(Age[i])){
    if (Age[i]%in%c("15-19","20-24","25-29")){
      Age[i]="15-29"
    }
    if (Age[i]%in%c("30-34","35-39")){
      Age[i]="30-39"
    }
    if (Age[i]%in%c("40-44","45-49")){
      Age[i]="40-49"
    }
    if (Age[i]%in%c("50-54","55-59")){
      Age[i]="50-59"
    }
    if (Age[i]%in%c("60-64","65-69")){
      Age[i]="60-69"
    }
    if (Age[i]%in%c("70-74","75-79")){
      Age[i]="70-89"
    }
    if (Age[i]%in%c("80-84","85-89")){
      Age[i]="70-89"
    }
    if (Age[i]==""){
      Age[i]=NA
    }
  }
  if (Sex[i]%in%c(""," ")){
    Sex[i]=NA
  }
}

datas_tot[,2]=Age
datas_tot[,3]=Sex

#datas_tot=datas_tot[datas_tot[,6]=="GBM",]
datas_tot=datas_tot[!is.na(datas_tot[,2])&!is.na(datas_tot[,3]),]
datas=datas_tot[,1:6]
counts=datas_tot[,-(1:6)]
```

We can look at the boxplot of the counts per patients, classified per type of disease

```{r,eval=T,echo=F,message=FALSE,warning=FALSE,cache=FALSE}

data_histo=cbind(datas[,1],datas[,6],counts)
colnames(data_histo)=c("ID","Type",colnames(counts))
data_histo=data_histo[order(data_histo[,2]),]
data_histo[,1]=factor(data_histo[,1],levels=data_histo[,1])
df.m <- melt(data_histo, id=c("ID","Type"))
plot1=ggplot(data = df.m, aes(x=ID, y=value)) + geom_boxplot(aes(fill=Type),outlier.shape = NA,lwd=0.1, alpha=1)+theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+ 
  scale_y_continuous(limits = quantile(df.m$value, c(0.1, 0.9)))+xlab("patients")+ylab("counts")

```

```{r,eval=T,echo=F,message=FALSE,warning=FALSE,cache=FALSE, fig.cap="Counts for each patients"}
plot1
```

Then, to see if there are any outliers, we performed a pca on the counts (after scaling the data)

```{r,eval=T,echo=F,message=FALSE,warning=FALSE,results=FALSE, fig.show='hide'}
datas=datas
counts_pca <- t(counts)
#counts_pca <- round((counts_pca+1),0)
#counts_pca <- vst(t(counts_pca))

counts_pca_scaled <- base::scale(t(counts_pca), center=TRUE, scale=TRUE)

res.pca=PCA(counts_pca_scaled , graph=FALSE)
#plot.PCA(res.pca, axes=c(1, 2), choix="ind")
Type=as.factor(datas[,6])
pca_data <- cbind.data.frame(res.pca$ind$coord[, 1:4],as.factor(datas[,6]),rownames(res.pca$ind$coord))
colnames(pca_data)=c("Dim.1","Dim.2","Dim.3","Dim.4","Type","ID")
table_coord=res.pca$ind$coord[,c(1,2)]
particuliers=table_coord[table_coord[,1]>quantile(table_coord[,1],0.99)|table_coord[,1]<quantile(table_coord[,1],0.01)|table_coord[,2]>quantile(table_coord[,2],0.99)|table_coord[,2]<quantile(table_coord[,2],0.01),]

for (i in 1:length(pca_data[,1])){
  if (!pca_data[i,6]%in%rownames(particuliers)){
    pca_data[i,6]=""
    }
}
p <- ggplot(pca_data, aes(x=Dim.1, y=Dim.2, colour = as.factor(datas[,6]),label=ID)) +
 # viridis::scale_color_viridis("Cold_diffuse", discrete = TRUE,end = 0.8)
  #  scale_color_manual(values= c("#E7B800", "#2E9FDF"))+
  geom_text_repel() +
  theme_bw() +
  geom_point(size=2, alpha=1) +
  guides(color=guide_legend(override.aes = list(alpha=1),title="Type")) +
  xlab(label=paste0("PC1 (", round(res.pca$eig[1,2], 2),"% of total variance explained)")) +
  ylab(label=paste0("PC2 (", round(res.pca$eig[2,2], 2), "% of total variance explained)")) +
  ggtitle("", subtitle = "First factorial plan")

pca_data <- cbind.data.frame(res.pca$ind$coord[, 1:4],datas[,4],datas[,5],rownames(res.pca$ind$coord))
colnames(pca_data)=c("Dim.1","Dim.2","Dim.3","Dim.4","Event","Time","ID")

p2 <- ggplot(pca_data, aes(x=Time, y=Dim.1, colour = as.factor(datas[,6]),shape=as.factor(Event))) +
 # viridis::scale_color_viridis("Cold_diffuse", discrete = TRUE,end = 0.8)
  #  scale_color_manual(values= c("#E7B800", "#2E9FDF"))
  theme_bw() +
  geom_point(size=2, alpha=1) +
  guides(color=guide_legend(override.aes = list(alpha=1),title="Type"),shape=guide_legend(title="Event")) +
  xlab(label="Survival time") +
  ylab(label=paste0("PC1 (", round(res.pca$eig[1,2], 2), "% of total variance explained)")) +
  ggtitle("", subtitle = "First factorial plan")

```

```{r,eval=T,echo=F,message=FALSE,warning=FALSE,cache=FALSE, fig.cap="PCA of the RNA data"}
p
```

We can also look at the first coordinate of PCA in function of the time of survival :

```{r,eval=T,echo=F,message=FALSE,warning=FALSE,cache=FALSE, fig.cap="PCA of the RNA data"}
p2
```

There seem to be no particular group of patients, the GBM patients are more separated from astro and oligo patients, who are mixed.

Here are the kaplan meier curves for the three types of disease:

```{r,eval=T,echo=F,message=FALSE,warning=FALSE,cache=FALSE, fig.cap="Kaplan meier curves for Astro, Oligo and GBM"}
fit <- survfit(Surv(time=as.numeric(OVERALL_SURVIVAL_MONTHS), event=as.numeric(EVENT_OS)) ~ DISEASE_TYPE, data = datas)

 p_type=ggsurvplot(fit, data = datas,
            surv.median.line = "hv", # Add medians survival

            # Change legends: title & labels
            legend.title = "Type",
            legend.labs = c("Astro", "GBM","Oligo"),

            conf.int = TRUE,
            # Add risk table
            risk.table = TRUE,
            tables.height = 0.2,
            tables.theme = theme_cleantable(),

            title="Type of disease",
            ggtheme = theme_bw() # Change ggplot2 theme
 )
 p_type

```

The GBM patients are more at risk thank the astro and oligo patients, who have similar survival curves. We can also look at the histogram of the survival times according to the disease type:

```{r,eval=T,echo=F,message=FALSE,warning=FALSE,cache=FALSE, fig.cap="survival time for each type"}
ggplot(datas,aes(x=OVERALL_SURVIVAL_MONTHS,fill=as.factor(EVENT_OS)))+geom_histogram()+facet_grid(.~DISEASE_TYPE)+ 
  scale_fill_discrete("Event")
```

### 2022-10-11 script/datamanage_rembrandt.R

@@ -1,0 +1,66 @@
+################################################# load packages ####################################
+library(dplyr)
+library(hgu133plus2.db)
+
+################################################# import data ####################################
+
+##### clinical data
+GSE108474_REMBRANDT_Clinical <- read.table(file = "GSE108474_REMBRANDT_clinical.data.txt",
+                                                 sep = "\t",
+                                                 header = T)
+##### gene-clinical mapping
+GSE108474_REMBRANDT_Mapping <- read.table(file = "GSE108474_REMBRANDT_biospecimen_mapping_GEO.txt",
+                                          sep = "\t",
+                                          header = T) %>%
+  mutate_all(.funs = function(x) gsub(pattern = "-", replacement = "_", x = x))
+
+##### gene expression data
+### import data
+GSE108474_REMBRANDT_GeneExpression <- read.table(file = "GSE108474_REMBRANDT_GeneExpression.txt",
+                                                 sep = "\t",
+                                                 header = T,
+                                                 row.names = 1) %>%
+  tibble::rownames_to_column(var = "PROBEID")
+
+### probe annotation
+dfProbeGene <- AnnotationDbi::select(x = hgu133plus2.db,
+                                     keys = GSE108474_REMBRANDT_GeneExpression$PROBEID,
+                                     columns = c("PROBEID", "SYMBOL"),
+                                     keytype = "PROBEID") %>%
+  dplyr::distinct() %>%
+  dplyr::rename("GENE" = "SYMBOL")
+
+dfGenes <- dfProbeGene %>%
+  dplyr::left_join(GSE108474_REMBRANDT_GeneExpression, by = "PROBEID") %>%
+  ## fix patients id to map to clinical data
+  dplyr::rename_with(.fn = function(x) ifelse(grepl(x,
+                                                    pattern = "^X|^PROBEID|^GENE"),
+                                              x,
+                                              paste0("X", x)) %>%
+                       gsub(pattern = "\\.",
+                            replacement = "_",
+                            x = .))
+
+### transpose the dataframe and aggregate with the mean when a gene is identified by several probes
+dfGenesWide <- dfGenes %>%
+  dplyr::select(-PROBEID) %>%
+  tidyr::pivot_longer(cols = X00518392_U133P2:XNormal_7_U133_P2,
[…]

### 2022-10-11 script/generation_figure.R

@@ -1,0 +1,80 @@
+### import data
+results=read.table("results/comparaison_survival_sets_bis_corr_beta6_bis_2000.txt",
+                   col.names = c("ite","beta","corr","cens","prop","size","gt","wt","bst","gbj"))
+combinaisons=read.table("data/simulations_table.txt",header=TRUE)
+
+# Je comprends rien à cette section
+###
+lines=seq(1,length(results[,1]),by=2)
+results=results[lines,]
+results_size=results
+results=results_size[results_size[,6]==50,-6]
+###
+
+### change les cp
+# 0.3 -> 0.4
+# 0.1 -> 0.3
+# 0.5 -> 0.5
+for (i in 1:length(combinaisons[,1])){
+  
+  if (combinaisons[i,1]==0.3){
+    combinaisons[i,1]=0.4
+  }
+  if (combinaisons[i,1]==0.1){
+    combinaisons[i,1]=0.3
+  }
+  if (combinaisons[i,1]==0.5){
+    combinaisons[i,1]=0.5
+  }
+  
+}
+gt=numeric(length(combinaisons[,1]))
+wt=numeric(length(combinaisons[,1]))
+gbt=numeric(length(combinaisons[,1]))
+gbj=numeric(length(combinaisons[,1]))
+for (i in 1:length(combinaisons[,1])){
+  results_combinaison=results[results[,2]==combinaisons[i,4]&results[,3]==combinaisons[i,2]&results[,4]==combinaisons[i,3]&results[,5]==combinaisons[i,1],]
+  #results_combinaison=distinct(results_combinaison,ite, .keep_all= TRUE)
+  gt[i]=length(results_combinaison[results_combinaison[,6]<0.05,1])/length(results_combinaison[,1])
+  wt[i]=length(results_combinaison[results_combinaison[,7]<0.05,1])/length(results_combinaison[,1])
+  gbt[i]=length(results_combinaison[results_combinaison[,8]<0.05,1])/length(results_combinaison[,1])
+  gbj[i]=length(results_combinaison[results_combinaison[,9]<0.05,1])/length(results_combinaison[,1])
+}
+
+resultats=cbind(combinaisons,gbj,gt,wt,gbt)
+
+resultats_total=resultats
+library(reshape2)
+library(ggplot2)
[…]

### 2022-10-11 script/thomas_rembrandt.R

@@ -1,0 +1,24 @@
+pathways=GSA.read.gmt("data/c6.all.v7.1.symbols.gmt")
+liste_pathways=pathways$genesets
+liste_names=pathways$geneset.names
+
+
+datas_tot=readRDS("data/dfGenesAndClinical.rds")
+datas_tot=datas_tot[,-c(1,3,5,6,11:34)]
+datas_tot=datas_tot[,-c(2,3,7)]
+datas_tot=datas_tot[!is.na(datas_tot[,2])&!is.na(datas_tot[,3]),]
+datas_tot=datas_tot[datas_tot[,4]%in%c("ASTROCYTOMA","GBM","OLIGODENDROGLIOMA"),]
+
+
+path=100
+
+counts_pathway=datas_tot[,colnames(datas_tot)%in%liste_pathways[[path]]]
+patients=datas_tot[,c(1,3,2,4)]
+patients=as.data.frame(patients)
+st=as.numeric(patients[,2])
+Recurrence=as.factor(patients[,3])
+
+colnames(patients)=c("ID","st","Recurrence","type")
+datas=cbind(patients,counts_pathway)
+datas=as.data.frame(datas)
+datas[,2]=as.numeric(as.character(datas[,2]))

### 2023-02-15 script/rembrandt/rembrandt_pathways.Rmd
update scri^pt mega bug globalboosttest

---
title: "Rembrandt pathway analysis"
author: "TF"
date: "08/06/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
library(dplyr)
library(ggplot2)
library(survival)
```

```{r data, results='hide'}
dfRemb <- readRDS(file = "../../data/rembrandt/dfGenesAndClinical.rds")
pathways <- GSA::GSA.read.gmt("../../data/rembrandt/c6.all.v7.1.symbols.gmt")
```

```{r}
lapply(pathways$genesets,
       function(vec_genes){
         dfRemb %>% select(any_of(vec_genes)) %>% ncol()
       }) %>%
  unlist() %>%
  hist(main = "Number of genes by pathway found in Rembrandt data")

lapply(pathways$genesets,
       length) %>%
  unlist() %>%
  hist(main = "Number of genes by pathway")

```

### 2023-02-15 functions/fct_generate_data.R

@@ -31,10 +31,10 @@ fct_generate_data <- function(case,
                                 nb_genes = nb_genes)
   
   ### generate genes values
-  x<-rmvnorm(n=nb_observations,
-             mean=rep(0,nb_genes),
-             sigma=mat_var_covar,
-             method="chol")
+  x<-mvtnorm::rmvnorm(n=nb_observations,
+                      mean=rep(0,nb_genes),
+                      sigma=mat_var_covar,
+                      method="chol")
   colnames(x) <- paste0("gene", 1:ncol(x))
   
   ### generate survival time

### 2023-02-15 functions/fct_generate_survival_data.R

@@ -5,6 +5,7 @@
 #' @param censoring censoring proportion
 #' @param vec_beta vector of beta for cox model
 #' @param x the gene set
+#' @param cp Censoring proportion
 #'
 #' @return A dataframe with survival time, event and gene set
 #' @export
@@ -21,13 +22,13 @@ fct_generate_survival_data <- function(censoring,
   time<-exp(-predictor)*(-log(1-randu))/slam
   
   # censoring
-  if (cp==0) {
+  if (censoring==0) {
     df_survival <- data.frame(time = time,
                               event = 1)
   }else{
-    clam<-exp(predictor)*slam*cp/(1-cp)
-    csg<-rexp(ns,rate=clam)
-    dim(csg)<-c(ns,1)
+    clam<-exp(predictor)*slam*censoring/(1-censoring)
+    csg<-rexp(length(vec_beta),rate=clam)
+    dim(csg)<-c(length(vec_beta),1)
     csgind<-(time<=csg)
     otime<-time*csgind+csg*(1-csgind)
     

### 2023-02-15 functions/fct_simulation_paper.R

@@ -1,0 +1,96 @@
+#' Function to retrieve p-values from the different scenarios
+#'
+#' @param nb_observations Number of observations (default = 50)
+#' @param nb_genes Number of genes (default = 50)
+#' @param prop_sig_gene Proportion of significant gene (default = 0.2)
+#' @param variance Variance (default = 2)
+#' @param case Case of simulation
+#' @param type Type of simulation
+#' @param censoring Censoring proportion
+#' @param nb_permutation Number of permutation for other tests
+#'
+#' @return A dataframe with the different p-values
+#' @export
+fct_simulation_paper <- function(nb_observations = 50,
+                                 nb_genes = 50,
+                                 prop_sig_gene = 0.2,
+                                 variance = 0.2,
+                                 case = 2,
+                                 type = "A",
+                                 censoring = 0.3,
+                                 nb_permutation = 1000){
+  
+  
+  ##### generate data #####
+  dfdata <- fct_generate_data(case = case,
+                              variance = variance,
+                              type = type,
+                              prop_sig_gene = prop_sig_gene,
+                              nb_genes = nb_genes,
+                              censoring = censoring,
+                              vec_beta = vec_beta,
+                              x = x)
+  
+  ##### sGBJ #####
+  resSGBJ <- sGBJ::sGBJ(surv = survival::Surv(dfdata$time, dfdata$event),
+                        factor_matrix = dfdata %>% select(-time, -event) %>% as.matrix())
+  
+  ##### Other methods #####
+  ots <- statsTest(vectime = dfdata$time,
+                   vecevent = dfdata$event,
+                   x = dfdata %>% select(-time, -event) %>% as.matrix())
+  rts<-matrix(0,nrow=3,ncol=nb_permutation)
+  
+  for(b in 1:nb_permutation){
+    if(b %% 10 == 0){
+      print(b)
+    }
+    
+    dfdatapermuted <- dfdata %>%
+      dplyr::select(time, event) %>%
+      tibble::as_tibble() %>%
+      dplyr::slice_sample(prop = 1, replace = TRUE)
+    
+    print(dfdatapermuted[1:5,])
+    
+    vectime = dfdatapermuted$time
+    vecevent = dfdatapermuted$event
+    
+    ng <- ncol(x)
+    ucox<-rep(0,ng)
+
+    survObj <- survival::Surv(vectime,vecevent)
+
[…]

### 2023-02-15 functions/statsTest.R

@@ -2,38 +2,41 @@
 #'
 #' @description Function which compute test statistics for Global Boost test, Wald test and Global Test
 #'
-#' @param time the time
-#' @param event the event
+#' @param vectime the time
+#' @param vecevent the event
 #' @param x the genes experession
 #'
 #' @return A vector with the statistical test of global test, adewale test and global boost test
 #' @export
 #'
-statsTest=function(time, event, x){
+statsTest=function(vectime, vecevent, x){
   ng <- ncol(x)
-  wtcx<-matrix(0,nrow=2,ncol=ng)
   ucox<-rep(0,ng)
-  escx0<-0
-  escx1<-0
-  gt<-0
-  aw<-0
-  bst<-0
+  
+  survObj <- survival::Surv(vectime,vecevent)
   
   for(i in 1:ng){
-    ocx<-coxph(Surv(time,event)~x[,i])
+    ocx<-survival::coxph(survObj~x[,i])
     ucox[i]<-ocx$coefficients/sqrt(ocx$var)
   }
   
   #global test
-  ogt<-gt(Surv(time,event),x[,1:gsize],model="cox")
-  gt<-z.score(ogt)
+  ogt<-globaltest::gt(survObj,
+                      x,
+                      model="cox")
+  gt<-globaltest::z.score(ogt)
   
   #Adewale test
-  aw<-sum(ucox[1:gsize]^2)
+  aw<-sum(ucox^2)
   
   #Global boost test
   mstop<-500
-  gbst<-globalboosttest(x[,1:gsize],Surv(time,event),Z=NULL,nperm=1,mstop=mstop,pvalueonly=FALSE)
+  gbst<-globalboosttest::globalboosttest(x,
+                                         survObj,
+                                         Z=NULL,
+                                         nperm=1,
+                                         mstop=mstop,
+                                         pvalueonly=FALSE)
   bst<--gbst$riskreal[500]
   
   ts=c(gt,aw,bst)

### 2023-02-15 script/rembrandt/thomas_rembrandt.R

@@ -1,0 +1,44 @@
+library(dplyr)
+
+##### import pathways
+pathways=GSA::GSA.read.gmt("data/rembrandt/c6.all.v7.1.symbols.gmt")
+liste_pathways=pathways$genesets
+liste_names=pathways$geneset.names
+
+##### import data
+datas_tot=readRDS("data/rembrandt/dfGenesAndClinical.rds") %>%
+  select(-all_of(c("SUBJECT_ID",
+                   "PARENT_ID", "TYPE",
+                   "TIMEPOINT", "AGE_RANGE",
+                   "GENDER",
+                   "WHO_GRADE", "RACE",
+                   "INSTITUTION_NAME", "KARNOFSKY",
+                   "NEURO_EXAM_SCORE", "DISEASE_EVAL_MRI",
+                   "STEROID_DOSE_STATUS", "ANTI_CONVULSANT_STATUS",
+                   "PRIOR_THERAPY_RADIATION_TYPE", "PRIOR_THERAPY_CHEMO_AGENT_NAME",
+                   "PRIOR_THERAPY_SURGERY_TUMOR_HISTOLOGY", "PRIOR_THERAPY_PROCEDURE",
+                   "PRIOR_THERAPY_SURGERY_OUTCOME", "ON_THERAPY_RADIATION_SITE",
+                   "ON_THERAPY_RADIATION_TYPE", "ON_THERAPY_CHEMO_AGENT_NAME",
+                   "ON_THERAPY_SURGERY_HIST_DIAG", "ON_THERAPY_SURGERY_TITLE",
+                   "ON_THERAPY_SURGERY_OUTCOME", "PRIOR_THERAPY_SURGERY_OUTCOME_2",
+                   "PT_NAME2", "PRIOR_THERAPY_SURGERY_OUTCOME_1",
+                   "PT_NAME1", "GENE_EXP",
+                   "COPY_NUMBER"))) %>%
+  filter(DISEASE_TYPE %in% c("ASTROCYTOMA","GBM","OLIGODENDROGLIOMA"),
+         !is.na(EVENT_OS))
+
+lapply(liste_pathways, FUN = function(path){
+  print(path)
+  ncol(datas_tot[,colnames(datas_tot) %in% path])
+})
+
+counts_pathway=datas_tot[,colnames(datas_tot)%in%liste_pathways[[path]]]
+patients=datas_tot[,c(1,3,2,4)]
+patients=as.data.frame(patients)
[…]

### 2023-02-15 script/script_simulation/GBJ_simulations_clean.R

@@ -5,53 +5,30 @@ source(file = "functions/fct_generate_beta.R")
 source(file = "functions/fct_generate_survival_data.R")
 source(file = "functions/fct_generate_data.R")
 source(file = "functions/statsTest.R")
+source(file = "functions/fct_simulation_paper.R")
 
 library(dplyr)
 
 ##### hyperparameters #####
 
-nb_observations = 50
-nb_genes = 50
-prop_sig_gene = 0.2
-variance = 0.2
-case = 2
-type = "A"
-censoring = 0.3
-nb_permutation = 2
+# paper default : 2000 repetition
+nb_observations = 50 # paper default = 50
+nb_genes = 50 # paper default = 50
+prop_sig_gene = 0.2 # pg = c(0.05, 0.1, 0.2)
+variance = 0.2 # paper default = 0.2
+case = 2 # case = c(1, 2, 3)
+type = "A" # type = c("A", "B", "C")
+censoring = 0.3 # censoring = c(0, 0.3)
+nb_permutation = 1000 # paper default = 1000 
 
-##### generate data #####
-
-dfdata <- fct_generate_data(case = case,
-                            variance = variance,
-                            type = type,
-                            prop_sig_gene = prop_sig_gene,
-                            nb_genes = nb_genes,
-                            censoring = censoring,
-                            vec_beta = vec_beta,
-                            x = x)
-
-##### sGBJ #####
-resSGBJ <- sGBJ::sGBJ(surv = Surv(dfdata$time, dfdata$event),
-                      factor_matrix = dfdata %>% select(-time, -event) %>% as.matrix())
-
-##### Other methods #####
-ots <- statsTest(time = dfdata$time, event = dfdata$event, x = dfdata %>% select(-time, -event) %>% as.matrix())
-rts<-matrix(0,nrow=3,ncol=nb_permutation)
-for(b in 1:nb_permutation){
-  print(b)
-  
-  permute<-sample(1:ns)
-  
-  dfdatapermuted <- dfdata %>%
-    select(time, event) %>%
-    sample_frac(size = 1, replace = TRUE) %>%
-    bind_cols(dfdata %>%
-                select(-c(time, event)))
-  
-  rts[,b]<-statsTest(time = dfdatapermuted$time,
-                     event = dfdatapermuted$event,
-                     x = dfdatapermuted %>% select(-time, -event) %>% as.matrix())
-}
-p.pvalue<-apply((rts>ots),1,mean)
+set.seed(2)
+fct_simulation_paper(nb_observations = 50,
[…]

### 2023-02-15 script/report_var_covar.Rmd
update functions ready for big simulations

---
title: "VarCovar Rembrandt"
author: "TF"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(dplyr)
library(ggplot2)
dfdata <- readRDS("data/rembrandt/dfresvarcovar.rds")
```

Each dot in the following figure corresponds to one pathway.

# Variance

Each dot in the figure corresponds to the mean and the variance of a pathway. In the draft, variance of the gene expression was set to 0.2.

```{r}
dfdata %>%
  filter(param == "var") %>%
  ggplot(mapping = aes(x = mean,
                       y = var,
                       color = signif)) +
  geom_point() +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "Mean of variance", y = "Variance of variance",
       title = "Rembrandt variance distribution of genes of a given pathway",
       color = "Significant",
       subtitle = "Paper : variance = 0.2")
```

# Covariance

Each dot in the figure corresponds to the mean and the variance of the covariance of a pathway. In the draft, we used three scenario represented above.

```{r}
data_paper <- data.frame(mean = c(0, 0.4*0),
                         var = c(0, 0.4^2*0.01^2))

dfdata %>%
  filter(param == "covar") %>%
  ggplot(mapping = aes(x = mean,
                       y = var,
                       color = signif)) +
  geom_point() +
  geom_point(data_paper, mapping = aes(color = "Draft"), size = 5) +
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_y_log10() +
  labs(x = "Mean", y = "Variance", color = "",
       title = "Rembrandt covariance distribution",
       subtitle = "Draft scenario II is not shown (mean = 0.4*0.4, var = 0.1^2*0.4^2)")
```

### 2023-02-15 functions/fct_generate_survival_data.R

@@ -36,7 +36,7 @@ fct_generate_survival_data <- function(censoring,
                               event = as.numeric(csgind))
   }
   
-  res <- dplyr::bind_cols(df_survival, x)
+  res <- cbind(df_survival, x)
   
   return(res)
 }

### 2023-02-15 functions/fct_simulation_paper.R

@@ -46,48 +46,19 @@ fct_simulation_paper <- function(nb_observations = 50,
       print(b)
     }
     
+    seed <- as.numeric(Sys.time())+b
+    
+    set.seed(seed)
+    
     dfdatapermuted <- dfdata %>%
       dplyr::select(time, event) %>%
-      tibble::as_tibble() %>%
       dplyr::slice_sample(prop = 1, replace = TRUE)
     
-    print(dfdatapermuted[1:5,])
-    
-    vectime = dfdatapermuted$time
-    vecevent = dfdatapermuted$event
-    
-    ng <- ncol(x)
-    ucox<-rep(0,ng)
-
-    survObj <- survival::Surv(vectime,vecevent)
-
-    for(i in 1:ng){
-      ocx<-survival::coxph(survObj~x[,i])
-      ucox[i]<-ocx$coefficients/sqrt(ocx$var)
-    }
-
-    #global test
-    ogt<-globaltest::gt(survObj,
-                        x,
-                        model="cox")
-    gt<-globaltest::z.score(ogt)
-
-    #Adewale test
-    aw<-sum(ucox^2)
-
-    #Global boost test
-    mstop<-500
-    gbst<-globalboosttest::globalboosttest(x,
-                                           survObj,
-                                           Z=NULL,
-                                           nperm=1,
-                                           mstop=mstop,
-                                           pvalueonly=FALSE)
-    bst<--gbst$riskreal[500]
-    # 
-    # ts=c(gt,aw,bst)
+    rts[,b] <- statsTest(vectime = dfdatapermuted$time,
+                     vecevent = dfdatapermuted$event,
+                     x = dfdata %>% select(-time, -event) %>% as.matrix())
   }
-  p.pvalue<-c(resSGBJ$sGBJ_pvalue, apply((rts>ots),1,mean))
+  p.pvalue<-c(resSGBJ$GBJ_pvalue, apply((rts>ots),1,mean))
   
   res <- data.frame(p_value = p.pvalue,
                     method = c("sGBJ", "Global Test", "Wald Test","Global Boost Test"))

### 2023-02-15 script/script_simulation/GBJ_simulations_clean.R

@@ -19,7 +19,7 @@ variance = 0.2 # paper default = 0.2
 case = 2 # case = c(1, 2, 3)
 type = "A" # type = c("A", "B", "C")
 censoring = 0.3 # censoring = c(0, 0.3)
-nb_permutation = 1000 # paper default = 1000 
+nb_permutation = 100 # paper default = 1000 
 
 set.seed(2)
 fct_simulation_paper(nb_observations = 50,

### 2023-02-15 script/thomas_rembrandt.R

@@ -1,24 +1,64 @@
-pathways=GSA.read.gmt("data/c6.all.v7.1.symbols.gmt")
+library(dplyr)
+library(survival)
+
+##### load data
+pathways = GSA::GSA.read.gmt("data/rembrandt/c6.all.v7.1.symbols.gmt")
+datas_tot = readRDS("data/rembrandt/dfGenesAndClinical.rds")
+
+##### data cleaning
 liste_pathways=pathways$genesets
 liste_names=pathways$geneset.names
+names(liste_pathways) <- liste_names
 
+datas_clean <- datas_tot %>%
+  select(-c("BIOSPECIMEN_ID")) %>%
+  filter(DISEASE_TYPE %in% c("ASTROCYTOMA","GBM","OLIGODENDROGLIOMA"))
 
-datas_tot=readRDS("data/dfGenesAndClinical.rds")
-datas_tot=datas_tot[,-c(1,3,5,6,11:34)]
-datas_tot=datas_tot[,-c(2,3,7)]
-datas_tot=datas_tot[!is.na(datas_tot[,2])&!is.na(datas_tot[,3]),]
-datas_tot=datas_tot[datas_tot[,4]%in%c("ASTROCYTOMA","GBM","OLIGODENDROGLIOMA"),]
+##### test 1 pathway
 
+extract_varcovar <- function(pathway_i){
+  cox_model <- datas_clean %>%
+    select(EVENT_OS, OVERALL_SURVIVAL_MONTHS, any_of(pathway_i)) %>%
+    coxph(Surv(OVERALL_SURVIVAL_MONTHS, EVENT_OS) ~ .,
+          data = .)
+  
+  vcov_path <- cox_model %>%
+    vcov()
+  
+  pval_path <- summary(cox_model)$coefficients[,"Pr(>|z|)"]
+  
+  sign_genes <- names(pval_path)[pval_path<0.05]
+  nosign_genes <- names(pval_path)[!pval_path<0.05]
+  all_genes <- names(pval_path)
+  
+  
+  ## get variance and covariance
+  dfres <- lapply(list(sign = sign_genes,
+                       nosign = nosign_genes,
+                       all = all_genes),
+                  FUN = function(x){
+                    vcov_path_select <- vcov_path[rownames(vcov_path) %in% x, colnames(vcov_path) %in% x]
+                    vec_variance <- diag(vcov_path_select)
+                    diag(vcov_path_select) <- 0
[…]

### 2023-02-16 functions/fct_generate_data.R
update two prop

@@ -10,6 +10,9 @@
 #' @param censoring The proportion of censoring
 #' @param vec_beta The vector of beta
 #' @param x The gene set
+#' @param nb_observations The number of observations
+#' @param prop_two_periods Should the beta be divided by 2 after a given period (not respect proportionality assumption)
+#' @param slam A constant used in the exponential simulation (default = 0.005)
 #'
 #' @return A dataframe with time, event and genes
 #' @export
@@ -18,8 +21,11 @@ fct_generate_data <- function(case,
                               type,
                               prop_sig_gene,
                               nb_genes,
+                              nb_observations,
                               censoring,
                               vec_beta,
+                              prop_two_periods = FALSE,
+                              slam = 0.005,
                               x){
   mat_var_covar <- fct_generate_varcovar(case = case,
                                          prop_sig_gene = prop_sig_gene,
@@ -40,6 +46,8 @@ fct_generate_data <- function(case,
   ### generate survival time
   dfsurvival <- fct_generate_survival_data(censoring = censoring,
                                            vec_beta = vec_beta,
+                                           prop_two_periods = prop_two_periods,
+                                           slam = slam,
                                            x = x)
   
   return(dfsurvival)

### 2023-02-16 functions/fct_generate_survival_data.R

@@ -4,22 +4,26 @@
 #'
 #' @param censoring censoring proportion
 #' @param vec_beta vector of beta for cox model
+#' @param prop_two_periods Should the beta be divided by 2 after a given period (not respect proportionality assumption)
+#' @param slam A constant used in the exponential simulation (default = 0.005)
 #' @param x the gene set
-#' @param cp Censoring proportion
 #'
 #' @return A dataframe with survival time, event and gene set
 #' @export
 #' 
 fct_generate_survival_data <- function(censoring,
                                        vec_beta,
+                                       prop_two_periods = FALSE,
+                                       slam = 0.005,
                                        x){
   
   # survival time
   predictor<-x%*%vec_beta
-  randu<-runif(nb_observations,min=0,max=1)
-  dim(randu)<-c(nb_observations,1)
-  slam=0.005
-  time<-exp(-predictor)*(-log(1-randu))/slam
+  time <- simu_simple_beta(x = x, predictor = predictor, slam=slam)
+  if(prop_two_periods){
+    time_shift = quantile(time, 0.25)
+    time <- simu_twoperiod_beta(x = x, predictor = predictor, slam=slam, time_shift = time_shift)
+  }
   
   # censoring
   if (censoring==0) {

### 2023-02-16 functions/simulation_time_varying_beta.R

@@ -1,0 +1,51 @@
+#' simu_simple_beta
+#'
+#' @description A function to simulate simple exponential survival function depending on beta
+#'
+#' @param slam A constant used in the exponential simulation (default = 0.005)
+#' @param x Matrix of genes
+#' @param predictor predictor
+#'
+#' @return A vector of failure time
+simu_simple_beta <- function(predictor,
+                             slam = 0.005,
+                             x){
+  nb_observations <- nrow(x)
+  randu<-runif(nb_observations,min=0,max=1)
+  dim(randu)<-c(nb_observations,1)
+  time<-exp(-predictor)*(-log(1-randu))/slam
+  return(time)
+}
+
+#################################################
+##### simu two times effect #####################
+#################################################
+
+
+#' simu_twoperiod_beta
+#'
+#' @description A function to simulate simple exponential survival function depending on beta. After time_shift period, beta is divided by two.
+#'
+#' @param slam A constant used in the exponential simulation
+#' @param x Matrix of genes
+#' @param time_shift Time point after which beta is divided by 2
+#' @param predictor predictor
+#'
+#' @return A vector of failure time
+simu_twoperiod_beta <- function(predictor,
+                                slam = 0.005,
+                                x,
+                                time_shift){
+  time1 <- simu_simple_beta(predictor = predictor,
+                            slam = slam,
+                            x = x)
+  
+  time2 <- simu_simple_beta(predictor = predictor,
+                            slam = slam,
+                            x = x)
+  
+  time <- ifelse(time1 > time_shift,
+                 yes = time_shift + time2,
+                 no = time1)
+  return(time)
+}

### 2023-02-16 functions/statsTest.R

@@ -6,7 +6,7 @@
 #' @param vecevent the event
 #' @param x the genes experession
 #'
-#' @return A vector with the statistical test of global test, adewale test and global boost test
+#' @return A vector with the statistical test of global test, adewale test
 #' @export
 #'
 statsTest=function(vectime, vecevent, x){
@@ -29,16 +29,6 @@ statsTest=function(vectime, vecevent, x){
   #Adewale test
   aw<-sum(ucox^2)
   
-  #Global boost test
-  mstop<-500
-  gbst<-globalboosttest::globalboosttest(x,
-                                         survObj,
-                                         Z=NULL,
-                                         nperm=1,
-                                         mstop=mstop,
-                                         pvalueonly=FALSE)
-  bst<--gbst$riskreal[500]
-  
-  ts=c(gt,aw,bst)
+  ts=c(gt,aw)
   return(ts)
 }

### 2023-02-16 script/script_simulation/GBJ_simulations_clean.R

@@ -1,34 +1,56 @@
 ##### load functions and package #####
-
-source(file = "functions/fct_generate_varcovar.R")
-source(file = "functions/fct_generate_beta.R")
-source(file = "functions/fct_generate_survival_data.R")
-source(file = "functions/fct_generate_data.R")
-source(file = "functions/statsTest.R")
-source(file = "functions/fct_simulation_paper.R")
-
 library(dplyr)
+library(doParallel)
+library(parallel)
+
+lapply(list.files(path = "functions/", full.names = TRUE),
+       source)
 
 ##### hyperparameters #####
 
 # paper default : 2000 repetition
-nb_observations = 50 # paper default = 50
-nb_genes = 50 # paper default = 50
-prop_sig_gene = 0.2 # pg = c(0.05, 0.1, 0.2)
-variance = 0.2 # paper default = 0.2
-case = 2 # case = c(1, 2, 3)
-type = "A" # type = c("A", "B", "C")
-censoring = 0.3 # censoring = c(0, 0.3)
-nb_permutation = 100 # paper default = 1000 
+# nb_observations = 50 # paper default = 50
+# nb_genes = 50 # paper default = 50
+# prop_sig_gene = 0.2 # pg = c(0.05, 0.1, 0.2)
+# variance = 0.2 # paper default = 0.2
+# case = 2 # case = c(1, 2, 3)
+# type = "A" # type = c("A", "B", "C")
+# censoring = 0.3 # censoring = c(0, 0.3)
+# nb_permutation = 100 # paper default = 1000
+# prop_two_periods = FALSE
+# slam = 0.005
 
 set.seed(2)
-fct_simulation_paper(nb_observations = 50,
-                     nb_genes = 50,
-                     prop_sig_gene = 0.2,
-                     variance = 0.2,
-                     case = 1,
-                     type = "A",
-                     censoring = 0.3,
-                     nb_permutation = 10)
+
+myfunction <- function(x){
+  library(dplyr)
+  fct_simulation_paper(nb_observations = 50,
+                       nb_genes = 50,
+                       prop_sig_gene = 0.2,
+                       variance = 0.2,
+                       case = 3,
+                       type = "C",
+                       censoring = 0.3,
+                       nb_permutation = 3,
+                       prop_two_periods = FALSE)
+}
+
+
+no_cores <- detectCores(logical = TRUE)
+cl <- makeCluster(no_cores-1)  
[…]

### 2023-03-03 functions/simulation_time_varying_beta.R
update functions correct bug prop hazard

@@ -40,7 +40,7 @@ simu_twoperiod_beta <- function(predictor,
                             slam = slam,
                             x = x)
   
-  time2 <- simu_simple_beta(predictor = predictor,
+  time2 <- simu_simple_beta(predictor = predictor/2,
                             slam = slam,
                             x = x)
   

### 2023-03-03 script/breast_cancer/01_datamanagement.R

@@ -1,0 +1,32 @@
+##### load packages #####
+library(dplyr)
+
+# add method to grid.draw
+grid.draw.ggsurvplot <- function(x){
+  survminer:::print.ggsurvplot(x, newpage = FALSE)
+}
+
+##### load data #####
+load("data/brest_cancer/BreastCancer-Vijver-Clinical.RData")
+load("data/brest_cancer/BreastCancer-Vijver.RData")
+load("data/brest_cancer/c2.cp.v2.5.GS.RData")
+
+##### data management #####
+
+### clinical data
+df_clinical <- clin.data %>%
+  mutate(event = as.numeric(evdeath == 1 | evmeta == 1),
+         time = if_else(tsurv > tmeta, tmeta, tsurv)) %>%
+  select(ID, grade, event, time, age)
+
+### add counts
+counts=t(llid_norm.mat)
+
+df_all <- df_clinical %>%
+  bind_cols(counts) %>%
+  as_tibble()
+
+### pathways
+saveRDS(object = list(data = df_all,
+                      pathways = GS.llids),
+        file = "data/brest_cancer/datamanaged.rds")

### 2023-03-03 script/breast_cancer/02_survival_analysis.R

@@ -1,0 +1,40 @@
+##### load packages #####
+library(dplyr)
+library(ggplot2)
+library(survival)
+library(survminer)
+
+##### survival analysis #####
+lsdata <- readRDS(file = "data/brest_cancer/datamanaged.rds")
+
+### marginal
+surv_marginal <- survfit(Surv(time = time, event = event) ~ 1,
+                         data = lsdata$data)
+
+plot_marginal_km <- ggsurvplot(surv_marginal,
+                               risk.table = TRUE,
+                               conf.int = TRUE,
+                               ggtheme = theme_bw(),
+                               palette = "black")
+
+ggsave(filename = "images/plot_marginal_km.pdf",
+       plot = plot_marginal_km,
+       device = "pdf",
+       height = 6,
+       width = 6)
+
+### stratified 
+surv_grade <- survfit(Surv(time = time, event = event) ~ grade,
+                      data = lsdata$data)
+
+plot_grade_km <- ggsurvplot(surv_grade,
+                            risk.table = TRUE,
+                            conf.int = TRUE,
+                            palette = c("#FFCD4D", "#F48C06", "#9D0208"),
+                            ggtheme = theme_bw())
+
+ggsave(filename = "images/plot_grade_km.pdf",
+       plot = plot_grade_km,
+       device = "pdf",
+       height = 6,
+       width = 6)

### 2023-03-03 script/breast_cancer/03_pathway_analysis.R

@@ -1,0 +1,79 @@
+##### load packages #####
+library(dplyr)
+library(ggplot2)
+library(survival)
+library(survminer)
+library(parallel)
+
+##### survival analysis #####
+lsdata <- readRDS(file = "data/brest_cancer/datamanaged.rds")
+source("script/breast_cancer/liste70pathways.R")
+
+##### marginal #####
+df_res_marginal <- mclapply(X = liste_70,
+                            mc.cores = 3,
+                            FUN = function(pathway_i){
+                              vec_pathways <- lsdata$pathways[[pathway_i]]
+                              
+                              surv_obj <- survival::Surv(time = lsdata$data$time,
+                                                         event = lsdata$data$event)
+                              
+                              factor_matrix <- lsdata$data %>%
+                                select(any_of(vec_pathways))
+                              
+                              covariates <- lsdata$data %>% select(age)
+                              
+                              test <- sGBJ::sGBJ(surv = surv_obj,
+                                                 factor_matrix = factor_matrix,
+                                                 covariates = covariates)
+                              
+                              res <- data.frame(pathway = pathway_i,
+                                                nb_genes = ncol(factor_matrix),
+                                                p_value = test$sGBJ_pvalue)
+                              
+                              return(res)
+                            }) %>%
+  bind_rows() %>%
+  mutate(p_value_adjusted = p.adjust(p = p_value, method = "hochberg"))
+
+saveRDS(df_res_marginal, file = "data/brest_cancer/df_res_marginal.rds")
+
+##### stratified #####
+df_res_stratified <- mclapply(X = liste_70,
+                              mc.cores = 3,
+                              FUN = function(pathway_i){
+                                vec_pathways <- lsdata$pathways[[pathway_i]]
+                                vec_grade <- unique(lsdata$data$grade)
+                                
+                                res <- lapply(X = vec_grade,
+                                              FUN = function(grade_i){
+                                                dfdata <- lsdata$data %>% filter(grade == grade_i)
+                                                
+                                                surv_obj <- survival::Surv(time = dfdata$time,
+                                                                           event = dfdata$event)
+                                                
+                                                factor_matrix <- dfdata %>%
+                                                  select(any_of(vec_pathways))
+                                                
+                                                covariates <- dfdata %>% select(age)
+                                                
+                                                test <- sGBJ::sGBJ(surv = surv_obj,
+                                                                   factor_matrix = factor_matrix,
+                                                                   covariates = covariates)
+                                                
[…]

### 2023-03-03 script/breast_cancer/update_stratified_analysis.R

@@ -1,3 +1,0 @@
-source("script/rembrandt/liste70pathways.R")
-
-liste_70

### 2023-04-03 script/script_simulation/figures_simu1.R
update results

@@ -47,10 +47,19 @@ fct_generate_fig1 <- function(data){
 dfScenario <- readRDS("data/dfScenario.rds") %>%
   tibble::rowid_to_column(var = "hp_row")
 
-dfres <- list.files("data/result_simu/results_simu1/", recursive = T, full.names = T) %>%
+dfres1 <- list.files("data/result_simu_1/", recursive = T, full.names = T) %>%
   lapply(FUN = readRDS) %>%
   bind_rows() %>%
-  full_join(dfScenario, by = "hp_row")
+  full_join(dfScenario, by = "hp_row") %>%
+  filter(!prop_two_periods)
+
+dfres2 <- list.files("data/result_simu_2/", recursive = T, full.names = T) %>%
+  lapply(FUN = readRDS) %>%
+  bind_rows() %>%
+  full_join(dfScenario, by = "hp_row") %>%
+  filter(prop_two_periods)
+
+
 
 ##### plot #####
 plot_prop_risk <- dfres %>%

### 2023-04-03 script/script_simulation/generation_figure.R

@@ -1,0 +1,74 @@
+
+
+results=read.table("comparaison_survival_sets_bis_corr_beta6_bis_2000_thomas.txt")
+colnames(results)=c("ite","beta","corr","cens","prop","size","gt","wt","bst","gbj")
+
+lines=seq(1,length(results[,1]),by=2)
+results=results[lines,]
+results_size=results
+
+results=results_size[results_size[,6]==50,-6]
+combinaisons=read.table("simulations_table.txt",header=TRUE)
+for (i in 1:length(combinaisons[,1])){
+  
+  if (combinaisons[i,1]==0.3){
+    combinaisons[i,1]=0.4
+  }
+  if (combinaisons[i,1]==0.1){
+    combinaisons[i,1]=0.3
+  }
+  if (combinaisons[i,1]==0.5){
+    combinaisons[i,1]=0.5
+  }
+  
+}
+gt=numeric(length(combinaisons[,1]))
+wt=numeric(length(combinaisons[,1]))
+gbt=numeric(length(combinaisons[,1]))
+gbj=numeric(length(combinaisons[,1]))
+for (i in 1:length(combinaisons[,1])){
+  results_combinaison=results[results[,2]==combinaisons[i,4]&results[,3]==combinaisons[i,2]&results[,4]==combinaisons[i,3]&results[,5]==combinaisons[i,1],]
+  #results_combinaison=distinct(results_combinaison,ite, .keep_all= TRUE)
+  gt[i]=length(results_combinaison[results_combinaison[,6]<0.05,1])/length(results_combinaison[,1])
+  wt[i]=length(results_combinaison[results_combinaison[,7]<0.05,1])/length(results_combinaison[,1])
+  gbt[i]=length(results_combinaison[results_combinaison[,8]<0.05,1])/length(results_combinaison[,1])
+  gbj[i]=length(results_combinaison[results_combinaison[,9]<0.05,1])/length(results_combinaison[,1])
+}
+
+resultats=cbind(combinaisons,gbj,gt,wt,gbt)
+
+resultats_total=resultats
+library(reshape2)
+library(ggplot2)
+datas=melt(resultats,measure.vars=c("gbj","gt","wt","gbt"))
+colnames(datas)=c("cp","Corr","Censoring","beta","Method","Power")
+for (i in 1:length(datas[,1])){
[…]

### 2023-05-10 script/breast_cancer/01_datamanagement.R
add modifications for review

@@ -7,9 +7,9 @@ grid.draw.ggsurvplot <- function(x){
 }
 
 ##### load data #####
-load("data/brest_cancer/BreastCancer-Vijver-Clinical.RData")
-load("data/brest_cancer/BreastCancer-Vijver.RData")
-load("data/brest_cancer/c2.cp.v2.5.GS.RData")
+load("data/breast_cancer/BreastCancer-Vijver-Clinical.RData")
+load("data/breast_cancer/BreastCancer-Vijver.RData")
+load("data/breast_cancer/c2.cp.v2.5.GS.RData")
 
 ##### data management #####
 
@@ -29,4 +29,4 @@ df_all <- df_clinical %>%
 ### pathways
 saveRDS(object = list(data = df_all,
                       pathways = GS.llids),
-        file = "data/brest_cancer/datamanaged.rds")
+        file = "data/breast_cancer/datamanaged.rds")

### 2023-05-10 script/breast_cancer/02_survival_analysis.R

@@ -5,7 +5,7 @@ library(survival)
 library(survminer)
 
 ##### survival analysis #####
-lsdata <- readRDS(file = "data/brest_cancer/datamanaged.rds")
+lsdata <- readRDS(file = "data/breast_cancer/datamanaged.rds")
 
 ### marginal
 surv_marginal <- survfit(Surv(time = time, event = event) ~ 1,

### 2023-05-10 script/breast_cancer/03_pathway_analysis.R

@@ -6,12 +6,12 @@ library(survminer)
 library(parallel)
 
 ##### survival analysis #####
-lsdata <- readRDS(file = "data/brest_cancer/datamanaged.rds")
+lsdata <- readRDS(file = "data/breast_cancer/datamanaged.rds")
 source("script/breast_cancer/liste70pathways.R")
 
 ##### marginal #####
 df_res_marginal <- mclapply(X = liste_70,
-                            mc.cores = 3,
+                            mc.cores = 19,
                             FUN = function(pathway_i){
                               vec_pathways <- lsdata$pathways[[pathway_i]]
                               
@@ -36,11 +36,11 @@ df_res_marginal <- mclapply(X = liste_70,
   bind_rows() %>%
   mutate(p_value_adjusted = p.adjust(p = p_value, method = "hochberg"))
 
-saveRDS(df_res_marginal, file = "data/brest_cancer/df_res_marginal.rds")
+saveRDS(df_res_marginal, file = "data/breast_cancer/df_res_marginal.rds")
 
 ##### stratified #####
 df_res_stratified <- mclapply(X = liste_70,
-                              mc.cores = 3,
+                              mc.cores = 19,
                               FUN = function(pathway_i){
                                 vec_pathways <- lsdata$pathways[[pathway_i]]
                                 vec_grade <- unique(lsdata$data$grade)
@@ -76,4 +76,4 @@ df_res_stratified <- mclapply(X = liste_70,
   group_by(grade) %>%
   mutate(p_value_adjusted = p.adjust(p = p_value, method = "hochberg"))
 
-saveRDS(df_res_stratified, file = "data/brest_cancer/df_res_stratified.rds")
+saveRDS(df_res_stratified, file = "data/breast_cancer/df_res_stratified.rds")

### 2023-05-10 script/script_simulation/figures_simu1.R

@@ -1,9 +1,9 @@
 ##### load package #####
 library(dplyr)
 library(ggplot2)
+library(stringr)
 
 ##### functions #####
-
 fct_generate_fig1 <- function(data){
   vec_labels_type <- paste0("Type ", c("A", "B", "C"))
   names(vec_labels_type) <- c("A", "B", "C")
@@ -36,7 +36,49 @@ fct_generate_fig1 <- function(data){
     labs(x = "Proportion of significant genes",
          y = "Statistical Power",
          color = "Method",
-         lty = "Censoring proportion") +
+         lty = "Censoring proportion",
+         subtitle = "(A) Non proportional hazard") +
+    lims(y = c(0, 1))
+  
+  return(res)
+}
+
+fct_generate_fig2 <- function(data){
+  vec_labels_type <- paste0("Type ", c("A", "B", "C"))
+  names(vec_labels_type) <- c("A", "B", "C")
+  vec_labels_case <- paste0("Case (", c("I", "II", "III"), ")")
+  names(vec_labels_case) <- c(1:3)
+  
+  res <- data %>%
+    mutate(p_val_sig = p_value < 0.05) %>%
+    group_by(hp_row, method, prop_sig_gene, case, type, censoring) %>%
+    summarise(power = mean(p_val_sig)) %>%
+    mutate(censoring = factor(censoring,
+                              levels = c(0, 0.3),
+                              labels = c("0%", "30%")),
+           method = factor(method,
+                           levels = c("Global Boost Test",
+                                      "Wald Test",
+                                      "Global Test",
+                                      "sGBJ")),
+           group_line = paste0(method, censoring)) %>%
+    ungroup() %>%
+    ggplot(mapping = aes(x = type,
+                         y = power,
+                         color = method,
+                         lty = censoring,
+                         group = group_line)) +
+    geom_point() +
+    geom_line(size = 1) +
+    scale_color_viridis_d() +
+    facet_grid(case ~ .,
+               labeller = labeller(case = vec_labels_case,
+                                   type = vec_labels_type)) +
+    labs(x = "Type",
+         y = "Statistical Power",
+         color = "Method",
+         lty = "Censoring proportion",
+         subtitle = "(B) Proportion of significant genes = 50%") +
     lims(y = c(0, 1))
   
   return(res)
[…]

### 2023-05-10 script/script_simulation/generation_figure.R

@@ -1,6 +1,4 @@
-
-
-results=read.table("comparaison_survival_sets_bis_corr_beta6_bis_2000_thomas.txt")
+results=read.table("data/result_simu_original/comparaison_survival_sets_bis_corr_controle2.txt")
 colnames(results)=c("ite","beta","corr","cens","prop","size","gt","wt","bst","gbj")
 
 lines=seq(1,length(results[,1]),by=2)
@@ -8,17 +6,17 @@ results=results[lines,]
 results_size=results
 
 results=results_size[results_size[,6]==50,-6]
-combinaisons=read.table("simulations_table.txt",header=TRUE)
+combinaisons=read.table("data/simulations_table.txt",header=TRUE)
 for (i in 1:length(combinaisons[,1])){
   
   if (combinaisons[i,1]==0.3){
-    combinaisons[i,1]=0.4
+    combinaisons[i,1]=0.05
   }
   if (combinaisons[i,1]==0.1){
-    combinaisons[i,1]=0.3
+    combinaisons[i,1]=0.1
   }
   if (combinaisons[i,1]==0.5){
-    combinaisons[i,1]=0.5
+    combinaisons[i,1]=0.2
   }
   
 }

### 2023-06-07 reporting/discordance LV TF.Rmd
add html for methods comparison

---
title: "Discordance LV - TF"
author: "TF"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

```{r eval=TRUE, fig.cap="Original plot", fig.height=20, include = TRUE, out.height=700, out.width=700}
knitr::include_graphics("../images/comparison_methods6-1.pdf")
```

```{r eval=TRUE, fig.cap="New plot (fails to replicate)", fig.height=20, include = TRUE, out.height=700, out.width=700}
knitr::include_graphics("../images/plot_original.pdf")
```

# Data generation

## Variance covariance

### TF

```{r}
if(case == 1){

  mat_var_covar <- matrix(data = 0, nrow = nb_genes, ncol = nb_genes)

} else if(case == 2){
  # varcovar for non significant genes
  mat_var_covar <- matrix(data = 0, nrow = nb_genes, ncol = nb_genes)

  # covar for significant genes
  nb_sig_gene = prop_sig_gene * nb_genes

  covar <- 0.4*rnorm(n = nb_sig_gene^2/2-nb_sig_gene/2,
                     mean = 0.4,
                     sd = 0.1)

  # fill upper triangle
  mat_var_covar[1:nb_sig_gene,1:nb_sig_gene][upper.tri(mat_var_covar[1:nb_sig_gene,1:nb_sig_gene])] <- covar
  # duplicate to lower triangle
  mat_var_covar[lower.tri(mat_var_covar)] <- t(mat_var_covar)[lower.tri(mat_var_covar)]

} else if(case == 3){
  # varcovar for significant genes
  covar <- 0.4*rnorm(n = nb_genes^2,
                     mean = 0,
                     sd = 0.01)
  mat_var_covar <- matrix(data = covar,
                          nrow = nb_genes,
                          ncol = nb_genes)
  # duplicate upper triangle to lower triangle
  mat_var_covar[lower.tri(mat_var_covar)] <- t(mat_var_covar)[lower.tri(mat_var_covar)]

} else {
  stop("case must be 1, 2 or 3")
}

# set the variance
diag(mat_var_covar) <- variance

return(mat_var_covar)
```

### LV

```{r}
xsigma<-diag(rep(0.2,ng)) # The variance corresponding to Cjj ?

if (corr==2) {
  xgs<-diag(rep(0.2,ngs))        
  xgs[lower.tri(xgs==TRUE)]<-0.4*(rnorm(ncov,0.4,0.1))
  tmp<-xgs+t(xgs)-diag(diag(xgs))
  xsigma[1:ngs,1:ngs]<-tmp
}
if (corr==3) {
  xgs<-diag(rep(0.2,ngs))        
  xgs[lower.tri(xgs==TRUE)]<-0.4*sample(c(-1,1), size = 1, prob = c(1/2,  1/2),replace = TRUE)*(rnorm(ncov,0.4,0.1))
  tmp<-xgs+t(xgs)-diag(diag(xgs))
  xsigma[1:ngs,1:ngs]<-tmp
}
if (corr==4) {
  xgs<-diag(rep(0.2,ng))        
  xgs[lower.tri(xgs==TRUE)]<-0.4*(rnorm(ncov2,0,0.01))
  tmp<-xgs+t(xgs)-diag(diag(xgs))
  xsigma[1:ng,1:ng]<-tmp
}       
x<-rmvnorm(n=ns,mean=xmean,sigma=xsigma,method="chol")
```

### Comparison

-   Diagonal variance is 0.2 in both
-   Covariance corr is not 2,3,4 in LV is similar to case == 1 in TF
-   corr == 2 in LV is similar to case == 2 in TF
-   corr == 4 in LV is similar to case == 3 in TF

## Beta

### TF

```{r}
nb_sig_gene = round(prop_sig_gene*nb_genes)

if(type == "A"){

  res <- c(rnorm(n = nb_sig_gene, mean = 0, sd = 0.5),
           rep(0, nb_genes-nb_sig_gene))

} else if(type == "B"){

  res <- c(rnorm(n = ceiling(nb_sig_gene/2), mean = 0.1, sd = 0.5),
           rnorm(n = floor(nb_sig_gene/2), mean = -0.1, sd = 0.5),
           rep(0, nb_genes-nb_sig_gene))

} else if(type == "C"){

  res <- c(rnorm(n = ceiling(nb_sig_gene/2), mean = 0.1, sd = 0.25),
           rnorm(n = floor(nb_sig_gene/2), mean = -0.1, sd = 1),
           rep(0, nb_genes-nb_sig_gene))

} else {
  stop("case must be A, B or C")
}

return(res)
```

### LV

```{r}
tau <- 0.5

if (beta==1) {
  bmean<-c(rep(0.1,floor(ngs/2)),
           rep(-0.5,(ngs-floor(ngs/2))),
           rep(0,ng-ngs)) #for power
  bsd<-c(rep(tau,floor(ngs/2)),
         rep(tau/2,(ngs-floor(ngs/2))),
         rep(0,ng-ngs)) #for power
}else{
  bmean<-c(rep(0.1,floor(ngs/2)),
           rep(-0.1,(ngs-floor(ngs/2))),
           rep(0,ng-ngs)) #for power
  bsd<-c(rep(tau/2,floor(ngs/2)),
         rep(tau*2,(ngs-floor(ngs/2))),
         rep(0,ng-ngs)) #for power
}

tbeta<-rnorm(ng,mean=bmean,sd=bsd)
```

### Comparison

-   Nothing corresponds to case == 'A' in LV script
-   There is no scenario in the paper where beta follows gaussian distribution with mean -0.5 (beta == 1 in LV script)
-   The variance defined in LV when beta != 1 corresponds to type == "C" in TF (because tau = 0.5)

It makes sense regarding the graph because results from type == "C" are mostly similar. Problem identified !!

## Survival

### TF

```{r}
predictor<-x%*%vec_beta
time <- simu_simple_beta(x = x, predictor = predictor, slam=slam)

if (censoring==0) {
  df_survival <- data.frame(time = time,
                            event = 1)
}else{
  clam<-exp(predictor)*slam*censoring/(1-censoring)
  csg<-rexp(length(vec_beta),rate=clam)
  dim(csg)<-c(length(vec_beta),1)
  csgind<-(time<=csg)
  otime<-time*csgind+csg*(1-csgind)

  df_survival <- data.frame(time = otime,
                            event = as.numeric(csgind))
}

simu_simple_beta <- function(predictor,
                             slam = 0.005,
                             x){
  nb_observations <- nrow(x)
  randu<-runif(nb_observations,min=0,max=1)
  dim(randu)<-c(nb_observations,1)
  time<-exp(-predictor)*(-log(1-randu))/slam
  return(time)
}

```

### LV

```{r}
tbeta<-rnorm(ng,mean=bmean,sd=bsd)
predictor<-x%*%tbeta
randu<-runif(ns,min=0,max=1)
dim(randu)<-c(ns,1)
slam=0.005
time<-exp(-predictor)*(-log(1-randu))/slam
if (cp==0) {
  otime<-time
  csgind<-rep(1,ns)
}else{
  clam<-exp(predictor)*slam*cp/(1-cp)
  csg<-rexp(ns,rate=clam)
  dim(csg)<-c(ns,1)
  csgind<-(time<=csg)
  otime<-time*csgind+csg*(1-csgind)
}
```

### Comparison

Same implementation

### 2023-06-07 reporting/discordance LV TF.Rmd
update results

@@
 date: "`r Sys.Date()`"
-output: html_document
+output:
+  bookdown::github_document2:
+    toc: TRUE
 ---
@@
 
-```{r eval=TRUE, fig.cap="Original plot", fig.height=20, include = TRUE, out.height=700, out.width=700}
-knitr::include_graphics("../images/comparison_methods6-1.pdf")
+# Plots
+
+## Draft plot
+
+```{r eval=TRUE, fig.cap="Original plot", fig.height=20, include = TRUE, out.height=700, out.width=700, echo=FALSE}
+knitr::include_graphics("images/comparison_methods6-1.png")
 ```
 
-```{r eval=TRUE, fig.cap="New plot (fails to replicate)", fig.height=20, include = TRUE, out.height=700, out.width=700}
-knitr::include_graphics("../images/plot_original.pdf")
+# New plot (fails to replicate)
+
+```{r eval=TRUE, fig.cap="New plot (fails to replicate)", fig.height=20, include = TRUE, out.height=700, out.width=700, echo=FALSE}
+knitr::include_graphics("images/plot_original.png")
 ```
@@
 ```{r}
+if (params[ite,1]==0.1){
+  gsp=0.3
+}
+if (params[ite,1]==0.3){
+  gsp=0.4
+}
+
+if (params[ite,1]==0.5){
+  gsp=0.5
+}
+gsize<-50 #gene set size
+#gsp<-agsp[sigp]
+ngs<-gsize*gsp
 tau <- 0.5
@@
 -   The variance defined in LV when beta != 1 corresponds to type == "C" in TF (because tau = 0.5)
-
-It makes sense regarding the graph because results from type == "C" are mostly similar. Problem identified !!
+- The proportion of significant gene does not match the ones in the paper
 

### 2023-06-07 reporting/discordance LV TF.Rmd
update results

@@
 
-```{r eval=TRUE, fig.cap="Original plot", fig.height=20, include = TRUE, out.height=700, out.width=700, echo=FALSE}
+```{r eval=TRUE, fig.cap="Original plot", fig.height=6, echo=FALSE}
 knitr::include_graphics("images/comparison_methods6-1.png")
@@
 
-```{r eval=TRUE, fig.cap="New plot (fails to replicate)", fig.height=20, include = TRUE, out.height=700, out.width=700, echo=FALSE}
+```{r eval=TRUE, fig.cap="New plot (fails to replicate)", fig.height=6, echo=FALSE}
 knitr::include_graphics("images/plot_original.png")

### 2023-06-07 script/comparaisons_sets_bis_corr_beta6_2.R
upload another old lv

@@ -1,0 +1,233 @@
+
+library(mvtnorm)
+library(GBJ)
+library(survival)
+library(globaltest)
+library(globalboosttest)
+ite_base=as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))+999+999
+ite=(ite_base%%48)+1
+
+if (ite_base<=1920){
+ns=50
+}
+
+if (ite_base>1920 & ite_base<=3840){
+ns=80
+}
+
+
+
+params=read.table("data/simulations_table.txt",header=TRUE)
+if (params[ite,1]==0.1){
+gsp=0.05
+}
+if (params[ite,1]==0.3){
+gsp=0.1
+}
+
+if (params[ite,1]==0.5){
+gsp=0.2
+}
+
+cp=params[ite,3]
+corr=params[ite,2]
+#corr=4
+beta=params[ite,4] #beta type: 1(positively) or 2(random)
+#corr=1 #correlation type: 1(independence), 2(equal), 3(AR), 4(unstructured)
+cenp=1 #censoring type: 1(no), 2(30%)
+sigp=1 #significant gene proportion: 1(10%), 2(30%), 3(50%))
+tau<-0.5 #correlation coefficient between genes
+acp<-c(0.0,0.3) #censoring fraction
+agsp<-c(0.1,0.3,0.5) #significant gene set proportion
+nstat<-3 #test statistics of interest: GSEA with p=0,0.5,1, global test, and Adewale test
+ng<-200 #gene size
+#ns<-80 #sanple size
+B<-1000 #permuation
+iter<-50 #iteration
+gsize<-50 #gene set size
+#gsp<-agsp[sigp]
+ngs<-gsize*gsp
+ncov<-ngs*(ngs-1)/2
+ncov2<-ng*(ng-1)/2
+p.pvalue<-matrix(0,nrow=nstat,ncol=iter)
+ots<-matrix(0,nrow=nstat,ncol=iter)
+p5<-rep(0,nstat)
+cf<-rep(0,iter)
+
+stats2=function(data,ng,gsize,nstat){
+  time<-data[,1]
+  csg<-data[,2]
+  x<-data[,3:ncol(data)]
+  wtcx<-matrix(0,nrow=2,ncol=ng)
+  ucox<-rep(0,ng)
+  escx0<-0
+  escx1<-0
+  gt<-0
+  aw<-0
+  bst<-0
[…]

### 2023-06-21 reporting/simu_rembrandt_breast.Rmd
rename varcovar

---
title: "VarCovar Rembrandt"
author: "TF"
date: "`r Sys.Date()`"
output:
  bookdown::github_document2:
    toc: True
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(dplyr)
library(ggplot2)

source(file = "functions/fct_report_simu_setting.R")

dfRembrandt <- readRDS("data/rembrandt/dfresvarcovar.rds")
dfBreast <- readRDS("data/breast_cancer/dfresvarcovar.rds")
```

Each dot in the following figure corresponds to one pathway.

# Rembrandt

## Proportion of significant genes

```{r}
plot_R <- fct_prop_sign_genes(dfRembrandt) + ggtitle("Rembrandt")
plot_B <- fct_prop_sign_genes(dfBreast) + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

Choices :

- number of genes : $NG = \{10, 50, 200\}$
- proportion of significant genes : $p_g = \{0.2, 0.5\}$

## Beta

```{r}
lsbeta_R <- fct_beta_dist(dfRembrandt, caption = "Rembrandt : Proportion of positive, negative and non significant genes")
lsbeta_B <- fct_beta_dist(dfBreast, caption = "Breast cancer : Proportion of positive, negative and non significant genes")

lsbeta_R$table_beta
lsbeta_B$table_beta
```

Proportions are roughly similar between positive and negative genes.

```{r}
plot_R <- lsbeta_R$plot_unique_beta + ggtitle("Rembrandt")
plot_B <- lsbeta_B$plot_unique_beta + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

```{r}
plot_R <- lsbeta_R$plot_multiple_beta + ggtitle("Rembrandt")
plot_B <- lsbeta_B$plot_multiple_beta + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

Choices :

- $\beta \sim \mathcal{N}(0, 0.4^2)$
- half of genes follows $\beta \sim \mathcal{N}(-0.4, 0.2^2)$ and half follows $\beta \sim \mathcal{N}(0.4, 0.2^2)$
- half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$

## Variance

```{r}
plot_R <- fct_plot_variance(dfRembrandt) + ggtitle("Rembrandt")
plot_B <- fct_plot_variance(dfBreast, nb_filter_plot = 3) + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

Choice :

- Variance : $C_{jj} = 0.2$

## Correlation

```{r}
ls_R <- fct_plot_correlation(dfRembrandt, caption = "Rembrandt : Shapes of non-standard beta distribution by type of gene")
ls_B <- fct_plot_correlation(dfBreast, caption = "Breast : Shapes of non-standard beta distribution by type of gene")

plot_R <- ls_R$plot_unique + ggtitle("Rembrandt")
plot_B <- ls_B$plot_unique + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

```{r}
ls_R$table_shapes_correlation
ls_B$table_shapes_correlation
```

```{r}
plot_R <- ls_R$plot_multiple + ggtitle("Rembrandt")
plot_B <- ls_B$plot_multiple + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

Choice :

- Overall correlation follows a non-standard beta $corr \sim NSBeta(20, 20, min=-1, max =1)$
- Correlation between significant genes $corr \sim NSBeta(10, 10, min=-1, max =1)$ and between other genes $corr \sim NSBeta(25, 25, min=-1, max =1)$

## Summary Rembrandt

Taking paper notations :

- number of genes : $NG = \{10, 50, 200\}$
- proportion of significant genes : $p_g = \{0.2, 0.5\}$
- Variance : $C_{jj} = 0.2$
- Type (D) : $\beta \sim \mathcal{N}(0, 0.4^2)$
- Type (E) : half of genes follows $\beta \sim \mathcal{N}(-0.4, 0.2^2)$ and half follows $\beta \sim \mathcal{N}(0.4, 0.2^2)$
- Type (F) : half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$
- Case (IV) : Overall correlation follows a non-standard beta $corr \sim NSBeta(20, 20, min=-1, max =1)$
- Case (V) : Correlation between significant genes $corr \sim NSBeta(10, 10, min=-1, max =1)$ and between other genes $corr \sim NSBeta(25, 25, min=-1, max =1)$

Nb of simulations to perform = $3*2*3*2 = 36$

### 2023-06-21 script/script_simulation/figures_simu_rembrandt_breast_v1.R
update with new results

@@ -59,19 +59,24 @@ fct_generate_fig1 <- function(data,
 
 ##### data #####
 
-dfScenario <- readRDS("data/dfScenario_breast_rembrandt.rds") %>%
+dfScenario1 <- readRDS("data/dfScenario_breast_rembrandt.rds") %>%
   tibble::rowid_to_column(var = "hp_row")
 
-dfScenario %>% filter(nb_genes == 10)
+dfScenario2 <- readRDS("data/dfScenario_breast_rembrandt_to_finish.rds") %>%
+  tibble::rowid_to_column(var = "hp_row")
 
-dfres <- list.files("data/result_simu_rembrandt_breast/", recursive = T, full.names = T) %>%
+dfres1 <- list.files("data/result_simu_rembrandt_breast/result_9799043/", recursive = T, full.names = T) %>%
   lapply(FUN = readRDS) %>%
   bind_rows() %>%
-  inner_join(dfScenario, by = "hp_row")
+  inner_join(dfScenario1, by = "hp_row")
 
-dfScenario %>%
-  select(-hp_row) %>%
-  summarise_all(.funs = function(x) list(unique(x)))
+dfres2 <- list.files("data/result_simu_rembrandt_breast/result_9839694/", recursive = T, full.names = T) %>%
+  lapply(FUN = readRDS) %>%
+  bind_rows() %>%
+  inner_join(dfScenario2, by = "hp_row")
+
+dfres <- bind_rows(dfres1, dfres2)
+
 
 ##### plot #####
 vec_labels_type <- paste0("Type ", c("A", "B", "C", "D", "E", "F"))
@@ -81,7 +86,8 @@ names(vec_labels_case) <- c(1:length(vec_labels_case))
 
 dfres %>%
   mutate(case_type = paste0("Case : ", case, " ; Type : ", type),
-         n_p = paste0("N = ", nb_observations, " ; NG = ", nb_genes)) %>%
+         n_p = paste0("N = ", nb_observations, " ; NG = ", nb_genes),
+         prop_sig_gene = as.factor(prop_sig_gene)) %>%
   mutate(p_val_sig = p_value < 0.05) %>%
   group_by(hp_row, method, prop_sig_gene, case_type, censoring, n_p) %>%
   summarise(power = sum(p_val_sig)/n()) %>%
@@ -95,21 +101,28 @@ dfres %>%
                                     "sGBJ"))) %>%
   ggplot(mapping = aes(x = prop_sig_gene,
                        y = power,
[…]

### 2023-06-21 reporting/simu_rembrandt_breast.Rmd
update results with simulations

@@
 
-# Rembrandt
+# Introduction
+
+Document is separated into two parts
+
+- First part goal is to find relevant simulation setting from Rembrandt and Breast cancer study.
+- Second part will present the results from the simulation studies
+
+# Finding the simulation setting
 
@@
 Nb of simulations to perform = $3*2*3*2 = 36$
+
+# Results of simulation study
+
+```{r fig.height=10, fig.cap="Statistical power depending on simulation setting. N is the number of patients and NG the number of genes in the pathway."}
+source("script/script_simulation/figures_simu_rembrandt_breast_v1.R")
+
+plot_simu
+```
+
+Overall, all methods have close statistical power but sGBJ and Global boost test tends to be among the worst methods. There are some exceptions to note :
+
+- Global Boost Test performs well in type F
+- sGBJ performs well when N = 100 and NG = 50 which might be related to its asymptotic nature.
+- In other cases, Wald and Global Test slightly outperformed other methods and Wald seems better than Global Test.
+
+## Comprehend tests
+
+see : https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3196970/
+
+### Wald test
+
+- estimate one beta by gene using cox model
+- sum the wald test statistics
+- H0 : variance of beta is equal to 0.
+
+### Global test
+
+Jelle J.
+Goeman and others, Testing association of a pathway with survival using gene expression data, Bioinformatics, Volume 21, Issue 9, May 2005, Pages 1950–1957, https://doi.org/10.1093/bioinformatics/bti267
+
+- estimate one beta by gene using cox model
+- makes the assumption that beta are centered on 0
+- H0 : variance of beta is equal to 0.
+
+### Global boost test
+
+Compute the cox model negative likelihood using a boosted multivariate (all genes) cox model. P-value is computed by permutation on the negative likelihood.
+
+### Thinking
+
+So basically, Wald and Global are super similar, Global test would have been expected to perform better in D case because the hypothesis on beta distribution is closer.
+One explanation could be that this test is optimal for small deviations from null hypothesis.
+This might explain why is seems better than Wald where power is low and why Wald is better when power is high.
+
+Global boost test relies on one model for all the data contrary to other models that perform one model by gene which might makes it more adapted to large number of genes.
+This is partially in line with good performance when N = 50.
+
+sGBJ is similar to Wald and Global test in the sense that it is based on one cox model by feature.
+However, it takes into account the covariance of genes in its formula.
+And the data are simulated with no covariance of beta => Problem ?
+look at GBJ paper

### 2023-06-21 script/script_simulation/figures_simu_rembrandt_breast_v1.R

@@ -75,7 +75,21 @@ dfres2 <- list.files("data/result_simu_rembrandt_breast/result_9839694/", recurs
   bind_rows() %>%
   inner_join(dfScenario2, by = "hp_row")
 
-dfres <- bind_rows(dfres1, dfres2)
+dfres <- bind_rows(dfres1, dfres2) %>%
+  mutate(case_type = paste0("Case : ", case, " ; Type : ", type),
+         n_p = paste0("N = ", nb_observations, " ; NG = ", nb_genes),
+         prop_sig_gene = as.factor(prop_sig_gene)) %>%
+  mutate(p_val_sig = p_value < 0.05) %>%
+  group_by(hp_row, method, prop_sig_gene, case_type, case, type, censoring, n_p, nb_observations, nb_genes) %>%
+  summarise(power = sum(p_val_sig)/n(), .groups = "drop") %>%
+  mutate(censoring = factor(censoring,
+                            levels = c(0, 0.3),
+                            labels = c("0%", "30%")),
+         method = factor(method,
+                         levels = c("Global Boost Test",
+                                    "Wald Test",
+                                    "Global Test",
+                                    "sGBJ")))
 
 
 ##### plot #####
@@ -84,21 +98,7 @@ names(vec_labels_type) <- c("A", "B", "C", "D", "E", "F")
 vec_labels_case <- paste0("Case (", c("I", "II", "III", "IV", "V"), ")")
 names(vec_labels_case) <- c(1:length(vec_labels_case))
 
-dfres %>%
-  mutate(case_type = paste0("Case : ", case, " ; Type : ", type),
-         n_p = paste0("N = ", nb_observations, " ; NG = ", nb_genes),
-         prop_sig_gene = as.factor(prop_sig_gene)) %>%
-  mutate(p_val_sig = p_value < 0.05) %>%
-  group_by(hp_row, method, prop_sig_gene, case_type, censoring, n_p) %>%
-  summarise(power = sum(p_val_sig)/n()) %>%
-  mutate(censoring = factor(censoring,
-                            levels = c(0, 0.3),
-                            labels = c("0%", "30%")),
-         method = factor(method,
-                         levels = c("Global Boost Test",
-                                    "Wald Test",
-                                    "Global Test",
-                                    "sGBJ"))) %>%
+plot_simu <- dfres %>%
   ggplot(mapping = aes(x = prop_sig_gene,
                        y = power,
                        linetype = censoring,
[…]

### 2023-07-04 reporting/simu_rembrandt_breast.Rmd
fix bug retrieve previous results

@@
 - number of genes : $NG = \{10, 50, 200\}$
+- number of individuals : $N = \{50, 100\}$
 - proportion of significant genes : $p_g = \{0.2, 0.5\}$
+- proportion of censored genes : $c = \{0, 0.3\}$
 - Variance : $C_{jj} = 0.2$
@@
 
-Nb of simulations to perform = $3*2*3*2 = 36$
+Nb of simulations to perform = $3*2*2*2*3*2 = 144$
 
@@
 And the data are simulated with no covariance of beta => Problem ?
-look at GBJ paper

### 2023-07-04 functions/rf_permutation.R

@@ -40,7 +40,7 @@ rfpermutation=function(dfdata, nb_permutation = 10, ntree = 10){
     unlist()
   
     
-  p.pvalue <- mean(oob_error < vec_ooberror)
+  p.pvalue <- mean(oob_error > vec_ooberror)
   
-  return(ts)
+  return(p.pvalue)
 }

### 2023-07-04 functions/fct_generate_beta.R
reuse gene correlation to generate beta

@@ -5,13 +5,15 @@
 #' @param type The scenario type either A, B or C
 #' @param nb_genes The number of genes
 #' @param prop_sig_gene The proportion of significant genes
+#' @param corr_mat The correlation matrix of beta
 #'
 #' @return A vector of the beta coefficients
 #' @export
 #' 
 fct_generate_beta <- function(type,
                               nb_genes,
-                              prop_sig_gene){
+                              prop_sig_gene,
+                              corr_mat = NULL){
   
   nb_sig_gene = round(prop_sig_gene*nb_genes)
   
@@ -49,8 +51,42 @@ fct_generate_beta <- function(type,
              rnorm(n = floor(nb_sig_gene/2), mean = 0.8, sd = 0.4),
              rep(0, nb_genes-nb_sig_gene))
     
+  } else if(type == "G"){
+    vec_sd <- c(rep(0.4, nb_sig_gene),
+                rep(0, nb_genes-nb_sig_gene))
+    # Compute the covariance matrix
+    mat_var_covar <- diag(vec_sd) %*% corr_mat %*% diag(vec_sd)
+    
+    vec_mu <- rep(0, nb_genes)
+    
+    res <- MASS::mvrnorm(n = 1, mu = vec_mu, Sigma = mat_var_covar)
+    
+  } else if(type == "H"){
+    vec_sd <- c(rep(0.4, nb_sig_gene),
+                rep(0, nb_genes-nb_sig_gene))
+    # Compute the covariance matrix
+    mat_var_covar <- diag(vec_sd) %*% corr_mat %*% diag(vec_sd)
+    
+    vec_mu <- c(rep(-0.4, ceiling(nb_sig_gene/2)),
+                rep(0.4, floor(nb_sig_gene/2)),
+                rep(0, nb_genes-nb_sig_gene))
+    
+    res <- MASS::mvrnorm(n = 1, mu = vec_mu, Sigma = mat_var_covar)
+    
+  } else if(type == "I"){
+    vec_sd <- c(rep(0.4, nb_sig_gene),
+                rep(0, nb_genes-nb_sig_gene))
+    # Compute the covariance matrix
+    mat_var_covar <- diag(vec_sd) %*% corr_mat %*% diag(vec_sd)
+    
+    vec_mu <- c(rep(-0.8, ceiling(nb_sig_gene/2)),
+                rep(0.8, floor(nb_sig_gene/2)),
+                rep(0, nb_genes-nb_sig_gene))
+    
[…]

### 2023-07-04 functions/fct_generate_data.R

@@ -34,12 +34,13 @@ fct_generate_data <- function(case,
   
   vec_beta <- fct_generate_beta(type = type,
                                 prop_sig_gene = prop_sig_gene,
-                                nb_genes = nb_genes)
+                                nb_genes = nb_genes,
+                                corr_mat = mat_var_covar$corr_mat)
   
   ### generate genes values
   x<-mvtnorm::rmvnorm(n=nb_observations,
                       mean=rep(0,nb_genes),
-                      sigma=mat_var_covar,
+                      sigma=mat_var_covar$mat_var_covar,
                       method="chol")
   colnames(x) <- paste0("gene", 1:ncol(x))
   

### 2023-07-04 functions/fct_generate_varcovar.R

@@ -14,6 +14,8 @@ fct_generate_varcovar <- function(case,
                                   prop_sig_gene,
                                   nb_genes,
                                   variance){
+  corr_mat <- NULL
+  
   nb_sig_gene = round(prop_sig_gene * nb_genes)
   
   if(case == 1){
@@ -83,7 +85,8 @@ fct_generate_varcovar <- function(case,
   # set the variance
   diag(mat_var_covar) <- variance
   
-  return(mat_var_covar)
+  return(list(mat_var_covar = mat_var_covar,
+              corr_mat = corr_mat))
 }
 
 #' fct_impute_nsbeta

### 2023-07-04 functions/fct_simulation_paper.R
strengthen the log

@@ -121,7 +121,19 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "other", "rf"),
     method = names(ls_pvalue)
   ) %>%
     na.omit() %>%
-    tibble::remove_rownames()
+    tibble::remove_rownames() %>%
+    mutate(methods = paste0(methods, collapse = ", "),
+           nb_observations = nb_observations,
+           nb_genes = nb_genes,
+           prop_sig_gene = prop_sig_gene,
+           variance = variance,
+           case = case,
+           type = type,
+           censoring = censoring,
+           nb_permutation = nb_permutation,
+           prop_two_periods = prop_two_periods,
+           nperm_sGBJ = nperm_sGBJ,
+           slam = slam)
   
   return(res)
 }

### 2023-07-04 script/script_simulation/meta_parameters_remb_breast.R
add correlation metaparam

@@ -36,3 +36,15 @@ readRDS(file = "data/dfScenario_breast_rembrandt.rds") %>%
   filter(!rowid %in% vec_done) %>%
   select(-rowid) %>%
   saveRDS(file = "data/dfScenario_breast_rembrandt_to_finish.rds")
+
+##### new set exp with correlated beta
+expand.grid(nb_observations = c(50,100),
+            nb_genes = c(10,50),
+            prop_sig_gene = c(0.2),
+            variance = 0.2,
+            case = c(4, 5),
+            type = c("G", "H", "I"),
+            censoring = c(0.3),
+            nb_permutation = 1000,
+            prop_two_periods = FALSE) |>
+  saveRDS(file = "data/dfScenarioCorr.rds")

### 2023-07-04 functions/fct_simulation_paper.R
separate tests and compute time

@@ -14,7 +14,7 @@
 #'
 #' @return A dataframe with the different p-values
 #' @export
-fct_simulation_paper <- function(methods = c("sgbj", "gbt", "other", "rf"),
+fct_simulation_paper <- function(methods = c("sgbj", "gbt", "gt", "wald", "rf"),
                                  nb_observations = 50,
                                  nb_genes = 50,
                                  prop_sig_gene = 0.2,
@@ -34,6 +34,9 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "other", "rf"),
       "Global Boost Test" = NA,
       "RF" = NA
     )
+  
+  ls_time <- ls_pvalue
+  
   ##### generate data #####
   dfdata <- fct_generate_data(
     case = case,
@@ -51,18 +54,22 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "other", "rf"),
   
   ##### sGBJ #####
   if("sgbj" %in% methods){
+    start_time <- Sys.time()
     resSGBJ <-
       sGBJ::sGBJ(
         surv = survival::Surv(dfdata$time, dfdata$event),
         factor_matrix = dfdata %>% select(-time,-event) %>% as.matrix(),
         nperm = nperm_sGBJ
       )$sGBJ_pvalue
+    end_time <- Sys.time()
     
     ls_pvalue$sGBJ <- resSGBJ
+    ls_time$sGBJ <- as.numeric(end_time - start_time)
   }
   
   #Global boost test
   if("gbt" %in% methods){
+    start_time <- Sys.time()
     gbst <-
       globalboosttest::globalboosttest(
         X = dfdata %>% select(-time,-event) %>% as.matrix(),
@@ -72,52 +79,52 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "other", "rf"),
         mstop = 500,
         pvalueonly = TRUE
       )$pvalue[1]
+    end_time <- Sys.time()
     
     ls_pvalue$`Global Boost Test` <- gbst
+    ls_time$`Global Boost Test` <- as.numeric(end_time - start_time)
   }
   
-  ##### Other methods #####
-  if("other" %in% methods){
-    ots <- statsTest(
-      vectime = dfdata$time,
-      vecevent = dfdata$event,
-      x = dfdata %>% select(-time,-event) %>% as.matrix()
-    )
-    rts <- matrix(0, nrow = 2, ncol = nb_permutation)
+  #Wald test
[…]

### 2023-07-04 functions/statsTest.R

@@ -32,3 +32,70 @@ statsTest=function(vectime, vecevent, x){
   ts=c(gt,aw)
   return(ts)
 }
+
+
+
+#' pre_Wald_Test
+#'
+#' @description Function used for wald test computation
+#'
+#' @param vectime the time
+#' @param vecevent the event
+#' @param x the genes experession
+#'
+#' @return A vector with the statistical test of global test, adewale test
+#' @export
+#'
+pre_Wald_Test=function(vectime, vecevent, x){
+  # baseline
+  survObj <- survival::Surv(vectime,vecevent)
+  ucox <- lapply(X = 1:ncol(x),
+         FUN = function(x){
+           ocx<-survival::coxph(survObj~x[,i])
+           res<-ocx$coefficients/sqrt(ocx$var)
+           return(res)
+         }) %>%
+    unlist()
+  
+  res <- sum(ucox^2)
+  
+  return(res)
+}
+
+#' Wald_Test_Perm
+#'
+#' @description Function which compute Wald test
+#'
+#' @param vectime the time
+#' @param vecevent the event
+#' @param x the genes experession
+#' @param nb_permutation Number of permutation
+#'
+#' @return A vector with the statistical test of global test, adewale test
+#' @export
+#'
+Wald_Test_Perm=function(vectime, vecevent, x, nb_permutation){
+  # baseline
+  baseline_ucox <- pre_Wald_Test(vectime = vectime,
+                                 vecevent = vecevent,
+                                 x = x)
+  
+  # permutation
+  permut_ucox <- lapply(X = 1:nb_permutation,
+                        FUN = function(b){
+                          id_sample <- sample(1:length(vectime), replace = FALSE)
+                          vectime_perm <- vectime[id_sample]
+                          vecevent_perm <- vecevent[id_sample]
+                          
+                          res <- pre_Wald_Test(vectime = vectime_perm,
+                                               vecevent = vecevent_perm,
+                                               x = x)
+                          return(res)
+                        })
+    
+  res <- mean(permut_ucox > baseline_ucox)
+  
+  return(res)
+}
+
+

### 2023-07-04 functions/fct_simulation_paper.R
fix bugs

@@ -101,6 +101,7 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "gt", "wald", "rf"),
   #Global Test
   if("gt" %in% methods){
     start_time <- Sys.time()
+    survObj <- survival::Surv(dfdata$time,dfdata$event)
     ogt<-globaltest::gt(survObj,
                         x,
                         model="cox",

### 2023-07-04 functions/statsTest.R

@@ -50,7 +50,7 @@ pre_Wald_Test=function(vectime, vecevent, x){
   # baseline
   survObj <- survival::Surv(vectime,vecevent)
   ucox <- lapply(X = 1:ncol(x),
-         FUN = function(x){
+         FUN = function(i){
            ocx<-survival::coxph(survObj~x[,i])
            res<-ocx$coefficients/sqrt(ocx$var)
            return(res)

### 2023-07-04 script/script_simulation/curta/05_simu_from_data_corr.R
prepare slurm script

@@ -1,0 +1,61 @@
+print(paste0("Start at ", Sys.time(), "..."))
+
+slar_taskid <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+# slar_taskid <- 1
+slar_jobid <- as.numeric(Sys.getenv("SLURM_ARRAY_JOB_ID"))
+
+
+##### load packages and functions
+message("load packages and functions")
+
+library(dplyr)
+
+lapply(list.files(path = "functions/", full.names = TRUE),
+       source)
+
+##### load data
+message("load dfScenario")
+
+dfScenario <- readRDS("data/dfScenarioCorr.rds")
+hp_row <- slar_taskid
+vec_hp <- dfScenario[hp_row,]
+
+##### start computation
+message("start computation")
+
+res <- pbapply::pblapply(1:500,
+                         FUN = function(iter){
+                           res <- fct_simulation_paper(methods = c("sgbj", "gbt", "gt", "wald"),
+                                                       nb_observations = vec_hp[["nb_observations"]],
+                                                       nb_genes = vec_hp[["nb_genes"]],
+                                                       prop_sig_gene =  vec_hp[["prop_sig_gene"]],
+                                                       variance =  vec_hp[["variance"]],
+                                                       case =  vec_hp[["case"]],
+                                                       type =  vec_hp[["type"]],
+                                                       censoring =  vec_hp[["censoring"]],
+                                                       nb_permutation =  vec_hp[["nb_permutation"]],
+                                                       prop_two_periods =  vec_hp[["prop_two_periods"]],
+                                                       nperm_sGBJ = vec_hp[["nb_permutation"]]) %>%
+                             mutate(hp_row = hp_row,
+                                    iter = iter)
+                           
+                           return(res)
+                         }) %>%
+  bind_rows()
+
+##### save results
+print(paste0("Saving start at ", Sys.time(), "..."))
+
+subDir <- paste0("data/result_simu/result_", slar_jobid)
+if (!file.exists(subDir)){
+  dir.create(subDir)
+}
+
+date_str <- gsub(x = Sys.Date(),"-","")
+
+filename_r <- paste0(subDir, "/result_job",
+                     slar_taskid, "_", date_str, ".rds")
+
+saveRDS(res, file = filename_r)
+
+print(paste0("Saving end at ", Sys.time(), "..."))

### 2023-09-21 reporting/simu_rembrandt_breast.Rmd
update with new simu

--- 2023-06-21 reporting/simu_rembrandt_breast.Rmd
@@
 - number of genes : $NG = \{10, 50, 200\}$
+- number of individuals : $N = \{50, 100\}$
 - proportion of significant genes : $p_g = \{0.2, 0.5\}$
+- proportion of censored genes : $c = \{0, 0.3\}$
 - Variance : $C_{jj} = 0.2$
+
 - Type (D) : $\beta \sim \mathcal{N}(0, 0.4^2)$
@@
 - Type (F) : half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$
+
+**Note : scenario D, E and F do not rely on correlation matrix to simulate variance covariance matrix so the those matrix are unlikely to be positive definite.**
+
+- Type (G), (H) and (I) : same as type (D), (E), (F) where covariance matrix of $\beta$ is obtained from the correlation matrix of gene expression defined by "Case" (instead of no correlation).
+
 - Case (IV) : Overall correlation follows a non-standard beta $corr \sim NSBeta(20, 20, min=-1, max =1)$
@@
 
-Nb of simulations to perform = $3*2*3*2 = 36$
+**Note : Case IV and V are not guaranteed to generate positive definite matrix, after the correlation values are sampled, if the matrix is not positive definite, the nearest positive definite matrix is computed using the algorithm defined by Higham 2002 (see Matrix::nearPD).**
+
+- Case (V) : Correlation between significant genes is $0.2$, correlation with non significant genes is $0$.
 
@@
 source("script/script_simulation/figures_simu_rembrandt_breast_v1.R")
+
+plot_simu
+```
+
+**Note : see methods section, the covariance matrix are unlikely to be positive definite.**
+
+```{r fig.height=10, fig.cap="Statistical power depending on simulation setting (second batch). N is the number of patients and NG the number of genes in the pathway."}
+source("script/script_simulation/figures_simu_rembrandt_breast_v2.R")
 

### 2023-09-21 functions/fct_generate_varcovar.R

@@ -50,10 +50,9 @@ fct_generate_varcovar <- function(case,
   } else if(case == 4){
     variances <- rep(variance, nb_genes)
     
-    corr_mat <- fct_impute_nsbeta(variance = variance,
-                      nb_genes = nb_genes,
-                      shape1 = 20,
-                      shape2 = 20)
+    corr_mat <- fct_impute_nsbeta(nb_genes = nb_genes,
+                                  shape1 = 20,
+                                  shape2 = 20)
     
     if(!matrixcalc::is.positive.definite(corr_mat)){
       corr_mat <- Matrix::nearPD(corr_mat, corr = TRUE, base.matrix = TRUE)$mat
@@ -65,13 +64,11 @@ fct_generate_varcovar <- function(case,
   } else if(case == 5){
     variances <- rep(variance, nb_genes)
     
-    corr_mat <- fct_impute_nsbeta(variance = variance,
-                                  nb_genes = nb_genes,
+    corr_mat <- fct_impute_nsbeta(nb_genes = nb_genes,
                                   shape1 = 25,
                                   shape2 = 25)
     
-    corr_mat_sig <- fct_impute_nsbeta(variance = variance,
-                                      nb_genes = nb_sig_gene,
+    corr_mat_sig <- fct_impute_nsbeta(nb_genes = nb_sig_gene,
                                       shape1 = 10,
                                       shape2 = 10)
     
@@ -84,8 +81,18 @@ fct_generate_varcovar <- function(case,
     # Compute the covariance matrix
     mat_var_covar <- diag(sqrt(variances)) %*% corr_mat %*% diag(sqrt(variances))
     
+  } else if(case == 6){
+    
+    ls_cor <- fct_correlation_matrix_simple(nb_genes = nb_genes,
+                                            nb_sig_gene = nb_sig_gene,
+                                            variance = variance,
+                                            correlation_value_sig = 0.2,
+                                            correlation_value_non_sig = 0)
+    mat_var_covar <- ls_cor$mat_var_covar
+    corr_mat <- ls_cor$corr_mat
+    
   } else {
-    stop("case must be between 1 and 5")
+    stop("case must be between 1 and 7")
   }
   
   # set the variance
@@ -103,20 +110,15 @@ fct_generate_varcovar <- function(case,
 #' @description Build a correlation matrix with nsbeta law
 #'
 #' @param nb_genes The number of genes
-#' @param variance The variance (diagonal of the matrix)
[…]

### 2023-09-21 script/script_simulation/figures_simu_rembrandt_breast_corr.R

@@ -1,0 +1,114 @@
+##### load package #####
+library(dplyr)
+library(ggplot2)
+library(stringr)
+
+##### functions #####
+fct_generate_fig1 <- function(data,
+                              subtitle = "(A) Non proportional hazard",
+                              confint = FALSE){
+  vec_labels_type <- paste0("Type ", c("A", "B", "C", "D", "E", "F"))
+  names(vec_labels_type) <- c("A", "B", "C", "D", "E", "F")
+  vec_labels_case <- paste0("Case (", c("I", "II", "III", "IV", "V"), ")")
+  names(vec_labels_case) <- c(1:length(vec_labels_case))
+  
+  res <- data %>%
+    mutate(p_val_sig = p_value < 0.05) %>%
+    group_by(hp_row, method, prop_sig_gene, case, type, censoring) %>%
+    summarise(success = sum(p_val_sig),
+              trials = n()) %>%
+    # wait for case 41
+    filter(!is.na(success)) %>%
+    ungroup() %>%
+    mutate(power = binom::binom.confint(x = success, n = trials, methods = "exact")$mean,
+           lower = binom::binom.confint(x = success, n = trials, methods = "exact")$lower,
+           upper = binom::binom.confint(x = success, n = trials, methods = "exact")$upper) %>%
+    mutate(censoring = factor(censoring,
+                              levels = c(0, 0.3),
+                              labels = c("0%", "30%")),
+           method = factor(method,
+                           levels = c("Global Boost Test",
+                                      "Wald Test",
+                                      "Global Test",
+                                      "sGBJ"))) %>%
+    ggplot(mapping = aes(x = prop_sig_gene,
+                         y = power,
+                         color = method,
+                         lty = censoring)) +
+    geom_point() +
+    geom_line(size = 1) +
+    scale_color_viridis_d() +
+    facet_grid(case ~ type,
+               labeller = labeller(case = vec_labels_case,
+                                   type = vec_labels_type)) +
+    labs(x = "Proportion of significant genes",
+         y = "Statistical Power",
+         color = "Method",
+         lty = "Censoring proportion",
+         subtitle = subtitle) +
+    lims(y = c(0, 1))
+  
+  if(confint){
+    res <- res +
+      ggplot2::geom_errorbar(mapping = aes(ymin = lower, ymax = upper),
+                             width = 0)
+  }
+  
[…]

### 2023-09-21 script/script_simulation/figures_simu_rembrandt_breast_v2.R

@@ -1,0 +1,55 @@
+##### load package #####
+library(dplyr)
+library(ggplot2)
+##### data #####
+
+dfres <- list.files("data/result_10187723/", recursive = T, full.names = T) %>%
+  lapply(FUN = readRDS) %>%
+  bind_rows() %>%
+  mutate(case_type = paste0("Case : ", case, " ; Type : ", type),
+         n_p = paste0("N = ", nb_observations, " ; NG = ", nb_genes),
+         prop_sig_gene = as.factor(prop_sig_gene)) %>%
+  mutate(p_val_sig = p_value < 0.05) %>%
+  group_by(hp_row, method, prop_sig_gene, case_type, case, type, censoring, n_p, nb_observations, nb_genes) %>%
+  summarise(power = sum(p_val_sig)/n(), .groups = "drop") %>%
+  mutate(censoring = factor(censoring,
+                            levels = c(0, 0.3),
+                            labels = c("0%", "30%")),
+         method = factor(method,
+                         levels = c("Global Boost Test",
+                                    "Wald Test",
+                                    "Global Test",
+                                    "sGBJ")))
+
+
+##### plot #####
+vec_labels_type <- paste0("Type ", c("A", "B", "C", "D", "E", "F"))
+names(vec_labels_type) <- c("A", "B", "C", "D", "E", "F")
+vec_labels_case <- paste0("Case (", c("I", "II", "III", "IV", "V"), ")")
+names(vec_labels_case) <- c(1:length(vec_labels_case))
+
+plot_simu <- dfres %>%
+  ggplot(mapping = aes(x = prop_sig_gene,
+                       y = power,
+                       linetype = censoring,
+                       color = censoring,
+                       group = method,
+                       fill = method)) +
+  geom_bar(stat = "identity",
+           position = "dodge") +
+  scale_fill_viridis_d() +
+  scale_color_manual(values = c("black", "red")) +
+  facet_grid(case_type ~ n_p,
+             labeller = labeller(case = vec_labels_case,
+                                 type = vec_labels_type)) +
+  labs(x = "Proportion of significant genes",
+       y = "Statistical Power",
+       fill = "Method",
+       color = "Censoring proportion",
+       lty = "Censoring proportion") +
+  lims(y = c(0, 1)) +
+  guides(fill = guide_legend(nrow = 4),
+         lty = guide_legend(nrow = 2),
+         color = guide_legend(nrow = 2)) +
+  theme_bw() +
+  theme(legend.position = "bottom")

### 2023-09-21 script/script_simulation/curta/06_simu_from_data_simple_corr.R
prepare next simu

@@ -1,0 +1,60 @@
+print(paste0("Start at ", Sys.time(), "..."))
+
+slar_taskid <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+# slar_taskid <- 1
+slar_jobid <- as.numeric(Sys.getenv("SLURM_ARRAY_JOB_ID"))
+
+
+##### load packages and functions
+message("load packages and functions")
+
+library(dplyr)
+
+lapply(list.files(path = "functions/", full.names = TRUE),
+       source)
+
+##### load data
+message("load dfScenario")
+
+dfScenario <- readRDS("data/dfScenarioSimpleCorr.rds") %>%
+  tibble::rowid_to_column(var = "hp_row")
+
+##### start computation
+message("start computation")
+
+res <- pbapply::pbapply(X = dfScenario,
+                        MARGIN = 1,
+                        FUN = function(row){
+                          fct_simulation_paper(methods = c("sgbj", "gbt", "gt", "wald"),
+                                               nb_observations = as.numeric(row[["nb_observations"]]),
+                                               nb_genes = as.numeric(row[["nb_genes"]]),
+                                               prop_sig_gene = as.numeric(row[["prop_sig_gene"]]),
+                                               variance =  as.numeric(row[["variance"]]),
+                                               case =  row[["case"]],
+                                               type =  row[["type"]],
+                                               censoring =  as.numeric(row[["censoring"]]),
+                                               nb_permutation =  as.numeric(row[["nb_permutation"]]),
+                                               prop_two_periods =  row[["prop_two_periods"]],
+                                               nperm_sGBJ = as.numeric(row[["nb_permutation"]])) %>%
+                            mutate(hp_row = row[["hp_row"]],
+                                   iter = slar_taskid) %>%
+                            return()
+                        }) %>%
+  bind_rows()
+
+##### save results
+print(paste0("Saving start at ", Sys.time(), "..."))
+
+subDir <- paste0("data/result_simu/result_", slar_jobid)
+if (!file.exists(subDir)){
+  dir.create(subDir)
+}
+
+date_str <- gsub(x = Sys.Date(),"-","")
+
+filename_r <- paste0(subDir, "/result_job",
+                     slar_taskid, "_", date_str, ".rds")
+
+saveRDS(res, file = filename_r)
+
+print(paste0("Saving end at ", Sys.time(), "..."))

### 2023-09-21 script/script_simulation/meta_parameters_remb_breast.R

@@ -48,3 +48,17 @@ expand.grid(nb_observations = c(50,100),
             nb_permutation = 1000,
             prop_two_periods = FALSE) |>
   saveRDS(file = "data/dfScenarioCorr.rds")
+
+##### new set exp with simple correlation setting
+expand.grid(nb_observations = c(50,100),
+            nb_genes = c(10,50),
+            prop_sig_gene = c(0.2),
+            variance = 0.2,
+            case = c(6),
+            type = c("G", "H", "I"),
+            censoring = c(0.3),
+            nb_permutation = 1000,
+            prop_two_periods = FALSE) |>
+  saveRDS(file = "data/dfScenarioSimpleCorr.rds")
+
+

### 2023-09-21 script/script_simulation/figures_simu_rembrandt_breast_v2.R
update report with new simulation

@@ -2,9 +2,15 @@
 library(dplyr)
 library(ggplot2)
 ##### data #####
+ls_folder_path <- c("data/result_11020779", "data/result_10187723")
 
-dfres <- list.files("data/result_10187723/", recursive = T, full.names = T) %>%
-  lapply(FUN = readRDS) %>%
+dfres <- lapply(ls_folder_path,
+                FUN = function(x) list.files(x, recursive = T, full.names = T)) %>%
+  unlist() %>%
+  lapply(FUN = function(x) readRDS(x) %>%
+           mutate(case = as.character(case),
+                  prop_two_periods = as.character(prop_two_periods),
+                  hp_row = as.character(hp_row))) %>%
   bind_rows() %>%
   mutate(case_type = paste0("Case : ", case, " ; Type : ", type),
          n_p = paste0("N = ", nb_observations, " ; NG = ", nb_genes),

### 2023-09-22 functions/fct_simulation_paper.R
Fix bug time computing problem

@@ -67,7 +67,7 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "gt", "wald", "rf"),
     end_time <- Sys.time()
     
     ls_pvalue$sGBJ <- resSGBJ
-    ls_time$sGBJ <- as.numeric(end_time - start_time)
+    ls_time$sGBJ <- as.numeric(difftime(end_time, start_time, units = "secs"))
   }
   
   #Global boost test
@@ -86,7 +86,7 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "gt", "wald", "rf"),
     end_time <- Sys.time()
     
     ls_pvalue$`Global Boost Test` <- gbst
-    ls_time$`Global Boost Test` <- as.numeric(end_time - start_time)
+    ls_time$`Global Boost Test` <- as.numeric(difftime(end_time, start_time, units = "secs"))
   }
   
   #Wald test
@@ -100,7 +100,7 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "gt", "wald", "rf"),
     end_time <- Sys.time()
     
     ls_pvalue$`Wald Test` <- wald_pval
-    ls_time$`Wald Test` <- as.numeric(end_time - start_time)
+    ls_time$`Wald Test` <- as.numeric(difftime(end_time, start_time, units = "secs"))
   }
   
   #Global Test
@@ -116,7 +116,7 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "gt", "wald", "rf"),
     end_time <- Sys.time()
     
     ls_pvalue$`Global Test` <- gt_pval
-    ls_time$`Global Test` <- as.numeric(end_time - start_time)
+    ls_time$`Global Test` <- as.numeric(difftime(end_time, start_time, units = "secs"))
   }
   
   #RF Test
@@ -127,7 +127,7 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "gt", "wald", "rf"),
     end_time <- Sys.time()
     
     ls_pvalue$RF <- p_val_rf
-    ls_time$RF <- as.numeric(end_time - start_time)
+    ls_time$RF <- as.numeric(difftime(end_time, start_time, units = "secs"))
   }
   
   res <- data.frame(

### 2023-09-22 script/script_simulation/curta/06_simu_from_data_simple_corr_compute_time.R

@@ -1,0 +1,39 @@
+##### load packages and functions
+library(dplyr)
+library(parallel)
+
+lapply(list.files(path = "functions/", full.names = TRUE),
+       source)
+
+##### load data
+dfScenario <- readRDS("data/dfScenarioSimpleCorr.rds") %>%
+  tibble::rowid_to_column(var = "hp_row")
+
+##### start computation
+parallel_fct <- function(row) {
+  fct_simulation_paper(
+    methods = c("sgbj", "gbt", "gt", "wald"),
+    nb_observations = as.numeric(row[["nb_observations"]]),
+    nb_genes = as.numeric(row[["nb_genes"]]),
+    prop_sig_gene = as.numeric(row[["prop_sig_gene"]]),
+    variance = as.numeric(row[["variance"]]),
+    case = row[["case"]],
+    type = row[["type"]],
+    censoring = as.numeric(row[["censoring"]]),
+    nb_permutation = as.numeric(row[["nb_permutation"]]),
+    prop_two_periods = row[["prop_two_periods"]],
+    nperm_sGBJ = as.numeric(row[["nb_permutation"]])
+  ) %>%
+    mutate(hp_row = row[["hp_row"]])
+}
+
+# Split the data into chunks for parallel processing
+chunks <- apply(dfScenario, 1, function(x) return(x), simplify = FALSE)
+
+# Parallel execution using 12 cores
+result_list <- mclapply(chunks, parallel_fct, mc.cores = nrow(dfScenario))
+
+# Combine the results into a single data frame
+res <- bind_rows(result_list)
+
+saveRDS(res, file = "data/result_11020779_time.rds")

### 2023-09-22 reporting/simu_rembrandt_breast.Rmd
update report with new simulation
Update report with new results

--- 2023-06-21 reporting/simu_rembrandt_breast.Rmd
@@
 
-Each dot in the following figure corresponds to one pathway.
-
 # Introduction
@@
 # Finding the simulation setting
+
+We used Rembrandt and breast cancer datasets. For each datasets we explored the parameters of interest to perform the simulation study.
+
+Each dot in the following figure corresponds to one pathway.
 
@@
 
-## Beta
+## Beta from Cox model
 
@@
 
-## Variance
+## Gene expression variance
 
@@
 
-## Correlation
+## Gene expression correlation
+
+### Overall gene expression correlation
 
@@
 
+### Gene expression correlation depending on gene significance
+
 ```{r}
@@
 
-## Summary Rembrandt
+## Summary Rembrandt and simulation settings
 
@@
 - number of genes : $NG = \{10, 50, 200\}$
+- number of individuals : $N = \{50, 100\}$
 - proportion of significant genes : $p_g = \{0.2, 0.5\}$
+- proportion of censored genes : $c = \{0, 0.3\}$
 - Variance : $C_{jj} = 0.2$
+
+<!-- **Note : scenario D, E and F do not rely on correlation matrix to simulate variance covariance matrix so the those matrix are unlikely to be positive definite.** -->
+
+### Case
+
+- Case 4 : Overall correlation follows a non-standard beta $corr \sim NSBeta(20, 20, min=-1, max =1)$
+- Case 5 : Correlation between significant genes $corr \sim NSBeta(10, 10, min=-1, max =1)$ and between other genes $corr \sim NSBeta(25, 25, min=-1, max =1)$
+
+**Note : Case IV and V are not guaranteed to generate positive definite matrix, after the correlation values are sampled, if the matrix is not positive definite, the nearest positive definite matrix is computed using the algorithm defined by Higham 2002 (see Matrix::nearPD).
+This correction was not implemented for scenario D, E and F.**
+
+- Case 6 : Correlation between significant genes is $0.2$, correlation with non significant genes is $0$.
+
+### Type
+
 - Type (D) : $\beta \sim \mathcal{N}(0, 0.4^2)$
@@
 - Type (F) : half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$
-- Case (IV) : Overall correlation follows a non-standard beta $corr \sim NSBeta(20, 20, min=-1, max =1)$
-- Case (V) : Correlation between significant genes $corr \sim NSBeta(10, 10, min=-1, max =1)$ and between other genes $corr \sim NSBeta(25, 25, min=-1, max =1)$
-
-Nb of simulations to perform = $3*2*3*2 = 36$
+- Type (G) : $\beta \sim \mathcal{N}(0, 0.4^2)$. Correlation matrix of $\beta$ is the same as correlation matrix of genes.
+- Type (H) : half of genes follows $\beta \sim \mathcal{N}(-0.4, 0.2^2)$ and half follows $\beta \sim \mathcal{N}(0.4, 0.2^2)$.
+Correlation matrix of $\beta$ is the same as correlation matrix of genes.
+- Type (I) : half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$.
+Correlation matrix of $\beta$ is the same as correlation matrix of genes.
 
@@
 
-```{r fig.height=10, fig.cap="Statistical power depending on simulation setting. N is the number of patients and NG the number of genes in the pathway."}
+```{r fig.height=10, fig.cap="Statistical power depending on simulation setting.
+N is the number of patients and NG the number of genes in the pathway.
+Note : scenario D, E and F do not rely on correlation matrix to simulate variance covariance matrix so the those matrix are unlikely to be positive definite."}
 source("script/script_simulation/figures_simu_rembrandt_breast_v1.R")
@@
 
-Overall, all methods have close statistical power but sGBJ and Global boost test tends to be among the worst methods. There are some exceptions to note :
+```{r fig.height=14, fig.cap="Statistical power depending on simulation setting (second batch). N is the number of patients and NG the number of genes in the pathway."}
+source("script/script_simulation/figures_simu_rembrandt_breast_v2.R")
 
-- Global Boost Test performs well in type F
-- sGBJ performs well when N = 100 and NG = 50 which might be related to its asymptotic nature.
-- In other cases, Wald and Global Test slightly outperformed other methods and Wald seems better than Global Test.
+plot_simu
+```
+
+```{r fig.height=6, fig.cap="Time depending on simulation setting (second batch). N is the number of patients and NG the number of genes in the pathway."}
+plot_time
+```
+
+# Discussion
+
+Overall :
+
+- Global Test is fast compared to other approach. It is not completely clear to me why because it also relies on permutation and is similar to Wald test. There is probably a clever implementation ?
+- In terms of power Wald test is the best option, closely followed by Global test
+- sGBJ is slow when the number of genes rise. This is due to permutation for variance matrix estimation that might be reduced to improve computation time.
+- sGBJ performance is similar to most approach except when the number of genes increases.
 
@@
 - sum the wald test statistics
-- H0 : variance of beta is equal to 0.
 
@@
 sGBJ is similar to Wald and Global test in the sense that it is based on one cox model by feature.
-However, it takes into account the covariance of genes in its formula.
-And the data are simulated with no covariance of beta => Problem ?
-look at GBJ paper
+However, it takes into account the covariance of genes in its formula so it should perform better when beta are correlated.

### 2023-09-22 script/script_simulation/figures_simu_rembrandt_breast_v2.R

@@ -17,7 +17,8 @@ dfres <- lapply(ls_folder_path,
          prop_sig_gene = as.factor(prop_sig_gene)) %>%
   mutate(p_val_sig = p_value < 0.05) %>%
   group_by(hp_row, method, prop_sig_gene, case_type, case, type, censoring, n_p, nb_observations, nb_genes) %>%
-  summarise(power = sum(p_val_sig)/n(), .groups = "drop") %>%
+  summarise(power = sum(p_val_sig)/n(), .groups = "drop",
+            mean_time = mean(time)) %>%
   mutate(censoring = factor(censoring,
                             levels = c(0, 0.3),
                             labels = c("0%", "30%")),
@@ -45,15 +46,51 @@ plot_simu <- dfres %>%
            position = "dodge") +
   scale_fill_viridis_d() +
   scale_color_manual(values = c("black", "red")) +
-  facet_grid(case_type ~ n_p,
-             labeller = labeller(case = vec_labels_case,
-                                 type = vec_labels_type)) +
+  facet_grid(case_type ~ n_p) +
   labs(x = "Proportion of significant genes",
        y = "Statistical Power",
        fill = "Method",
        color = "Censoring proportion",
        lty = "Censoring proportion") +
   lims(y = c(0, 1)) +
+  guides(fill = guide_legend(nrow = 4),
+         lty = guide_legend(nrow = 2),
+         color = guide_legend(nrow = 2)) +
+  theme_bw() +
+  theme(legend.position = "bottom")
+
+# Time
+dfrestime <- readRDS(file = "data/result_11020779_time.rds") %>%
+  mutate(case_type = paste0("Case : ", case, " ; Type : ", type),
+         n_p = paste0("N = ", nb_observations, " ; NG = ", nb_genes),
+         prop_sig_gene = as.factor(prop_sig_gene)) %>%
+  select(hp_row, method, prop_sig_gene, case_type, case, type, censoring, n_p, nb_observations, nb_genes, time) %>%
+  mutate(censoring = factor(censoring,
+                            levels = c(0, 0.3),
+                            labels = c("0%", "30%")),
+         method = factor(method,
+                         levels = c("Global Boost Test",
+                                    "Wald Test",
+                                    "Global Test",
+                                    "sGBJ")))
+
+plot_time <- dfrestime %>%
+  ggplot(mapping = aes(x = prop_sig_gene,
+                       y = time,
+                       linetype = censoring,
+                       color = censoring,
[…]

### 2023-09-22 script/script_simulation/meta_parameters_remb_breast.R

@@ -61,4 +61,17 @@ expand.grid(nb_observations = c(50,100),
             prop_two_periods = FALSE) |>
   saveRDS(file = "data/dfScenarioSimpleCorr.rds")
 
+##### new set exp for alpha risk
+expand.grid(nb_observations = c(50,100),
+            nb_genes = c(10,50),
+            prop_sig_gene = c(0.2),
+            variance = 0.2,
+            case = c(4, 5, 6),
+            type = c("Z"),
+            censoring = c(0.3),
+            nb_permutation = 1000,
+            prop_two_periods = FALSE) |>
+  saveRDS(file = "data/dfScenarioAlpha.rds")
+
+
 

### 2023-09-22 functions/fct_generate_beta.R
alpha risk simulation

@@ -17,7 +17,11 @@ fct_generate_beta <- function(type,
   
   nb_sig_gene = round(prop_sig_gene*nb_genes)
   
-  if(type == "A"){
+  if(type == "Z"){
+    
+    res <- rep(0, nb_genes)
+    
+  } else if(type == "A"){
     
     res <- c(rnorm(n = nb_sig_gene, mean = 0, sd = 0.5),
              rep(0, nb_genes-nb_sig_gene))

### 2023-09-22 script/script_simulation/curta/07_simu_from_data_simple_corr_alpha.R

@@ -1,0 +1,60 @@
+print(paste0("Start at ", Sys.time(), "..."))
+
+slar_taskid <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+# slar_taskid <- 1
+slar_jobid <- as.numeric(Sys.getenv("SLURM_ARRAY_JOB_ID"))
+
+
+##### load packages and functions
+message("load packages and functions")
+
+library(dplyr)
+
+lapply(list.files(path = "functions/", full.names = TRUE),
+       source)
+
+##### load data
+message("load dfScenario")
+
+dfScenario <- readRDS("data/dfScenarioAlpha.rds") %>%
+  tibble::rowid_to_column(var = "hp_row")
+
+##### start computation
+message("start computation")
+
+res <- pbapply::pbapply(X = dfScenario,
+                        MARGIN = 1,
+                        FUN = function(row){
+                          fct_simulation_paper(methods = c("sgbj", "gbt", "gt", "wald"),
+                                               nb_observations = as.numeric(row[["nb_observations"]]),
+                                               nb_genes = as.numeric(row[["nb_genes"]]),
+                                               prop_sig_gene = as.numeric(row[["prop_sig_gene"]]),
+                                               variance =  as.numeric(row[["variance"]]),
+                                               case =  row[["case"]],
+                                               type =  row[["type"]],
+                                               censoring =  as.numeric(row[["censoring"]]),
+                                               nb_permutation =  as.numeric(row[["nb_permutation"]]),
+                                               prop_two_periods =  row[["prop_two_periods"]],
+                                               nperm_sGBJ = as.numeric(row[["nb_permutation"]])) %>%
+                            mutate(hp_row = row[["hp_row"]],
+                                   iter = slar_taskid) %>%
+                            return()
+                        }) %>%
+  bind_rows()
+
+##### save results
+print(paste0("Saving start at ", Sys.time(), "..."))
+
+subDir <- paste0("data/result_simu/result_", slar_jobid)
+if (!file.exists(subDir)){
+  dir.create(subDir)
+}
+
+date_str <- gsub(x = Sys.Date(),"-","")
+
+filename_r <- paste0(subDir, "/result_job",
+                     slar_taskid, "_", date_str, ".rds")
+
+saveRDS(res, file = filename_r)
+
+print(paste0("Saving end at ", Sys.time(), "..."))

### 2023-09-22 reporting/simu_rembrandt_breast.Rmd
update with type-I error

@@
 
+- Type (Z) : $\beta = 0$ evaluates the type-I error
 - Type (D) : $\beta \sim \mathcal{N}(0, 0.4^2)$
@@
 
+```{r fig.height=6, fig.cap="Type-I error (Type = Z). N is the number of patients and NG the number of genes in the pathway."}
+plot_simu_alpha
+```
+
 ```{r fig.height=6, fig.cap="Time depending on simulation setting (second batch). N is the number of patients and NG the number of genes in the pathway."}
@@
 - In terms of power Wald test is the best option, closely followed by Global test
+- Type-I error is well controlled by the different methods. We note that sGBJ is too conservative when the number of genes is high.
 - sGBJ is slow when the number of genes rise. This is due to permutation for variance matrix estimation that might be reduced to improve computation time.
-- sGBJ performance is similar to most approach except when the number of genes increases.
+- sGBJ performance is similar to most approach except when the number of genes increases which is partly explained by the over-conservative control of Type-I error.
 
-## Comprehend tests
+# Notes
 
@@
 
-### Wald test
+## Wald test
 
@@
 
-### Global test
+## Global test
 
@@
 
-### Global boost test
+## Global boost test
 
@@
 
-### Thinking
+## Thinking
 

### 2023-09-22 script/script_simulation/figures_simu_rembrandt_breast_v2.R

@@ -2,7 +2,7 @@
 library(dplyr)
 library(ggplot2)
 ##### data #####
-ls_folder_path <- c("data/result_11020779", "data/result_10187723")
+ls_folder_path <- c("data/result_11020779", "data/result_10187723", "data/result_11021425/")
 
 dfres <- lapply(ls_folder_path,
                 FUN = function(x) list.files(x, recursive = T, full.names = T)) %>%
@@ -17,8 +17,14 @@ dfres <- lapply(ls_folder_path,
          prop_sig_gene = as.factor(prop_sig_gene)) %>%
   mutate(p_val_sig = p_value < 0.05) %>%
   group_by(hp_row, method, prop_sig_gene, case_type, case, type, censoring, n_p, nb_observations, nb_genes) %>%
-  summarise(power = sum(p_val_sig)/n(), .groups = "drop",
-            mean_time = mean(time)) %>%
+  summarise(success = sum(p_val_sig),
+            trials = n(),
+            mean_time = mean(time),
+            power = success/trials,
+            power_lower = binom::binom.confint(methods = "exact", x = success, n = trials)$lower,
+            power_upper = binom::binom.confint(methods = "exact", x = success, n = trials)$upper,
+            .groups = "drop") %>%
+  select(-success, -trials) %>%
   mutate(censoring = factor(censoring,
                             levels = c(0, 0.3),
                             labels = c("0%", "30%")),
@@ -30,12 +36,8 @@ dfres <- lapply(ls_folder_path,
 
 
 ##### plot #####
-vec_labels_type <- paste0("Type ", c("A", "B", "C", "D", "E", "F"))
-names(vec_labels_type) <- c("A", "B", "C", "D", "E", "F")
-vec_labels_case <- paste0("Case (", c("I", "II", "III", "IV", "V"), ")")
-names(vec_labels_case) <- c(1:length(vec_labels_case))
-
 plot_simu <- dfres %>%
+  filter(type != "Z") %>%
   ggplot(mapping = aes(x = prop_sig_gene,
                        y = power,
                        linetype = censoring,
@@ -53,6 +55,36 @@ plot_simu <- dfres %>%
        color = "Censoring proportion",
        lty = "Censoring proportion") +
   lims(y = c(0, 1)) +
+  guides(fill = guide_legend(nrow = 4),
+         lty = guide_legend(nrow = 2),
+         color = guide_legend(nrow = 2)) +
[…]

### 2023-09-22 script/script_simulation/figures_simu_rembrandt_breast_v2.R
update report

@@ -63,7 +63,7 @@ plot_simu <- dfres %>%
 
 plot_simu_alpha <- dfres %>%
   filter(type == "Z") %>%
-  ggplot(mapping = aes(x = prop_sig_gene,
+  ggplot(mapping = aes(x = method,
                        y = power,
                        ymin = power_lower,
                        ymax = power_upper,
@@ -71,16 +71,14 @@ plot_simu_alpha <- dfres %>%
                        color = censoring,
                        group = method,
                        fill = method)) +
-  geom_errorbar(position = position_dodge(width=0.45),
-                width = 0) +
-  geom_point(position = position_dodge(width=0.45),
-             shape = 22,
+  geom_errorbar(width = 0) +
+  geom_point(shape = 22,
              size = 3) +
   geom_hline(yintercept = 0.05, lty = 2) +
   scale_fill_viridis_d() +
   scale_color_manual(values = c("black", "red")) +
   facet_grid(case_type ~ n_p) +
-  labs(x = "Proportion of significant genes",
+  labs(x = "",
        y = "Type-I error",
        fill = "Method",
        color = "Censoring proportion",
@@ -89,7 +87,8 @@ plot_simu_alpha <- dfres %>%
          lty = guide_legend(nrow = 2),
          color = guide_legend(nrow = 2)) +
   theme_bw() +
-  theme(legend.position = "bottom")
+  theme(legend.position = "right",
+        axis.text.x = element_text(angle = 45, hjust=1))
 
 # Time
 dfrestime <- readRDS(file = "data/result_11020779_time.rds") %>%

### 2024-06-05 script/rembrandt/rembrant_description.Rmd
recode rembrandt description

---
title: "Rembrant"
author: "TF"
output: 
  bookdown::html_document2:
    toc: yes
    toc_float: true
    number_sections: FALSE
    fig_caption: TRUE
    table_caption: TRUE
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r setup, include=FALSE, eval=TRUE,message=FALSE,warning=FALSE,cache=FALSE}

library(dplyr)
library(ggplot2)
library(survival)

datas_tot=readRDS(here::here("data/rembrandt/dfGenesAndClinical.rds"))

```

```{r dm}
##### Datamanagement

datas_filtered <- datas_tot |> 
  filter(DISEASE_TYPE %in% c("ASTROCYTOMA", "OLIGODENDROGLIOMA"),
         !is.na(OVERALL_SURVIVAL_MONTHS),
         !is.na(EVENT_OS),
         GENDER != "",
         AGE_RANGE != "") |> 
  select(-c(SUBJECT_ID,
            COPY_NUMBER,
            GENE_EXP,
            PARENT_ID,
            TYPE,
            TIMEPOINT,
            WHO_GRADE,
            RACE,
            INSTITUTION_NAME,
            KARNOFSKY,
            NEURO_EXAM_SCORE,
            DISEASE_EVAL_MRI,
            STEROID_DOSE_STATUS,
            ANTI_CONVULSANT_STATUS,
            starts_with("PRIOR_THERAPY_"),
            starts_with("ON_THERAPY"),
            starts_with("PT_NAME"))) |> 
  mutate(GENDER = as.factor(GENDER),
         AGE_RANGE = as.factor(AGE_RANGE),
         AGE_RANGE = forcats::fct_collapse(AGE_RANGE,
                                           "< 40" = c("15-19","20-24",
                                                      "25-29", "30-34",
                                                      "35-39"),
                                           ">= 40" = c("40-44","45-49",
                                                       "50-54","55-59",
                                                       "60-64","65-69",
                                                       "70-74","75-79",
                                                       "80-84","85-89"))) |> 
  sjlabelled::var_labels(AGE_RANGE = "Age",
                         GENDER = "Sex",
                         EVENT_OS = "Recurrence",
                         OVERALL_SURVIVAL_MONTHS = "Follow-up period in months",
                         DISEASE_TYPE = "Tumor type")
```

```{r desc}
##### Description
datas_filtered |> 
  dplyr::select(AGE_RANGE,
                GENDER,
                EVENT_OS,
                OVERALL_SURVIVAL_MONTHS,
                DISEASE_TYPE) |> 
  gtsummary::tbl_summary(by = DISEASE_TYPE)
```

```{r genecount, fig.cap="Rembrandt count per patient per tumor type"}
datas_filtered |> 
  dplyr::select(-c(AGE_RANGE,
                   GENDER,
                   EVENT_OS,
                   OVERALL_SURVIVAL_MONTHS)) |> 
  tidyr::pivot_longer(cols = -c(BIOSPECIMEN_ID, DISEASE_TYPE)) |> 
  ggplot(mapping = aes(x = BIOSPECIMEN_ID, y = value, fill = DISEASE_TYPE)) +
  geom_boxplot(outlier.shape = NA,lwd=0.1, alpha=1) +
  facet_wrap(.~DISEASE_TYPE, scales = "free_x") +
  theme_minimal() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "none")+
  labs(x = "patients",
       y = "counts")

```

```{r rembrandtpca, fig.cap="First two PCA factorial plans"}
pca_model <- datas_filtered |> 
  dplyr::select(-c(BIOSPECIMEN_ID,
                   DISEASE_TYPE,
                   AGE_RANGE,
                   GENDER,
                   EVENT_OS,
                   OVERALL_SURVIVAL_MONTHS)) |> 
  FactoMineR::PCA(ncp = 4, graph = FALSE)

pca_coord <- pca_model$ind$coord |> 
  as.data.frame() |> 
  mutate(BIOSPECIMEN_ID = datas_filtered$BIOSPECIMEN_ID,
         DISEASE_TYPE = datas_filtered$DISEASE_TYPE,
         .before = 1)

p1 <- plot(pca_model, choix = "ind", axes = c(1, 2))
p2 <- plot(pca_model, choix = "ind", axes = c(3, 4))

ggpubr::ggarrange(p1, p2, common.legend = TRUE)
```

Here are the kaplan meier curves for the two types of disease:

```{r kmrembrandt, fig.cap="Kaplan meier curves for Astro, Oligo"}
fit <- survfit(Surv(time=OVERALL_SURVIVAL_MONTHS, event=EVENT_OS) ~ DISEASE_TYPE,
               data = datas_filtered)

p_type=survminer::ggsurvplot(fit,
                             data = datas_filtered,
                             surv.median.line = "hv", # Add medians survival

                             palette = c("#8338EC", "#FFB703"),

                             # Change legends: title & labels
                             legend.title = "Type",
                             legend.labs = c("Astrocytoma", "Oligodendroglioma"),

                             conf.int = TRUE,
                             # Add risk table
                             risk.table = TRUE,
                             tables.height = 0.2,
                             tables.theme = survminer::theme_cleantable(),

                             title="Type of disease",
                             ggtheme = theme_bw() # Change ggplot2 theme
)
p_type

```

### 2024-06-05 script/rembrandt/rembrant_description.Rmd
clean rembrandt folder

@@
 
-```{r setup, include=FALSE, eval=TRUE,message=FALSE,warning=FALSE,cache=FALSE}
+```{r setup}
 
@@
 
-datas_tot=readRDS(here::here("data/rembrandt/dfGenesAndClinical.rds"))
+df_rembrandt = readRDS(here::here("data/rembrandt/dfGenesAndClinical.rds"))
 
 ```
+
+# Describe Rembrandt cohort
 
@@
 
-datas_filtered <- datas_tot |> 
+df_rembrandt_clean <- df_rembrandt |> 
   filter(DISEASE_TYPE %in% c("ASTROCYTOMA", "OLIGODENDROGLIOMA"),
@@
 ##### Description
-datas_filtered |> 
+df_rembrandt_clean |> 
   dplyr::select(AGE_RANGE,
@@
 ```{r genecount, fig.cap="Rembrandt count per patient per tumor type"}
-datas_filtered |> 
+df_rembrandt_clean |> 
   dplyr::select(-c(AGE_RANGE,
@@
 ```{r rembrandtpca, fig.cap="First two PCA factorial plans"}
-pca_model <- datas_filtered |> 
+pca_model <- df_rembrandt_clean |> 
   dplyr::select(-c(BIOSPECIMEN_ID,
@@
   as.data.frame() |> 
-  mutate(BIOSPECIMEN_ID = datas_filtered$BIOSPECIMEN_ID,
-         DISEASE_TYPE = datas_filtered$DISEASE_TYPE,
+  mutate(BIOSPECIMEN_ID = df_rembrandt_clean$BIOSPECIMEN_ID,
+         DISEASE_TYPE = df_rembrandt_clean$DISEASE_TYPE,
          .before = 1)
@@
 fit <- survfit(Surv(time=OVERALL_SURVIVAL_MONTHS, event=EVENT_OS) ~ DISEASE_TYPE,
-               data = datas_filtered)
+               data = df_rembrandt_clean)
 
 p_type=survminer::ggsurvplot(fit,
-                             data = datas_filtered,
+                             data = df_rembrandt_clean,
                              surv.median.line = "hv", # Add medians survival

### 2024-06-05 script/rembrandt/thomas_rembrandt_old.R

@@ -1,44 +1,0 @@
-library(dplyr)
-
-##### import pathways
-pathways=GSA::GSA.read.gmt("data/rembrandt/c6.all.v7.1.symbols.gmt")
-liste_pathways=pathways$genesets
-liste_names=pathways$geneset.names
-
-##### import data
-datas_tot=readRDS("data/rembrandt/dfGenesAndClinical.rds") %>%
-  select(-all_of(c("SUBJECT_ID",
-                   "PARENT_ID", "TYPE",
-                   "TIMEPOINT", "AGE_RANGE",
-                   "GENDER",
-                   "WHO_GRADE", "RACE",
-                   "INSTITUTION_NAME", "KARNOFSKY",
-                   "NEURO_EXAM_SCORE", "DISEASE_EVAL_MRI",
-                   "STEROID_DOSE_STATUS", "ANTI_CONVULSANT_STATUS",
-                   "PRIOR_THERAPY_RADIATION_TYPE", "PRIOR_THERAPY_CHEMO_AGENT_NAME",
-                   "PRIOR_THERAPY_SURGERY_TUMOR_HISTOLOGY", "PRIOR_THERAPY_PROCEDURE",
-                   "PRIOR_THERAPY_SURGERY_OUTCOME", "ON_THERAPY_RADIATION_SITE",
-                   "ON_THERAPY_RADIATION_TYPE", "ON_THERAPY_CHEMO_AGENT_NAME",
-                   "ON_THERAPY_SURGERY_HIST_DIAG", "ON_THERAPY_SURGERY_TITLE",
-                   "ON_THERAPY_SURGERY_OUTCOME", "PRIOR_THERAPY_SURGERY_OUTCOME_2",
-                   "PT_NAME2", "PRIOR_THERAPY_SURGERY_OUTCOME_1",
-                   "PT_NAME1", "GENE_EXP",
-                   "COPY_NUMBER"))) %>%
-  filter(DISEASE_TYPE %in% c("ASTROCYTOMA","GBM","OLIGODENDROGLIOMA"),
-         !is.na(EVENT_OS))
-
-lapply(liste_pathways, FUN = function(path){
-  print(path)
-  ncol(datas_tot[,colnames(datas_tot) %in% path])
-})
-
-counts_pathway=datas_tot[,colnames(datas_tot)%in%liste_pathways[[path]]]
-patients=datas_tot[,c(1,3,2,4)]
-patients=as.data.frame(patients)
[…]

### 2024-06-05 script/rembrandt/rembrant_description.Rmd
ready functions for rembrandt curta analysis

@@
 
-```{r setup}
+```{r setup, results='hide'}
 
@@
 
-df_rembrandt = readRDS(here::here("data/rembrandt/dfGenesAndClinical.rds"))
+lapply(list.files(path = here::here("functions"), full.names = TRUE), source)
+
+df_rembrandt <- fct_load_clean_rembrandt()
 
@@
 
-```{r dm}
-##### Datamanagement
-
-df_rembrandt_clean <- df_rembrandt |> 
-  filter(DISEASE_TYPE %in% c("ASTROCYTOMA", "OLIGODENDROGLIOMA"),
-         !is.na(OVERALL_SURVIVAL_MONTHS),
-         !is.na(EVENT_OS),
-         GENDER != "",
-         AGE_RANGE != "") |> 
-  select(-c(SUBJECT_ID,
-            COPY_NUMBER,
-            GENE_EXP,
-            PARENT_ID,
-            TYPE,
-            TIMEPOINT,
-            WHO_GRADE,
-            RACE,
-            INSTITUTION_NAME,
-            KARNOFSKY,
-            NEURO_EXAM_SCORE,
-            DISEASE_EVAL_MRI,
-            STEROID_DOSE_STATUS,
-            ANTI_CONVULSANT_STATUS,
-            starts_with("PRIOR_THERAPY_"),
-            starts_with("ON_THERAPY"),
-            starts_with("PT_NAME"))) |> 
-  mutate(GENDER = as.factor(GENDER),
-         AGE_RANGE = as.factor(AGE_RANGE),
-         AGE_RANGE = forcats::fct_collapse(AGE_RANGE,
-                                           "< 40" = c("15-19","20-24",
-                                                      "25-29", "30-34",
-                                                      "35-39"),
-                                           ">= 40" = c("40-44","45-49",
-                                                       "50-54","55-59",
-                                                       "60-64","65-69",
-                                                       "70-74","75-79",
-                                                       "80-84","85-89"))) |> 
-  sjlabelled::var_labels(AGE_RANGE = "Age",
-                         GENDER = "Sex",
-                         EVENT_OS = "Recurrence",
-                         OVERALL_SURVIVAL_MONTHS = "Follow-up period in months",
-                         DISEASE_TYPE = "Tumor type")
-```
-
 ```{r desc}
 ##### Description
-df_rembrandt_clean |> 
+df_rembrandt |> 
   dplyr::select(AGE_RANGE,
@@
 ```{r genecount, fig.cap="Rembrandt count per patient per tumor type"}
-df_rembrandt_clean |> 
+df_rembrandt |> 
   dplyr::select(-c(AGE_RANGE,
@@
 ```{r rembrandtpca, fig.cap="First two PCA factorial plans"}
-pca_model <- df_rembrandt_clean |> 
+pca_model <- df_rembrandt |> 
   dplyr::select(-c(BIOSPECIMEN_ID,
@@
   as.data.frame() |> 
-  mutate(BIOSPECIMEN_ID = df_rembrandt_clean$BIOSPECIMEN_ID,
-         DISEASE_TYPE = df_rembrandt_clean$DISEASE_TYPE,
+  mutate(BIOSPECIMEN_ID = df_rembrandt$BIOSPECIMEN_ID,
+         DISEASE_TYPE = df_rembrandt$DISEASE_TYPE,
          .before = 1)
@@
 fit <- survfit(Surv(time=OVERALL_SURVIVAL_MONTHS, event=EVENT_OS) ~ DISEASE_TYPE,
-               data = df_rembrandt_clean)
+               data = df_rembrandt)
 
 p_type=survminer::ggsurvplot(fit,
-                             data = df_rembrandt_clean,
+                             data = df_rembrandt,
                              surv.median.line = "hv", # Add medians survival

### 2024-06-05 functions/fct_breast_cancer_analysis.R

@@ -1,0 +1,39 @@
+# function to analyse breast cancer dataset
+fct_breast_cancer_analysis <- function(df_data,
+                                       ls_pathways,
+                                       vec_covariates,
+                                       nb_permutation = 1000){
+  
+  vec_grade <- c("all", unique(df_data$grade))
+  
+  res <- tidyr::expand_grid(grade = vec_grade,
+                            pathway = names(ls_pathways)) |> 
+    apply(MARGIN = 1,
+          FUN = function(row_i){
+            grade_i <- row_i[["grade"]]
+            pathway_i <- row_i[["pathway"]]
+            
+            message(paste0("------- pathway : ", pathway_i, ", grade : ", grade_i))
+            
+            vec_genes_of_pathway <- ls_pathways[[pathway_i]]
+            
+            if(grade_i != "all"){
+              df_data_analysis <- df_data |> 
+                filter(grade == grade_i)
+            } else {
+              df_data_analysis <- df_data
+            }
+            
+            res <- fct_perform_tests(df_data_analysis = df_data_analysis,
+                                     vec_genes_of_pathway = vec_genes_of_pathway,
+                                     vec_covariates = vec_covariates,
+                                     nb_permutation = nb_permutation) |> 
+              mutate(pathway = pathway_i,
+                     grade = grade_i,
+                     .before = 1)
+          }) |> 
+    bind_rows()
+  
+  return(res)
+  
+}

### 2024-06-05 functions/fct_load_clean_rembrandt.R

@@ -1,0 +1,45 @@
+fct_load_clean_rembrandt <- function(path = here::here("data/rembrandt/dfGenesAndClinical.rds")){
+  df_rembrandt = readRDS(path)
+  
+  df_rembrandt_clean <- df_rembrandt |> 
+    filter(DISEASE_TYPE %in% c("ASTROCYTOMA", "OLIGODENDROGLIOMA"),
+           !is.na(OVERALL_SURVIVAL_MONTHS),
+           !is.na(EVENT_OS),
+           GENDER != "",
+           AGE_RANGE != "") |> 
+    select(-c(SUBJECT_ID,
+              COPY_NUMBER,
+              GENE_EXP,
+              PARENT_ID,
+              TYPE,
+              TIMEPOINT,
+              WHO_GRADE,
+              RACE,
+              INSTITUTION_NAME,
+              KARNOFSKY,
+              NEURO_EXAM_SCORE,
+              DISEASE_EVAL_MRI,
+              STEROID_DOSE_STATUS,
+              ANTI_CONVULSANT_STATUS,
+              starts_with("PRIOR_THERAPY_"),
+              starts_with("ON_THERAPY"),
+              starts_with("PT_NAME"))) |> 
+    mutate(GENDER = as.factor(GENDER),
+           AGE_RANGE = as.factor(AGE_RANGE),
+           AGE_RANGE = forcats::fct_collapse(AGE_RANGE,
+                                             "< 40" = c("15-19","20-24",
+                                                        "25-29", "30-34",
+                                                        "35-39"),
+                                             ">= 40" = c("40-44","45-49",
+                                                         "50-54","55-59",
+                                                         "60-64","65-69",
+                                                         "70-74","75-79",
+                                                         "80-84","85-89"))) |> 
+    sjlabelled::var_labels(AGE_RANGE = "Age",
+                           GENDER = "Sex",
+                           EVENT_OS = "Death",
+                           OVERALL_SURVIVAL_MONTHS = "Follow-up period in months",
+                           DISEASE_TYPE = "Tumor type")
+  
+  return(df_rembrandt_clean)
+}

### 2024-06-05 functions/fct_perform_tests.R

@@ -1,0 +1,81 @@
+# function to perform all 4 tests on a dataset depending on genes and covariates
+fct_perform_tests <- function(df_data_analysis,
+                              vec_genes_of_pathway,
+                              vec_covariates,
+                              nb_permutation = 1000){
+  
+  surv_obj <- survival::Surv(time = df_data_analysis$time,
+                             event = df_data_analysis$event)
+  
+  factor_matrix <- df_data_analysis %>% select(any_of(vec_genes_of_pathway))
+  
+  covariates <- df_data_analysis %>% select(any_of(vec_covariates))
+  
+  all_expl_features <- bind_cols(factor_matrix, covariates)
+  
+  alternative_hyp <- paste0("~ ",
+                            paste0("`", colnames(factor_matrix), "`", collapse = " + "))
+  null_hyp <- paste0("~ ",
+                     paste0(colnames(covariates), collapse = " + "))
+  
+  ##### TESTS
+  ### sGBJ
+  message("--- sGBJ")
+  time_start <- Sys.time()
+  sGBJ <- sGBJ::sGBJ(surv = surv_obj,
+                     factor_matrix = factor_matrix,
+                     covariates = covariates,
+                     nperm = nb_permutation)$sGBJ_pvalue
+  time_end <- Sys.time()
+  time_sGBJ <- difftime(time_end, time_start, units = "secs")
+  ### GBT
+  message("--- GBT")
+  time_start <- Sys.time()
+  GBT <- globalboosttest::globalboosttest(
+    X = factor_matrix,
+    Y = surv_obj,
+    Z = covariates,
+    nperm = nb_permutation,
+    mstop = 500,
+    pvalueonly = TRUE
+  )$pvalue |>
+    as.numeric()
+  time_end <- Sys.time()
+  time_GBT <- difftime(time_end, time_start, units = "secs")
+  ### WALD
+  message("--- Wald")
+  time_start <- Sys.time()
+  wald_pval <- Wald_Test_Perm(vectime = df_data_analysis$time,
+                              vecevent = df_data_analysis$event,
+                              x = as.matrix(factor_matrix),
+                              covariates = as.matrix(covariates),
+                              nb_permutation = nb_permutation)
+  time_end <- Sys.time()
+  time_wald <- difftime(time_end, time_start, units = "secs")
+  
+  ### GT
+  message("--- Global Test")
+  time_start <- Sys.time()
+  GT<-globaltest::gt(response = surv_obj,
+                     data = all_expl_features,
[…]

### 2024-06-05 functions/fct_simulation_paper.R

@@ -124,6 +124,14 @@ fct_simulation_paper <- function(methods = c("sgbj", "gbt", "gt", "wald", "rf"),
     start_time <- Sys.time()
     set.seed(as.numeric(start_time) + base_seed)
     p_val_rf <- rfpermutation(dfdata = dfdata, nb_permutation = nb_permutation)
+    # fit_ranger <- ranger::ranger(survival::Surv(time,event) ~ .,
+    #                              data = dfdata,
+    #                              num.trees = 1000)
+    # 
+    # p_val_rf <- survcomp::concordance.index(x = rowSums(fit_ranger$chf),
+    #                                         surv.time = dfdata$time,
+    #                                         surv.event = dfdata$event,
+    #                                         method = "noether")$p.value
     end_time <- Sys.time()
     
     ls_pvalue$RF <- p_val_rf

### 2024-06-05 functions/statsTest.R

@@ -42,20 +42,31 @@ statsTest=function(vectime, vecevent, x){
 #' @param vectime the time
 #' @param vecevent the event
 #' @param x the genes experession
+#' @param covariates a dataframe of covariates (default is NULL)
 #'
 #' @return A vector with the statistical test of global test, adewale test
 #' @export
 #'
-pre_Wald_Test=function(vectime, vecevent, x){
+pre_Wald_Test=function(vectime, vecevent, x, covariates = NULL){
   # baseline
   survObj <- survival::Surv(vectime,vecevent)
-  ucox <- lapply(X = 1:ncol(x),
-         FUN = function(i){
-           ocx<-survival::coxph(survObj~x[,i])
-           res<-ocx$coefficients/sqrt(ocx$var)
-           return(res)
-         }) %>%
-    unlist()
+  if(is.null(covariates)){
+    ucox <- lapply(X = 1:ncol(x),
+                   FUN = function(i){
+                     ocx<-survival::coxph(survObj~x[,i])
+                     res<-ocx$coefficients/sqrt(ocx$var)
+                     return(res)
+                   }) %>%
+      unlist()
+  } else {
+    ucox <- lapply(X = 1:ncol(x),
+                   FUN = function(i){
+                     ocx<-survival::coxph(survObj~x[,i]+covariates)
+                     res<-ocx$coefficients[1]/sqrt(ocx$var[1])
+                     return(res)
+                   }) %>%
+      unlist()
+  }
   
   res <- sum(ucox^2)
   
@@ -74,7 +85,7 @@ pre_Wald_Test=function(vectime, vecevent, x){
 #' @return A vector with the statistical test of global test, adewale test
 #' @export
 #'
-Wald_Test_Perm=function(vectime, vecevent, x, nb_permutation){
+Wald_Test_Perm=function(vectime, vecevent, x, covariates, nb_permutation){
   # baseline
   baseline_ucox <- pre_Wald_Test(vectime = vectime,
                                  vecevent = vecevent,

### 2024-06-05 script/rembrandt/rembrandt_pathway_analysis.R

@@ -1,0 +1,31 @@
+taskid <- Sys.getenv('SLURM_ARRAY_TASK_ID')
+print(taskid)
+######## PATWAY ANALYSIS OF REMBRANDT FILE #########
+##### load packages #####
+library(dplyr)
+library(ggplot2)
+library(survival)
+library(survminer)
+library(parallel)
+
+lapply(list.files(path = here::here("functions"), full.names = TRUE), source)
+
+##### load data #####
+df_rembrandt <- fct_load_clean_rembrandt()
+
+##### load pathways #####
+pathways = GSA::GSA.read.gmt("data/rembrandt/c6.all.v7.1.symbols.gmt")
+liste_pathways=pathways$genesets
+liste_names=pathways$geneset.names
+names(liste_pathways) <- liste_names
+
+##### analysis
+df_analysis_rembrandt <- fct_breast_cancer_analysis(df_data = df_rembrandt |> 
+                                                      rename(grade = DISEASE_TYPE,
+                                                             time = OVERALL_SURVIVAL_MONTHS,
+                                                             event = EVENT_OS),
+                                                    ls_pathways = liste_pathways[taskid],
+                                                    vec_covariates = c("AGE_RANGE", "GENDER"),
+                                                    nb_permutation = 1000)
+
+saveRDS(df_analysis_rembrandt, file = here::here("results/rembrandt/df_analysis_rembrandt.rds"))

### 2024-06-06 script/rembrandt/rembrant_description.Rmd
update rembrandt analysis

@@
 ```
+
+# Rembrandt pathway analysis
+
+```{r importrembrandtpathway}
+df_rembrandt_pathway <- lapply(list.files(here::here("results/rembrandt/"), full.names = T),
+                               readRDS) |> 
+  bind_rows() |> 
+  select(-starts_with("time_"), -nb_genes) |> 
+  tidyr::pivot_longer(cols = c("sGBJ", "GBT", "Wald", "GT"),
+                      names_to = "method",
+                      values_to = "p_value") |> 
+  group_by(grade, method) |> 
+  mutate(p_value_BH = p.adjust(p_value)) |> 
+  ungroup() |> 
+  mutate(significant = as.integer(p_value_BH < 0.05))
+```
+
+```{r upsetplot, fig.cap="Upset plot of the different methods", fig.height=8}
+ls_plots <- df_rembrandt_pathway |> 
+  filter(significant == 1) |>
+  group_by(pathway, grade) |> 
+  summarise(methods = list(unique(method))) |> 
+  group_by(grade) |> 
+  group_split() |> 
+  lapply(FUN = function(df_i){
+    title_i <- unique(df_i$grade)
+    plot_i <- df_i |> 
+      ggplot(aes(x=methods)) +
+      geom_bar() +
+      ggupset::scale_x_upset(n_intersections = 6, order_by = "freq") +
+      theme_minimal() +
+      ggtitle(title_i)
+
+    return(plot_i)
+  })
+ggpubr::ggarrange(plotlist = ls_plots, ncol = 1)
+```
+
+```{r}
+nb_pathway = df_rembrandt_pathway |> 
+  pull(pathway) |> 
+  unique() |> 
+  length()
+
+df_sGBJ_rank <- df_rembrandt_pathway |> 
+  group_by(grade) |> 
+  filter(method == "sGBJ") |> 
+  mutate(sGBJ_rank = rank(p_value)) |> 
+  ungroup() |> 
+  select(pathway, grade, sGBJ_rank)
+
+df_BH_limit <- data.frame(sGBJ_rank = seq(1:nb_pathway)) |> 
+  mutate(p_value = 0.05/nb_pathway * sGBJ_rank)
+
+df_rembrandt_pathway |> 
+  left_join(df_sGBJ_rank, by = c("pathway", "grade")) |> 
+  ggplot(mapping = aes(x = sGBJ_rank, y = p_value)) +
+  geom_hline(mapping = aes(color = "5% threshold", yintercept = 0.05)) +
+  geom_line(data = df_BH_limit, mapping = aes(color = "BH limit")) +
+  ggnewscale::new_scale_color() +
+  geom_point(mapping = aes(color = method)) +
+  facet_grid(grade ~ ., scales = "free_y") +
+  scale_y_log10() +
+  annotation_logticks(sides = "l") +
+  theme_minimal()
+```

### 2024-07-01 script/rembrandt/rembrandt_pathway_analysis.R
change save rds files

@@ -1,12 +1,9 @@
-taskid <- Sys.getenv('SLURM_ARRAY_TASK_ID')
+taskid <- as.numeric(Sys.getenv('SLURM_ARRAY_TASK_ID'))
 print(taskid)
 ######## PATWAY ANALYSIS OF REMBRANDT FILE #########
 ##### load packages #####
 library(dplyr)
-library(ggplot2)
 library(survival)
-library(survminer)
-library(parallel)
 
 lapply(list.files(path = here::here("functions"), full.names = TRUE), source)
 
@@ -28,4 +25,4 @@ df_analysis_rembrandt <- fct_breast_cancer_analysis(df_data = df_rembrandt |>
                                                     vec_covariates = c("AGE_RANGE", "GENDER"),
                                                     nb_permutation = 1000)
 
-saveRDS(df_analysis_rembrandt, file = here::here("results/rembrandt/df_analysis_rembrandt.rds"))
+saveRDS(df_analysis_rembrandt, file = here::here(paste0("results/rembrandt/df_analysis_rembrandt_pw_", taskid, ".rds")))

### 2024-07-01 script/rembrandt/rembrant_description.Rmd
ready breast cancer scripts

--- 2024-06-05 script/rembrandt/rembrant_description.Rmd
@@
 
+pathways = GSA::GSA.read.gmt(here::here("data/rembrandt/c6.all.v7.1.symbols.gmt"))
+liste_pathways=pathways$genesets
+liste_names=pathways$geneset.names
+names(liste_pathways) <- liste_names
 ```
@@
 ```
+
+# Rembrandt pathway analysis
+
+## Description of pathways
+
+```{r ecdf, fig.cap="Empirical Cumulative Distribution Function of number of genes by pathway. Rembrandt study."}
+lapply(liste_pathways,
+       function(pathway_i) sum(pathway_i %in% colnames(df_rembrandt))) |> 
+  bind_rows() |> 
+  tidyr::pivot_longer(cols = everything(),
+                      names_to = "pathway",
+                      values_to = "nb_genes") |> 
+  ggplot(mapping = aes(x = nb_genes)) +
+  stat_ecdf() +
+  geom_point(y = 0, shape = "|") +
+  theme_minimal() +
+  labs(x = "Number of genes",
+       y = "Empirical Cumulative Distribution")
+
+```
+
+## Methods comparison
+
+```{r importrembrandtpathway}
+df_rembrandt_pathway <- lapply(list.files(here::here("results/rembrandt/"), full.names = T),
+                               readRDS) |> 
+  bind_rows() |> 
+  select(-starts_with("time_"), -nb_genes) |> 
+  tidyr::pivot_longer(cols = c("sGBJ", "GBT", "Wald", "GT"),
+                      names_to = "method",
+                      values_to = "p_value") |> 
+  group_by(grade, method) |> 
+  mutate(p_value_BH = p.adjust(p_value, method = "BH"),
+         method = factor(method,
+                         levels = c("sGBJ", "GBT", "GT", "Wald"),
+                         labels = c("sGBJ", "Global Boost Test", "Global Test", "Wald")),
+         grade = factor(grade,
+                         levels = c("all", "OLIGODENDROGLIOMA", "ASTROCYTOMA"),
+                         labels = c("All", "Oligodendroglioma", "Astrocytoma"))) |> 
+  ungroup() |> 
+  mutate(significant = as.integer(p_value < 0.05),
+         significant_bh = as.integer(p_value_BH < 0.05)) |> 
+  mutate(method = forcats::fct_relevel(method,
+                                       "sGBJ",
+                                       "Global Test",
+                                       "Wald",
+                                       "Global Boost Test"))
+
+# a = df_rembrandt_pathway |> filter(grade == "Astrocytoma", method == "Wald") |> pull(p_value) |> p.adjust(method = "BH")
+# b = df_rembrandt_pathway |> filter(grade == "Astrocytoma", method == "Wald") |> pull(p_value_BH)
+# cor(a, b)
+```
+
+```{r nbsign, fig.height=5, fig.cap="Number of significant pathways by method."}
+vec_pval_method <- c("Raw", "Benjamini-Hockberg")
+names(vec_pval_method) <- c("nb_sign", "nb_sign_bh")
+
+df_rembrandt_pathway |> 
+  group_by(grade, method) |> 
+  summarise(nb_sign_bh = sum(significant_bh),
+            nb_sign = sum(significant)) |> 
+  tidyr::pivot_longer(cols = starts_with("nb_sign")) |> 
+  ggplot(mapping = aes(x = method, fill = method, y = value)) +
+  geom_bar(stat = "identity") +
+  facet_grid(grade ~ name, scales = "free_y", labeller = labeller(name = vec_pval_method)) +
+  scale_fill_viridis_d(direction = -1) +
+  theme_minimal() +
+  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
+  labs(x = "Method",
+       y = "Nb of significant pathways",
+       fill = "")
+```
+
+```{r upsetplot, fig.cap="Upset plot of the Benjamini-Hockberg p-value agreement according to the different methods", fig.height=8}
+ls_plots <- df_rembrandt_pathway |> 
+  filter(significant_bh == 1) |>
+  group_by(pathway, grade) |> 
+  summarise(methods = list(unique(as.character(method)))) |> 
+  group_by(grade) |> 
+  group_split() |> 
+  lapply(FUN = function(df_i){
+    title_i <- unique(df_i$grade)
+    plot_i <- df_i |> 
+      ggplot(aes(x=methods)) +
+      geom_bar() +
+      ggupset::scale_x_upset(n_intersections = 6, order_by = "freq") +
+      theme_minimal() +
+      theme(plot.margin = margin(10, 10, 10, 50)) +
+      labs(title = title_i,
+           x = "Methods",
+           y = "Number of significant pathways")
+
+    return(plot_i)
+  })
+ggpubr::ggarrange(plotlist = ls_plots, ncol = 1)
+```
+
+```{r figpvaluesmethodsrembrandt, fig.cap="Raw p-values in function of the ordered ranks of sGBJ for the 4 methods (sGBJ , global boost test, Wald test and global test), with the 5% threshold and the Benjamini Hochberg limit, computed for astrocytoma, oligodendroglioma and all patients.
+Nota Bene: The Benjamini Hochberg limit only applies for the sGBJ method, as the ranks are computed for sGBJ only.", fig.height=8}
+nb_pathway = df_rembrandt_pathway |> 
+  pull(pathway) |> 
+  unique() |> 
+  length()
+
+df_sGBJ_rank <- df_rembrandt_pathway |> 
+  group_by(grade) |> 
+  filter(method == "sGBJ") |> 
+  mutate(sGBJ_rank = rank(p_value)) |> 
+  ungroup() |> 
+  select(pathway, grade, sGBJ_rank)
+
+df_BH_limit <- data.frame(sGBJ_rank = seq(1:nb_pathway)) |> 
+  mutate(p_value = 0.05/nb_pathway * sGBJ_rank)
+
+df_rembrandt_pathway |> 
+  left_join(df_sGBJ_rank, by = c("pathway", "grade")) |> 
+  ggplot(mapping = aes(x = sGBJ_rank, y = p_value)) +
+  geom_point(mapping = aes(color = method)) +
+  scale_color_viridis_d(direction = -1) +
+  ggnewscale::new_scale_color() +
+  geom_hline(mapping = aes(color = "5% threshold",
+                           linetype = "5% threshold", yintercept = 0.05),
+                         linewidth = 1) +
+  geom_line(data = df_BH_limit,
+            linewidth = 1,
+            mapping = aes(color = "BH limit", linetype = "BH limit")) +
+  scale_color_manual(values = c("red", "black")) +
+  scale_linetype_manual(values = c(2,1)) +
+  labs(color = "", linetype = "") +
+
+  facet_grid(grade ~ ., scales = "free_y") +
+  scale_y_log10() +
+  annotation_logticks(sides = "l") +
+  theme_minimal() +
+  labs(x = "SGBJ rank", y = "Raw p-value")
+```
+
+NB about GT test : https://www.bioconductor.org/packages/release/bioc/vignettes/globaltest/inst/doc/GlobalTest.pdf : "Because permutations require an exchangeable null hypothesis, such a permutation p-value is only available for the linear model and for the exchangeable null hypotheses ~1 and ~0 in other models."

### 2024-07-01 script/breast_cancer/04_all_pathway_analysis.R

@@ -1,0 +1,88 @@
+##### load packages #####
+library(dplyr)
+library(ggplot2)
+library(survival)
+library(survminer)
+library(parallel)
+
+##### survival analysis #####
+lsdata <- readRDS(file = "data/breast_cancer/datamanaged.rds")
+ls_pathways <- lsdata$pathways
+vec_grade <- unique(lsdata$data$grade)
+nperm <- 1000
+
+##### stratified #####
+# vec_pathways <- ls_pathways[[1]]
+# grade_i <- vec_grade[1]
+# nperm <- 10
+
+df_res_stratified <- mclapply(X = ls_pathways,
+                              mc.cores = 19,
+                              FUN = function(vec_pathways){
+                                res <- lapply(X = vec_grade,
+                                              FUN = function(grade_i){
+                                                ##### PREP DATA
+                                                dfdata <- lsdata$data %>% filter(grade == grade_i)
+                                                
+                                                surv_obj <- survival::Surv(time = dfdata$time,
+                                                                           event = dfdata$event)
+                                                
+                                                factor_matrix <- dfdata %>%
+                                                  select(any_of(vec_pathways))
+                                                
+                                                covariates <- dfdata %>% select(age)
+                                                
+                                                all_expl_features <- bind_cols(factor_matrix, covariates)
+                                                
+                                                alternative_hyp <- paste0("~ ",
+                                                                          paste0("`", colnames(factor_matrix), "`", collapse = " + "))
+                                                null_hyp <- paste0("~ ",
+                                                                   paste0(colnames(covariates), collapse = " + "))
+                                                
+                                                ##### TESTS
+                                                ### sGBJ
+                                                sGBJ <- sGBJ::sGBJ(surv = surv_obj,
+                                                                   factor_matrix = factor_matrix,
+                                                                   covariates = covariates,
+                                                                   nperm = nperm)$sGBJ_pvalue
+                                                ### GBT
+                                                GBT <- globalboosttest::globalboosttest(
+                                                  X = factor_matrix,
+                                                  Y = surv_obj,
+                                                  Z = covariates,
+                                                  nperm = nperm,
+                                                  mstop = 500,
+                                                  pvalueonly = TRUE
+                                                )$pvalue |> 
+                                                  as.numeric()
+                                                ### WALD
+                                                wald_pval <- Wald_Test_Perm(vectime = dfdata$time,
+                                                                            vecevent = dfdata$event,
+                                                                            x = dfdata %>% select(-time,-event) %>% as.matrix(),
[…]

### 2024-07-01 script/breast_cancer/05_master_analysis_breast.R

@@ -1,0 +1,25 @@
+taskid <- Sys.getenv('SLURM_ARRAY_TASK_ID')
+print(taskid)
+######## PATWAY ANALYSIS OF BREAST CANCER FILE #########
+##### load packages #####
+library(dplyr)
+library(ggplot2)
+library(survival)
+library(survminer)
+library(parallel)
+
+##### survival analysis #####
+lsdata <- readRDS(file = "data/breast_cancer/datamanaged.rds")
+
+lapply(X = list.files("functions/", full.names = T), source)
+
+ls_pathways <- lsdata$pathways
+df_data <- lsdata$data
+
+##### analysis
+df_analysis_breast <- fct_breast_cancer_analysis(df_data = df_data,
+                                                 ls_pathways = ls_pathways[taskid],
+                                                 vec_covariates = "age",
+                                                 nb_permutation = 1000)
+
+saveRDS(df_analysis_breast, file = here::here(paste0("results/breast_cancer/df_analysis_breast_pw_", taskid, ".rds")))

### 2024-07-01 script/breast_cancer/breast_description.Rmd
add breast cancer report

---
title: "Breast cancer"
author: "TF"
output: 
  bookdown::html_document2:
    toc: yes
    toc_float: true
    number_sections: FALSE
    fig_caption: TRUE
    table_caption: TRUE
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r setup, results='hide'}

library(dplyr)
library(ggplot2)
library(survival)

lapply(list.files(path = here::here("functions"), full.names = TRUE), source)

lsdata <- readRDS(file = here::here("data/breast_cancer/datamanaged.rds"))

lapply(X = list.files(here::here("functions/"), full.names = T), source)

ls_pathways <- lsdata$pathways
df_data <- lsdata$data

```

# Describe Breast Cancer cohort

```{r desc}
##### Description
df_data |> 
  dplyr::select(age,
                event,
                time,
                grade) |> 
  gtsummary::tbl_summary()
```

```{r genecount, fig.cap="Breast cancer count per patient per tumor type"}
df_data |> 
  dplyr::select(-c(age,
                   event,
                   time)) |> 
  tidyr::pivot_longer(cols = -c(ID, grade)) |> 
  ggplot(mapping = aes(x = ID, y = value, fill = as.factor(grade))) +
  geom_boxplot(outlier.shape = NA,lwd=0.1, alpha=1) +
  facet_wrap(.~grade, scales = "free_x") +
  theme_minimal() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "none")+
  labs(x = "Patients",
       y = "Normalized counts")

```

```{r breastcancerpca, fig.cap="First two PCA factorial plans"}
pca_model <- df_data |> 
  dplyr::select(-c(ID,
                   grade,
                   age,
                   event,
                   time)) |> 
  FactoMineR::PCA(ncp = 4, graph = FALSE)

pca_coord <- pca_model$ind$coord |> 
  as.data.frame() |> 
  mutate(ID = df_data$ID,
         grade = df_data$grade,
         .before = 1)

p1 <- plot(pca_model, choix = "ind", axes = c(1, 2))
p2 <- plot(pca_model, choix = "ind", axes = c(3, 4))

ggpubr::ggarrange(p1, p2, common.legend = TRUE)
```

Here are the kaplan meier curves for the two types of disease:

```{r kmbreastcancer, fig.cap="Kaplan meier curves for Astro, Oligo"}
fit <- survfit(Surv(time=time, event=event) ~ grade,
               data = df_data)

p_type=survminer::ggsurvplot(fit,
                             data = df_data,
                             surv.median.line = "hv", # Add medians survival

                             palette = c("#F48C06", "#DC2F02", "#9D0208"),

                             # Change legends: title & labels
                             legend.title = "",
                             # legend.labs = c("Astrocytoma", "Oligodendroglioma"),

                             conf.int = TRUE,
                             # Add risk table
                             risk.table = TRUE,
                             tables.height = 0.2,
                             tables.theme = survminer::theme_cleantable(),

                             title="Breast cancer patients time-to-event",
                             ggtheme = theme_bw() # Change ggplot2 theme
)
p_type

```

# Breast cancer pathway analysis

## Description of pathways

```{r ecdf, fig.cap="Empirical Cumulative Distribution Function of number of genes by pathway. Breast cancer study."}
lapply(ls_pathways,
       function(pathway_i) sum(pathway_i %in% colnames(df_data))) |> 
  bind_rows() |> 
  tidyr::pivot_longer(cols = everything(),
                      names_to = "pathway",
                      values_to = "nb_genes") |> 
  ggplot(mapping = aes(x = nb_genes)) +
  stat_ecdf() +
  geom_point(y = 0, shape = "|") +
  theme_minimal() +
  labs(x = "Number of genes",
       y = "Empirical Cumulative Distribution")

```

## Methods comparison

```{r importbreastcancerpathway}
df_breast_cancer_pathway <- lapply(list.files(here::here("results/breast_cancer/"), full.names = T),
                               readRDS) |> 
  bind_rows() |> 
  select(-starts_with("time_"), -nb_genes) |> 
  tidyr::pivot_longer(cols = c("sGBJ", "GBT", "Wald", "GT"),
                      names_to = "method",
                      values_to = "p_value") |> 
  group_by(grade, method) |> 
  mutate(p_value_BH = p.adjust(p_value, method = "BH"),
         method = factor(method,
                         levels = c("sGBJ", "GBT", "GT", "Wald"),
                         labels = c("sGBJ", "Global Boost Test", "Global Test", "Wald")),
         grade = factor(grade,
                         levels = c("all", "1", "2", "3"),
                         labels = c("All", "Grade 1", "Grade 2", "Grade 3"))) |> 
  ungroup() |> 
  mutate(significant = as.integer(p_value < 0.05),
         significant_bh = as.integer(p_value_BH < 0.05)) |> 
  mutate(method = forcats::fct_relevel(method,
                                       "sGBJ",
                                       "Global Test",
                                       "Wald",
                                       "Global Boost Test"))
```

```{r nbsign, fig.height=5, fig.cap="Number of significant pathways by method."}
vec_pval_method <- c("Raw", "Benjamini-Hockberg")
names(vec_pval_method) <- c("nb_sign", "nb_sign_bh")

df_breast_cancer_pathway |> 
  group_by(grade, method) |> 
  summarise(nb_sign_bh = sum(significant_bh),
            nb_sign = sum(significant)) |> 
  tidyr::pivot_longer(cols = starts_with("nb_sign")) |> 
  ggplot(mapping = aes(x = method, fill = method, y = value)) +
  geom_bar(stat = "identity") +
  facet_grid(grade ~ name, scales = "free_y", labeller = labeller(name = vec_pval_method)) +
  scale_fill_viridis_d(direction = -1) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  labs(x = "Method",
       y = "Nb of significant pathways",
       fill = "")
```

```{r upsetplot, fig.cap="Upset plot of the Benjamini-Hockberg p-value agreement according to the different methods", fig.height=10}
ls_plots <- df_breast_cancer_pathway |> 
  filter(significant_bh == 1) |>
  group_by(pathway, grade) |> 
  summarise(methods = list(unique(as.character(method)))) |> 
  group_by(grade) |> 
  group_split() |> 
  lapply(FUN = function(df_i){
    title_i <- unique(df_i$grade)
    plot_i <- df_i |> 
      ggplot(aes(x=methods)) +
      geom_bar() +
      ggupset::scale_x_upset(n_intersections = 6, order_by = "freq") +
      theme_minimal() +
      theme(plot.margin = margin(10, 10, 10, 50)) +
      labs(title = title_i,
           x = "Methods",
           y = "Number of significant pathways")

    return(plot_i)
  })
ggpubr::ggarrange(plotlist = ls_plots, ncol = 1)
```

```{r figpvaluesmethodsbreastcancer, fig.cap="Raw p-values in function of the ordered ranks of sGBJ for the 4 methods (sGBJ , global boost test, Wald test and global test), with the 5% threshold and the Benjamini Hochberg limit, computed for astrocytoma, oligodendroglioma and all patients. Nota Bene: The Benjamini Hochberg limit only applies for the sGBJ method, as the ranks are computed for sGBJ only.", fig.height=8}
nb_pathway = df_breast_cancer_pathway |> 
  pull(pathway) |> 
  unique() |> 
  length()

df_sGBJ_rank <- df_breast_cancer_pathway |> 
  group_by(grade) |> 
  filter(method == "sGBJ") |> 
  mutate(sGBJ_rank = rank(p_value)) |> 
  ungroup() |> 
  select(pathway, grade, sGBJ_rank)

df_BH_limit <- data.frame(sGBJ_rank = seq(1:nb_pathway)) |> 
  mutate(p_value = 0.05/nb_pathway * sGBJ_rank)

df_breast_cancer_pathway |> 
  left_join(df_sGBJ_rank, by = c("pathway", "grade")) |> 
  ggplot(mapping = aes(x = sGBJ_rank, y = p_value)) +
  geom_point(mapping = aes(color = method)) +
  scale_color_viridis_d(direction = -1) +
  ggnewscale::new_scale_color() +
  geom_hline(mapping = aes(color = "5% threshold",
                           linetype = "5% threshold", yintercept = 0.05),
                         linewidth = 1) +
  geom_line(data = df_BH_limit,
            linewidth = 1,
            mapping = aes(color = "BH limit", linetype = "BH limit")) +
  scale_color_manual(values = c("red", "black")) +
  scale_linetype_manual(values = c(2,1)) +
  labs(color = "", linetype = "") +

  facet_grid(grade ~ ., scales = "free_y") +
  scale_y_log10() +
  annotation_logticks(sides = "l") +
  theme_minimal() +
  labs(x = "SGBJ rank", y = "Raw p-value")
```

NB about GT test : https://www.bioconductor.org/packages/release/bioc/vignettes/globaltest/inst/doc/GlobalTest.pdf : "Because permutations require an exchangeable null hypothesis, such a permutation p-value is only available for the linear model and for the exchangeable null hypotheses ~1 and ~0 in other models."

### 2024-07-01 script/breast_cancer/breast_description.Rmd
add markdown report for github

@@
   bookdown::html_document2:
+    keep_md: true
     toc: yes

### 2024-07-01 script/rembrandt/rembrant_description.Rmd
change to github format

--- 2024-07-01 script/rembrandt/rembrant_description.Rmd
@@
 output: 
-  bookdown::html_document2:
+  bookdown::github_document2:
     toc: yes

### 2024-07-01 script/breast_cancer/breast_description.Rmd
update to github format

--- 2024-07-01 script/breast_cancer/breast_description.Rmd
@@
 output: 
-  bookdown::html_document2:
+  bookdown::github_document2:
+    keep_md: true
     toc: yes

### 2024-08-07 reporting/discordance_previous_results/discordance LV TF.Rmd
reorganise folders

--- 2023-06-07 reporting/discordance LV TF.Rmd
@@
 ```{r eval=TRUE, fig.cap="Original plot", fig.height=6, echo=FALSE}
-knitr::include_graphics("images/comparison_methods6-1.png")
+knitr::include_graphics(here::here("images/comparison_methods6-1.png"))
 ```
@@
 ```{r eval=TRUE, fig.cap="New plot (fails to replicate)", fig.height=6, echo=FALSE}
-knitr::include_graphics("images/plot_original.png")
+knitr::include_graphics(here::here("images/plot_original.png"))
 ```

### 2024-08-07 script/thomas_rembrandt.R
move old scripts

@@ -1,93 +1,0 @@
-library(dplyr)
-library(survival)
-library(parallel)
-
-##### load data
-pathways = GSA::GSA.read.gmt("data/rembrandt/c6.all.v7.1.symbols.gmt")
-datas_tot = readRDS("data/rembrandt/dfGenesAndClinical.rds")
-
-##### data cleaning
-liste_pathways=pathways$genesets
-liste_names=pathways$geneset.names
-names(liste_pathways) <- liste_names
-
-datas_clean <- datas_tot %>%
-  select(-c("BIOSPECIMEN_ID")) %>%
-  filter(DISEASE_TYPE %in% c("ASTROCYTOMA","GBM","OLIGODENDROGLIOMA"))
-
-##### test 1 pathway
-extract_varcovar <- function(pathway_i){
-  library(dplyr)
-  library(survival)
-  
-  pathway_i <- pathway_i[pathway_i %in% colnames(datas_clean)]
-  
-  # return NULL if no gene in the pathway
-  if(length(pathway_i) == 0) return()
-  
-  ##### cox beta and significance by gene
-  dfCox <- lapply(pathway_i,
-         FUN = function(gene_i){
-           form <- paste0("Surv(OVERALL_SURVIVAL_MONTHS, EVENT_OS) ~ `", gene_i, "`")
-           
-           cox_model <- coxph(as.formula(form),
-                   data = datas_clean) %>%
-             summary()
-           
-           cox_model$coefficients %>%
-             janitor::clean_names() %>%
-             as.data.frame() %>%
-             dplyr::select(coef, se_coef, pr_z) %>%
-             dplyr::mutate(GENE = gene_i, .before = 1) %>%
-             return(.)
-         }) %>%
-    bind_rows() %>%
-    mutate(SIGN = pr_z < 0.05)
-  
-  ##### get the variance covariance
-  mat_var_covar <- datas_clean %>%
-    select(all_of(pathway_i)) %>%
-    var()
-  
-  dfVariance <- diag(mat_var_covar) %>%
-    as.data.frame() %>%
-    janitor::clean_names() %>%
-    rename("VARIANCE" = "x") %>%
-    tibble::rownames_to_column(var = "GENE")
-  
-  
-  dfCovar <- mat_var_covar %>%
-    as.data.frame() %>%
-    tibble::rownames_to_column(var = "GENE") %>%
[…]

### 2024-08-07 script/script_simulation/GBJ_simulations.R
reorganise_folder

@@ -1,301 +1,0 @@
-## package install
-# BiocManager::install("globaltest")
-# install.packages("coin")
-# install.packages("mboost")
-# install.packages("globalboosttest", repos = c("http://R-Forge.R-project.org"), dep = TRUE)
-
-############################################################################################
-##### Load packages ##### 
-
-library(mvtnorm)
-library(GBJ)
-library(survival)
-library(globaltest)
-library(globalboosttest)
-
-############################################################################################
-##### Define function to compute global testn adewale test and global boost test ##### 
-
-#' stats2
-#'
-#' @description Function which compute test statistics for Global Boost test, Wald test and Global Test
-#'
-#' @param data The data for which, censoring time is the first column, event is the second column and all other columns are genes
-#' @param ng The number of gene (seems redondant with data argument)
-#' @param gsize The number of observation per gene (seems redondant with data argument)
-#' @param nstat The number of statistical test (seems useless as it is always returning the 3 test statistics)
-#'
-#' @return A vector with the statistical test of global test, adewale test and global boost test
-#' @export
-#'
-stats2=function(data,ng,gsize,nstat){
-  time<-data[,1]
-  csg<-data[,2]
-  x<-data[,3:ncol(data)]
-  wtcx<-matrix(0,nrow=2,ncol=ng)
-  ucox<-rep(0,ng)
-  escx0<-0
-  escx1<-0
-  gt<-0
-  aw<-0
-  bst<-0
-  ts<-rep(0,nstat)
-  for(i in 1:ng){
-    ocx<-coxph(Surv(time,csg)~x[,i])
-    ucox[i]<-ocx$coefficients/sqrt(ocx$var)
-  }
-  
-  #global test
-  ogt<-gt(Surv(time,csg),x[,1:gsize],model="cox")
-  gt<-z.score(ogt)
-  
-  #Adewale test
-  aw<-sum(ucox[1:gsize]^2)
[…]

### 2024-08-07 script/script_simulation/GBJ_simulations_clean.R

@@ -1,56 +1,0 @@
-##### load functions and package #####
-library(dplyr)
-library(doParallel)
-library(parallel)
-
-lapply(list.files(path = "functions/", full.names = TRUE),
-       source)
-
-##### hyperparameters #####
-
-# paper default : 2000 repetition
-# nb_observations = 50 # paper default = 50
-# nb_genes = 50 # paper default = 50
-# prop_sig_gene = 0.2 # pg = c(0.05, 0.1, 0.2)
-# variance = 0.2 # paper default = 0.2
-# case = 2 # case = c(1, 2, 3)
-# type = "A" # type = c("A", "B", "C")
-# censoring = 0.3 # censoring = c(0, 0.3)
-# nb_permutation = 100 # paper default = 1000
-# prop_two_periods = FALSE
-# slam = 0.005
-
-set.seed(2)
-
-myfunction <- function(x){
-  library(dplyr)
-  fct_simulation_paper(nb_observations = 50,
-                       nb_genes = 50,
-                       prop_sig_gene = 0.2,
-                       variance = 0.2,
-                       case = 3,
-                       type = "C",
-                       censoring = 0.3,
-                       nb_permutation = 3,
-                       prop_two_periods = FALSE)
-}
-
-
-no_cores <- detectCores(logical = TRUE)
-cl <- makeCluster(no_cores-1)  
-registerDoParallel(cl)
-clusterExport(cl, ls())
-seq_compute <- 1:100
-
-results<- c(parLapply(cl = cl,
-                      X = seq_compute,
-                      fun= myfunction))
-
-
-results %>%
-  bind_rows(.id = "iter") %>%
-  mutate(signif = p_value < 0.05) %>%
-  group_by(method) %>%
-  summarise(power = sum(signif)/n())
-
-##### save results #####

### 2024-08-07 script/script_simulation/generation_figure.R

@@ -1,72 +1,0 @@
-results=read.table("data/result_simu_original/comparaison_survival_sets_bis_corr_controle2.txt")
-colnames(results)=c("ite","beta","corr","cens","prop","size","gt","wt","bst","gbj")
-
-lines=seq(1,length(results[,1]),by=2)
-results=results[lines,]
-results_size=results
-
-results=results_size[results_size[,6]==50,-6]
-combinaisons=read.table("data/simulations_table.txt",header=TRUE)
-for (i in 1:length(combinaisons[,1])){
-  
-  if (combinaisons[i,1]==0.3){
-    combinaisons[i,1]=0.05
-  }
-  if (combinaisons[i,1]==0.1){
-    combinaisons[i,1]=0.1
-  }
-  if (combinaisons[i,1]==0.5){
-    combinaisons[i,1]=0.2
-  }
-  
-}
-gt=numeric(length(combinaisons[,1]))
-wt=numeric(length(combinaisons[,1]))
-gbt=numeric(length(combinaisons[,1]))
-gbj=numeric(length(combinaisons[,1]))
-for (i in 1:length(combinaisons[,1])){
-  results_combinaison=results[results[,2]==combinaisons[i,4]&results[,3]==combinaisons[i,2]&results[,4]==combinaisons[i,3]&results[,5]==combinaisons[i,1],]
-  #results_combinaison=distinct(results_combinaison,ite, .keep_all= TRUE)
-  gt[i]=length(results_combinaison[results_combinaison[,6]<0.05,1])/length(results_combinaison[,1])
-  wt[i]=length(results_combinaison[results_combinaison[,7]<0.05,1])/length(results_combinaison[,1])
-  gbt[i]=length(results_combinaison[results_combinaison[,8]<0.05,1])/length(results_combinaison[,1])
-  gbj[i]=length(results_combinaison[results_combinaison[,9]<0.05,1])/length(results_combinaison[,1])
-}
-
-resultats=cbind(combinaisons,gbj,gt,wt,gbt)
-
-resultats_total=resultats
-library(reshape2)
-library(ggplot2)
-datas=melt(resultats,measure.vars=c("gbj","gt","wt","gbt"))
-colnames(datas)=c("cp","Corr","Censoring","beta","Method","Power")
-for (i in 1:length(datas[,1])){
[…]

### 2024-08-07 reporting/simulation_study/simu_rembrandt_breast.Rmd
reorganise and ensure reproducibility

--- 2023-09-22 reporting/simu_rembrandt_breast.Rmd
@@
 
-source(file = "functions/fct_report_simu_setting.R")
+source(file = here::here("functions/fct_report_simu_setting.R"))
 
-dfRembrandt <- readRDS("data/rembrandt/dfresvarcovar.rds")
-dfBreast <- readRDS("data/breast_cancer/dfresvarcovar.rds")
+dfRembrandt <- readRDS(here::here("data/rembrandt/dfresvarcovar.rds"))
+dfBreast <- readRDS(here::here("data/breast_cancer/dfresvarcovar.rds"))
 ```
@@
 Note : scenario D, E and F do not rely on correlation matrix to simulate variance covariance matrix so the those matrix are unlikely to be positive definite."}
-source("script/script_simulation/figures_simu_rembrandt_breast_v1.R")
+source(here::here("script/script_simulation/generate_figures/figures_simu_rembrandt_breast_v1.R"))
 
@@
 ```{r fig.height=14, fig.cap="Statistical power depending on simulation setting (second batch). N is the number of patients and NG the number of genes in the pathway."}
-source("script/script_simulation/figures_simu_rembrandt_breast_v2.R")
+source(here::here("script/script_simulation/generate_figures/figures_simu_rembrandt_breast_v2.R"))
 
@@
 
-```{r fig.height=6, fig.cap="Type-I error (Type = Z). N is the number of patients and NG the number of genes in the pathway."}
+```{r fig.height=6, fig.cap="Type-I error (Type = Z). N is the number of patients and NG the number of genes in the pathway. Vertical bars represent 95% confidence intervals with exact method."}
 plot_simu_alpha

### 2024-08-07 script/script_simulation/generate_figures/figures_simu_rembrandt_breast_v1.R

@@ -59,18 +59,18 @@ fct_generate_fig1 <- function(data,
 
 ##### data #####
 
-dfScenario1 <- readRDS("data/dfScenario_breast_rembrandt.rds") %>%
+dfScenario1 <- readRDS(here::here("data/dfScenario_breast_rembrandt.rds")) %>%
   tibble::rowid_to_column(var = "hp_row")
 
-dfScenario2 <- readRDS("data/dfScenario_breast_rembrandt_to_finish.rds") %>%
+dfScenario2 <- readRDS(here::here("data/dfScenario_breast_rembrandt_to_finish.rds")) %>%
   tibble::rowid_to_column(var = "hp_row")
 
-dfres1 <- list.files("data/result_simu_rembrandt_breast/result_9799043/", recursive = T, full.names = T) %>%
+dfres1 <- list.files(here::here("data/result_simu_rembrandt_breast/result_9799043/"), recursive = T, full.names = T) %>%
   lapply(FUN = readRDS) %>%
   bind_rows() %>%
   inner_join(dfScenario1, by = "hp_row")
 
-dfres2 <- list.files("data/result_simu_rembrandt_breast/result_9839694/", recursive = T, full.names = T) %>%
+dfres2 <- list.files(here::here("data/result_simu_rembrandt_breast/result_9839694/"), recursive = T, full.names = T) %>%
   lapply(FUN = readRDS) %>%
   bind_rows() %>%
   inner_join(dfScenario2, by = "hp_row")

### 2024-08-07 script/script_simulation/generate_figures/figures_simu_rembrandt_breast_v2.R

@@ -5,7 +5,7 @@ library(ggplot2)
 ls_folder_path <- c("data/result_11020779", "data/result_10187723", "data/result_11021425/")
 
 dfres <- lapply(ls_folder_path,
-                FUN = function(x) list.files(x, recursive = T, full.names = T)) %>%
+                FUN = function(x) list.files(here::here(x), recursive = T, full.names = T)) %>%
   unlist() %>%
   lapply(FUN = function(x) readRDS(x) %>%
            mutate(case = as.character(case),
@@ -91,7 +91,7 @@ plot_simu_alpha <- dfres %>%
         axis.text.x = element_text(angle = 45, hjust=1))
 
 # Time
-dfrestime <- readRDS(file = "data/result_11020779_time.rds") %>%
+dfrestime <- readRDS(file = here::here("data/result_11020779_time.rds")) %>%
   mutate(case_type = paste0("Case : ", case, " ; Type : ", type),
          n_p = paste0("N = ", nb_observations, " ; NG = ", nb_genes),
          prop_sig_gene = as.factor(prop_sig_gene)) %>%

### 2024-08-08 reporting/master_draft/draft_sgbj.qmd
update master report for publication and add README

---
title: "Draft - sGBJ"
author: "TF"
format: gfm
echo: false
message: false
warning: false
---

```{r results='hide'}
library(tidyverse)
library(survival)
library(ggsurvfit)
library(cowplot)

s <- lapply(list.files(path = here::here("functions"), full.names = TRUE),
            \(x) source(x))
rm(s)
```

# Simulation study

## Simulation setting

- number of genes : $NG = \{10, 50\}$
- number of individuals : $N = \{50, 100\}$
- proportion of significant genes = $0.2$
- proportion of censored individuals : $c = 0.3$
- Variance : $C_{jj} = 0.2$

### Case

- Case (I) : Overall correlation follows a non-standard beta $corr \sim NSBeta(20, 20, min=-1, max =1)$
- Case (II) : Correlation between significant genes $corr \sim NSBeta(10, 10, min=-1, max =1)$ and between other genes $corr \sim NSBeta(25, 25, min=-1, max =1)$

**Note : Case IV and V are not guaranteed to generate positive definite matrix, after the correlation values are sampled, if the matrix is not positive definite, the nearest positive definite matrix is computed using the algorithm defined by Higham 2002 (see Matrix::nearPD). This correction was not implemented for scenario D, E and F.**

- Case (III) : Correlation between significant genes is $0.2$, correlation with non significant genes is $0$.

### Type

- Type (Z) : $\beta = 0$ evaluates the type-I error
- Type (A) : $\beta \sim \mathcal{N}(0, 0.4^2)$. Correlation matrix of $\beta$ is the same as correlation matrix of genes.
- Type (B) : half of genes follows $\beta \sim \mathcal{N}(-0.4, 0.2^2)$ and half follows $\beta \sim \mathcal{N}(0.4, 0.2^2)$. Correlation matrix of $\beta$ is the same as correlation matrix of genes.
- Type (C) : half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$. Correlation matrix of $\beta$ is the same as correlation matrix of genes.

## Results of simulation study

```{r fig.height=10, fig.cap="Panel A represents the statistical power depending on simulation setting. Panel B represents the type-I error, expected to be 5%, depending on simulation setting."}
source(here::here("script/script_simulation/generate_figures/figures_simu_rembrandt_breast_v2.R"))

plot_simulation_power_alpha
ggsave(plot_simulation_power_alpha,
       filename = here::here("reporting/master_draft/figures/plot_simulation_power_alpha.pdf"),
       height = 10,
       width = 8.5)
```

# Applications

## Breast cancer

```{r}
lsdata <- readRDS(file = here::here("data/breast_cancer/datamanaged.rds"))

ls_pathways <- lsdata$pathways
df_data <- lsdata$data
```

### Description

```{r descbreast}
##### Description
df_data |> 
  dplyr::select(age,
                event,
                time,
                grade) |> 
  gtsummary::tbl_summary()
```

Here are the kaplan meier curves for the two types of disease:

```{r kmbreastcancer, fig.cap="Kaplan meier curves for Astro, Oligo"}
p_type <- survfit2(Surv(time=time, event=event) ~ grade,
               data = df_data) %>% 
  ggsurvfit(linewidth = 1.1) +
  add_confidence_interval() +
  add_risktable(risktable_stats = "n.risk") +
  scale_color_manual(values = c("#F48C06", "#DC2F02", "#9D0208")) +
  scale_fill_manual(values = c("#F48C06", "#DC2F02", "#9D0208")) +
  labs(fill = "Grade", color = "Grade")
```

### Breast cancer pathway analysis

Analysis of breast cancer pathway adjusted on age, overall and stratified by grade.

```{r importbreastcancerpathway}
df_breast_cancer_pathway <- lapply(list.files(here::here("results/breast_cancer/"), full.names = T),
                               readRDS) |> 
  bind_rows() |> 
  select(-starts_with("time_"), -nb_genes) |> 
  tidyr::pivot_longer(cols = c("sGBJ", "GBT", "Wald", "GT"),
                      names_to = "method",
                      values_to = "p_value") |> 
  group_by(grade, method) |> 
  mutate(p_value_BH = p.adjust(p_value, method = "BH"),
         method = factor(method,
                         levels = c("sGBJ", "GBT", "GT", "Wald"),
                         labels = c("sGBJ", "Global Boost Test", "Global Test", "Wald")),
         grade = factor(grade,
                         levels = c("all", "1", "2", "3"),
                         labels = c("All", "Grade 1", "Grade 2", "Grade 3"))) |> 
  ungroup() |> 
  mutate(significant = as.integer(p_value < 0.05),
         significant_bh = as.integer(p_value_BH < 0.05)) |> 
  mutate(method = forcats::fct_relevel(method,
                                       "sGBJ",
                                       "Global Test",
                                       "Wald",
                                       "Global Boost Test"))
```

```{r nbsignbreast, fig.height=5, fig.cap="Number of significant pathways by method."}
plot_breast_nb_genesets <- df_breast_cancer_pathway |> 
  group_by(grade, method) |> 
  summarise(nb_sign_bh = sum(significant_bh),
            nb_sign = sum(significant),
            .groups = "drop") |> 
  tidyr::pivot_longer(cols = starts_with("nb_sign")) |> 
  mutate(method = forcats::fct_rev(method),
         method = forcats::fct_recode(method, "Wald Test" = "Wald"),
         name = factor(name,
                       labels = c("Raw", "Benjamini-Hockberg"),
                       levels = c("nb_sign", "nb_sign_bh"))) |> 
  ggplot(mapping = aes(x = method, fill = method, y = value, linetype = name)) +
  geom_bar(stat = "identity", position = position_dodge(), color = "black", size = .6) +
  facet_grid(grade ~ ., scales = "free_y") +
  scale_fill_viridis_d(guide = "none") +
  scale_linetype_manual(values = c(1,2)) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  labs(x = "Method",
       y = "Nb of significant pathways",
       linetype = "")
```

```{r upsetplotbreast, fig.cap="Upset plot of the Benjamini-Hockberg p-value agreement according to the different methods", fig.height=10}
ls_plots <- df_breast_cancer_pathway |>
  filter(significant_bh == 1) |>
  group_by(pathway, grade) |>
  summarise(methods = list(unique(as.character(method)))) |>
  group_by(grade) |>
  group_split() |>
  lapply(FUN = function(df_i){
    title_i <- unique(df_i$grade)
    plot_i <- df_i |>
      ggplot(aes(x=methods)) +
      geom_bar() +
      ggupset::scale_x_upset(n_intersections = 6, order_by = "freq") +
      theme_minimal() +
      theme(plot.margin = margin(10, 10, 10, 50)) +
      labs(title = title_i,
           x = "",
           y = "")

    return(plot_i)
  })
plot_upset_breast <- ggpubr::ggarrange(plotlist = ls_plots,
                                       ncol = 3)

plot_upset_breast <- ggpubr::annotate_figure(plot_upset_breast,
                                             left = "Nb significant pathways",
                                             bottom = "Method")
```

```{r figpvaluesmethodsbreastcancer, fig.cap="Raw p-values in function of the ordered ranks of sGBJ for the 4 methods (sGBJ , global boost test, Wald test and global test), with the 5% threshold and the Benjamini Hochberg limit, computed for astrocytoma, oligodendroglioma and all patients. Nota Bene: The Benjamini Hochberg limit only applies for the sGBJ method, as the ranks are computed for sGBJ only.", fig.height=8}
nb_pathway = df_breast_cancer_pathway |> 
  pull(pathway) |> 
  unique() |> 
  length()

df_sGBJ_rank <- df_breast_cancer_pathway |> 
  group_by(grade) |> 
  filter(method == "sGBJ") |> 
  mutate(sGBJ_rank = rank(p_value)) |> 
  ungroup() |> 
  select(pathway, grade, sGBJ_rank)

df_BH_limit <- data.frame(sGBJ_rank = seq(1:nb_pathway)) |> 
  mutate(p_value = 0.05/nb_pathway * sGBJ_rank)

plot_pval_breast <- df_breast_cancer_pathway |> 
  left_join(df_sGBJ_rank, by = c("pathway", "grade")) |> 
  ggplot(mapping = aes(x = sGBJ_rank, y = p_value)) +
  geom_point(mapping = aes(color = method)) +
  scale_color_viridis_d(direction = -1) +
  labs(color = "") +
  ggnewscale::new_scale_color() +
  geom_hline(mapping = aes(color = "5% threshold",
                           linetype = "5% threshold", yintercept = 0.05),
                         linewidth = 1) +
  # geom_line(data = df_BH_limit,
  #           linewidth = 1,
  #           mapping = aes(color = "BH limit", linetype = "BH limit")) +
  scale_color_manual(values = c("red", "black")) +
  scale_linetype_manual(values = c(2,1)) +
  labs(color = "", linetype = "") +

  facet_grid(grade ~ ., scales = "free_y") +
  scale_y_log10() +
  annotation_logticks(sides = "l") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.box="vertical") +
  labs(x = "SGBJ rank", y = "Raw p-value")
```

```{r fig.height=10}
plot_breast <- ggdraw() +
  draw_plot(p_type, x = 0, y = .74, width = .4, height = .22) +
  draw_plot(plot_breast_nb_genesets, x = 0, y = .24, width = .4, height = .5) +
  draw_plot(plot_upset_breast, x = 0, y = 0, width = 1, height = 0.25) +
  draw_plot(plot_pval_breast, x = .4, y = .25, width = .6, height = .75) +
  draw_plot_label(label = c("A", "B", "C", "D"), size = 15,
                  x = c(0, 0, 0, .4), y = c(1, .75, .26, 1))

plot_breast

ggsave(plot_breast,
       filename = here::here("reporting/master_draft/figures/plot_breast.pdf"),
       height = 9,
       width = 7)

```

NB about GT test : https://www.bioconductor.org/packages/release/bioc/vignettes/globaltest/inst/doc/GlobalTest.pdf : "Because permutations require an exchangeable null hypothesis, such a permutation p-value is only available for the linear model and for the exchangeable null hypotheses ~1 and ~0 in other models."

## Rembrandt

Analysis of rembrandt pathway adjusted on age and sex, overall and stratified by tumor type.

```{r results='hide'}
df_rembrandt <- fct_load_clean_rembrandt()

pathways = GSA::GSA.read.gmt(here::here("data/rembrandt/c6.all.v7.1.symbols.gmt"))
liste_pathways=pathways$genesets
liste_names=pathways$geneset.names
names(liste_pathways) <- liste_names
```

### Describe Rembrandt cohort

```{r descrembrandt}
##### Description
df_rembrandt |> 
  dplyr::select(AGE_RANGE,
                GENDER,
                EVENT_OS,
                OVERALL_SURVIVAL_MONTHS,
                DISEASE_TYPE) |> 
  gtsummary::tbl_summary(by = DISEASE_TYPE)
```

Here are the kaplan meier curves for the two types of disease:

```{r kmrembrandt, fig.cap="Kaplan meier curves for Astro, Oligo"}
p_type <- survfit2(Surv(time=OVERALL_SURVIVAL_MONTHS, event=EVENT_OS) ~ DISEASE_TYPE,
               data = df_rembrandt |> 
                 mutate(DISEASE_TYPE = factor(DISEASE_TYPE,
                                              levels = c("OLIGODENDROGLIOMA",
                                                         "ASTROCYTOMA"),
                                              labels = c("Oligo",
                                                         "Astro")))) %>% 
  ggsurvfit(linewidth = 1.1) +
  add_confidence_interval() +
  add_risktable(risktable_stats = "n.risk") +
  scale_color_manual(values = c("#8338EC", "#FFB703")) +
  scale_fill_manual(values = c("#8338EC", "#FFB703")) +
  labs(fill = "", color = "")
```

### Rembrandt pathway analysis

```{r importrembrandtpathway}
df_rembrandt_pathway <- lapply(list.files(here::here("results/rembrandt/"), full.names = T),
                               readRDS) |> 
  bind_rows() |> 
  select(-starts_with("time_"), -nb_genes) |> 
  tidyr::pivot_longer(cols = c("sGBJ", "GBT", "Wald", "GT"),
                      names_to = "method",
                      values_to = "p_value") |> 
  group_by(grade, method) |> 
  mutate(p_value_BH = p.adjust(p_value, method = "BH"),
         method = factor(method,
                         levels = c("sGBJ", "GBT", "GT", "Wald"),
                         labels = c("sGBJ", "Global Boost Test", "Global Test", "Wald")),
         grade = factor(grade,
                         levels = c("all", "OLIGODENDROGLIOMA", "ASTROCYTOMA"),
                         labels = c("All", "Oligodendroglioma", "Astrocytoma"))) |> 
  ungroup() |> 
  mutate(significant = as.integer(p_value < 0.05),
         significant_bh = as.integer(p_value_BH < 0.05)) |> 
  mutate(method = forcats::fct_relevel(method,
                                       "sGBJ",
                                       "Global Test",
                                       "Wald",
                                       "Global Boost Test"))
```

```{r nbsignrembrandt, fig.height=5, fig.cap="Number of significant pathways by method."}
plot_rembrandt_nb_genesets <- df_rembrandt_pathway |> 
  group_by(grade, method) |> 
  summarise(nb_sign_bh = sum(significant_bh),
            nb_sign = sum(significant),
            .groups = "drop") |> 
  tidyr::pivot_longer(cols = starts_with("nb_sign")) |> 
  mutate(method = forcats::fct_rev(method),
         method = forcats::fct_recode(method, "Wald Test" = "Wald"),
         name = factor(name,
                       labels = c("Raw", "Benjamini-Hockberg"),
                       levels = c("nb_sign", "nb_sign_bh")),
         grade  = forcats::fct_recode(grade,
                                      "Oligo" = "Oligodendroglioma",
                                      "Astro" = "Astrocytoma")) |> 
  ggplot(mapping = aes(x = method, fill = method, y = value, linetype = name)) +
  geom_bar(stat = "identity", position = position_dodge(), color = "black", size = .6) +
  facet_grid(grade ~ ., scales = "free_y") +
  scale_fill_viridis_d(guide = "none") +
  scale_linetype_manual(values = c(1,2)) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  labs(x = "Method",
       y = "Nb of significant pathways",
       linetype = "")
```

```{r upsetplotrembrandt, fig.cap="Upset plot of the Benjamini-Hockberg p-value agreement according to the different methods", fig.height=8}
ls_plots <- df_rembrandt_pathway |>
  filter(significant_bh == 1) |>
  group_by(pathway, grade) |>
  summarise(methods = list(unique(as.character(method)))) |>
  group_by(grade) |>
  group_split() |>
  lapply(FUN = function(df_i){
    title_i <- unique(df_i$grade)
    plot_i <- df_i |>
      ggplot(aes(x=methods)) +
      geom_bar() +
      ggupset::scale_x_upset(n_intersections = 6, order_by = "freq") +
      theme_minimal() +
      theme(plot.margin = margin(10, 10, 10, 50)) +
      labs(title = title_i,
           x = "",
           y = "")

    return(plot_i)
  })
plot_upset_rembrandt <- ggpubr::ggarrange(plotlist = ls_plots,
                                       ncol = length(ls_plots))

plot_upset_rembrandt <- ggpubr::annotate_figure(plot_upset_rembrandt,
                                                left = "Nb significant pathways",
                                                bottom = "Method")
```

```{r figpvaluesmethodsrembrandt, fig.cap="Raw p-values in function of the ordered ranks of sGBJ for the 4 methods (sGBJ , global boost test, Wald test and global test), with the 5% threshold and the Benjamini Hochberg limit, computed for astrocytoma, oligodendroglioma and all patients. Nota Bene: The Benjamini Hochberg limit only applies for the sGBJ method, as the ranks are computed for sGBJ only.", fig.height=8}
nb_pathway = df_rembrandt_pathway |> 
  pull(pathway) |> 
  unique() |> 
  length()

df_sGBJ_rank <- df_rembrandt_pathway |> 
  group_by(grade) |> 
  filter(method == "sGBJ") |> 
  mutate(sGBJ_rank = rank(p_value)) |> 
  ungroup() |> 
  select(pathway, grade, sGBJ_rank)

df_BH_limit <- data.frame(sGBJ_rank = seq(1:nb_pathway)) |> 
  mutate(p_value = 0.05/nb_pathway * sGBJ_rank)

plot_pval_rembrandt <- df_rembrandt_pathway |> 
  left_join(df_sGBJ_rank, by = c("pathway", "grade")) |> 
  ggplot(mapping = aes(x = sGBJ_rank, y = p_value)) +
  geom_point(mapping = aes(color = method)) +
  scale_color_viridis_d(direction = -1) +
  labs(color = "") +
  ggnewscale::new_scale_color() +
  geom_hline(mapping = aes(color = "5% threshold",
                           linetype = "5% threshold", yintercept = 0.05),
                         linewidth = 1) +
  # geom_line(data = df_BH_limit,
  #           linewidth = 1,
  #           mapping = aes(color = "BH limit", linetype = "BH limit")) +
  scale_color_manual(values = c("red", "black")) +
  scale_linetype_manual(values = c(2,1)) +
  labs(color = "", linetype = "") +

  facet_grid(grade ~ ., scales = "free_y") +
  scale_y_log10() +
  annotation_logticks(sides = "l") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.box="vertical") +
  labs(x = "SGBJ rank", y = "Raw p-value")
```

NB about GT test : https://www.bioconductor.org/packages/release/bioc/vignettes/globaltest/inst/doc/GlobalTest.pdf : "Because permutations require an exchangeable null hypothesis, such a permutation p-value is only available for the linear model and for the exchangeable null hypotheses ~1 and ~0 in other models."

```{r fig.height=10}
plot_rembrandt <- ggdraw() +
  draw_plot(p_type, x = 0, y = .74, width = .4, height = .22) +
  draw_plot(plot_rembrandt_nb_genesets, x = 0, y = .24, width = .4, height = .5) +
  draw_plot(plot_upset_rembrandt, x = 0, y = 0, width = 1, height = 0.25) +
  draw_plot(plot_pval_rembrandt, x = .4, y = .25, width = .6, height = .75) +
  draw_plot_label(label = c("A", "B", "C", "D"), size = 15,
                  x = c(0, 0, 0, .4), y = c(1, .75, .26, 1))

plot_rembrandt

ggsave(plot_rembrandt,
       filename = here::here("reporting/master_draft/figures/plot_rembrandt.pdf"),
       height = 9,
       width = 7)

```

### 2024-08-08 script/script_simulation/generate_figures/figures_simu_rembrandt_breast_v2.R

@@ -127,3 +127,83 @@ plot_time <- dfrestime %>%
          color = guide_legend(nrow = 2)) +
   theme_bw() +
   theme(legend.position = "bottom")
+
+############### draft figure ######################
+dfres_draft <- dfres %>%
+  mutate(case_type = gsub(x = case_type, pattern = "Type : G", replacement = "Type : A"),
+         case_type = gsub(x = case_type, pattern = "Type : H", replacement = "Type : B"),
+         case_type = gsub(x = case_type, pattern = "Type : I", replacement = "Type : C"),
+         case_type = gsub(x = case_type, pattern = "Case : 4", replacement = "Case : (I)"),
+         case_type = gsub(x = case_type, pattern = "Case : 5", replacement = "Case : (II)"),
+         case_type = gsub(x = case_type, pattern = "Case : 6", replacement = "Case : (III)"))
+
+dfres_draft |> 
+  filter(type != "Z") |> 
+  group_by(case_type, n_p) |> 
+  summarise(power = max(power))
+
+plot_power_simu <- dfres_draft |> 
+  filter(type != "Z") %>%
+  group_by(case_type, n_p) |> 
+  mutate(max_power = max(power)) |>
+  ungroup() |> 
+  ggplot(mapping = aes(x = method,
+                       yintercept = max_power,
+                       y = power,
+                       ymin = power_lower,
+                       ymax = power_upper,
+                       group = method,
+                       fill = method)) +
+  geom_bar(stat = "identity") +
+  geom_errorbar(width = .1) +
+  geom_hline(mapping = aes(yintercept = max_power), color = "grey") +
+  scale_fill_viridis_d() +
+  scale_color_manual(values = c("black", "red")) +
+  scale_y_continuous(limits = c(0,1), breaks = c(0, .5, 1)) +
+  facet_grid(case_type ~ n_p) +
+  labs(x = "Method",
+       y = "Statistical Power",
+       fill = "",
+       color = "Censoring proportion",
+       lty = "Censoring proportion") +
+  guides(fill = guide_legend(nrow = 4),
+         lty = guide_legend(nrow = 2),
+         color = guide_legend(nrow = 2)) +
+  theme_minimal() +
+  theme(legend.position = "right",
+        axis.text.x = element_text(angle = 45, hjust=1),
+        strip.text.y.right = element_text(angle = 0))
+
+plot_alpha_simu <- dfres_draft %>%
+  filter(type == "Z") %>%
[…]

### 2024-08-12 reporting/master_draft/draft_sgbj.qmd
update figures

@@
 library(cowplot)
+library(gt)
 
@@
                        levels = c("nb_sign", "nb_sign_bh"))) |> 
-  ggplot(mapping = aes(x = method, fill = method, y = value, linetype = name)) +
-  geom_bar(stat = "identity", position = position_dodge(), color = "black", size = .6) +
+  ggplot(mapping = aes(x = method, fill = method, y = value, alpha = name)) +
+  geom_bar(stat = "identity", position = position_dodge(), size = .6) +
   facet_grid(grade ~ ., scales = "free_y") +
@@
   scale_linetype_manual(values = c(1,2)) +
+  scale_alpha_manual(values = c(0.5, 1)) +
   theme_minimal() +
@@
        y = "Nb of significant pathways",
-       linetype = "")
+       alpha = "")
 ```
@@
 plot_pval_breast <- df_breast_cancer_pathway |> 
+  filter(p_value != 0) |> 
   left_join(df_sGBJ_rank, by = c("pathway", "grade")) |> 
@@
   labs(x = "SGBJ rank", y = "Raw p-value")
+```
+
+```{r}
+df_breast_cancer_pathway |> 
+  group_by(grade, method) |> 
+  summarise(prop_sign = mean(significant_bh)) |> 
+  gt::gt() |> 
+  tab_header(
+    title = "Proportion of significant pathway by method and glioma"
+  ) %>%
+  fmt_number(
+    columns = prop_sign,
+    decimals = 3
+  )
 ```
@@
   add_risktable(risktable_stats = "n.risk") +
-  scale_color_manual(values = c("#8338EC", "#FFB703")) +
-  scale_fill_manual(values = c("#8338EC", "#FFB703")) +
+  scale_color_manual(values = c("#2191FB", "#E94F37")) +
+  scale_fill_manual(values = c("#2191FB", "#E94F37")) +
   labs(fill = "", color = "")
@@
 
+```{r}
+df_rembrandt_pathway |> 
+  group_by(grade, method) |> 
+  summarise(prop_sign = mean(significant_bh)) |> 
+  gt::gt() |> 
+  tab_header(
+    title = "Proportion of significant pathway by method and glioma"
+  ) %>%
+  fmt_number(
+    columns = prop_sign,
+    decimals = 3
+  )
+```
+
 ```{r nbsignrembrandt, fig.height=5, fig.cap="Number of significant pathways by method."}
@@
                                       "Astro" = "Astrocytoma")) |> 
-  ggplot(mapping = aes(x = method, fill = method, y = value, linetype = name)) +
-  geom_bar(stat = "identity", position = position_dodge(), color = "black", size = .6) +
+  ggplot(mapping = aes(x = method, fill = method, y = value, alpha = name)) +
+  geom_bar(stat = "identity", position = position_dodge(), size = .6) +
   facet_grid(grade ~ ., scales = "free_y") +
+  scale_alpha_manual(values = c(0.5, 1)) +
   scale_fill_viridis_d(guide = "none") +
@@
 plot_pval_rembrandt <- df_rembrandt_pathway |> 
+  filter(p_value != 0) |> 
   left_join(df_sGBJ_rank, by = c("pathway", "grade")) |> 

### 2024-08-12 script/script_simulation/generate_figures/figures_simu_rembrandt_breast_v2.R

@@ -145,7 +145,8 @@ dfres_draft |>
 plot_power_simu <- dfres_draft |> 
   filter(type != "Z") %>%
   group_by(case_type, n_p) |> 
-  mutate(max_power = max(power)) |>
+  mutate(max_power = max(power),
+         rank = as.factor(dense_rank(-power))) |>
   ungroup() |> 
   ggplot(mapping = aes(x = method,
                        yintercept = max_power,
@@ -154,18 +155,17 @@ plot_power_simu <- dfres_draft |>
                        ymax = power_upper,
                        group = method,
                        fill = method)) +
-  geom_bar(stat = "identity") +
+  geom_bar(stat = "identity", width = 0.5, linewidth = 1) +
   geom_errorbar(width = .1) +
   geom_hline(mapping = aes(yintercept = max_power), color = "grey") +
   scale_fill_viridis_d() +
-  scale_color_manual(values = c("black", "red")) +
+  # scale_color_manual(values = c("red", "orange", "grey", "black")) +
   scale_y_continuous(limits = c(0,1), breaks = c(0, .5, 1)) +
   facet_grid(case_type ~ n_p) +
   labs(x = "Method",
        y = "Statistical Power",
        fill = "",
-       color = "Censoring proportion",
-       lty = "Censoring proportion") +
+       color = "Rank") +
   guides(fill = guide_legend(nrow = 4),
          lty = guide_legend(nrow = 2),
          color = guide_legend(nrow = 2)) +
@@ -173,7 +173,7 @@ plot_power_simu <- dfres_draft |>
   theme(legend.position = "right",
         axis.text.x = element_text(angle = 45, hjust=1),
         strip.text.y.right = element_text(angle = 0))
-
+plot_power_simu
 plot_alpha_simu <- dfres_draft %>%
   filter(type == "Z") %>%
   ggplot(mapping = aes(x = method,
@@ -187,7 +187,6 @@ plot_alpha_simu <- dfres_draft %>%
              size = 3) +
   geom_hline(yintercept = 0.05, lty = 2) +
   scale_fill_viridis_d() +
-  scale_color_manual(values = c("black", "red")) +
   scale_y_continuous(breaks = c(0.0125, 0.05), trans = "log") +
   facet_grid(case_type ~ n_p) +
   labs(x = "Method",

### 2024-08-12 script/old_scripts/GBJ_simulations.R
remove old files

@@ -1,242 +1,0 @@
-## package install
-# BiocManager::install("globaltest")
-# install.packages("coin")
-# install.packages("mboost")
-# install.packages("globalboosttest", repos = c("http://R-Forge.R-project.org"), dep = TRUE)
-
-library(mvtnorm)
-library(GBJ)
-library(survival)
-library(globaltest)
-library(globalboosttest)
-ite_base=as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
-# ite_base=1
-ite=(ite_base%%48)+1
-
-if (ite_base<=1920){
-ns=50
-}
-
-if (ite_base>1920 & ite_base<=3840){
-ns=80
-}
-
-
-
-params=read.table("data/simulations_table.txt",header=TRUE)
-# params=read.table("data/simulations_table.txt",header=TRUE)[1:2,]
-if (params[ite,1]==0.1){
-gsp=0.3
-}
-if (params[ite,1]==0.3){
-gsp=0.4
-}
-
-if (params[ite,1]==0.5){
-gsp=0.5
-}
-
-cp=params[ite,3]
-corr=params[ite,2]
-#corr=4
-beta=params[ite,4] #beta type: 1(positively) or 2(random)
-#corr=1 #correlation type: 1(independence), 2(equal), 3(AR), 4(unstructured)
-cenp=1 #censoring type: 1(no), 2(30%)
-sigp=1 #significant gene proportion: 1(10%), 2(30%), 3(50%))
-tau<-0.5 #correlation coefficient between genes
-acp<-c(0.0,0.3) #censoring fraction
-agsp<-c(0.1,0.3,0.5) #significant gene set proportion
-nstat<-3 #test statistics of interest: GSEA with p=0,0.5,1, global test, and Adewale test
-ng<-200 #gene size
-#ns<-80 #sanple size
-B<-1000 #permuation
-# B<-2 #permuation
-iter<-50 #iteration
-# iter<-2 #iteration
-gsize<-50 #gene set size
-#gsp<-agsp[sigp]
-ngs<-gsize*gsp
-ncov<-ngs*(ngs-1)/2
-ncov2<-ng*(ng-1)/2
[…]

### 2024-08-12 script/old_scripts/GBJ_simulations_2.R

@@ -1,242 +1,0 @@
-## package install
-# BiocManager::install("globaltest")
-# install.packages("coin")
-# install.packages("mboost")
-# install.packages("globalboosttest", repos = c("http://R-Forge.R-project.org"), dep = TRUE)
-
-library(mvtnorm)
-library(GBJ)
-library(survival)
-library(globaltest)
-library(globalboosttest)
-ite_base=as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))+999
-# ite_base=1
-ite=(ite_base%%48)+1
-
-if (ite_base<=1920){
-ns=50
-}
-
-if (ite_base>1920 & ite_base<=3840){
-ns=80
-}
-
-
-
-params=read.table("data/simulations_table.txt",header=TRUE)
-# params=read.table("data/simulations_table.txt",header=TRUE)[1:2,]
-if (params[ite,1]==0.1){
-gsp=0.3
-}
-if (params[ite,1]==0.3){
-gsp=0.4
-}
-
-if (params[ite,1]==0.5){
-gsp=0.5
-}
-
-cp=params[ite,3]
-corr=params[ite,2]
-#corr=4
-beta=params[ite,4] #beta type: 1(positively) or 2(random)
-#corr=1 #correlation type: 1(independence), 2(equal), 3(AR), 4(unstructured)
-cenp=1 #censoring type: 1(no), 2(30%)
-sigp=1 #significant gene proportion: 1(10%), 2(30%), 3(50%))
-tau<-0.5 #correlation coefficient between genes
-acp<-c(0.0,0.3) #censoring fraction
-agsp<-c(0.1,0.3,0.5) #significant gene set proportion
-nstat<-3 #test statistics of interest: GSEA with p=0,0.5,1, global test, and Adewale test
-ng<-200 #gene size
-#ns<-80 #sanple size
-B<-1000 #permuation
-# B<-2 #permuation
-iter<-50 #iteration
-# iter<-2 #iteration
-gsize<-50 #gene set size
-#gsp<-agsp[sigp]
-ngs<-gsize*gsp
-ncov<-ngs*(ngs-1)/2
-ncov2<-ng*(ng-1)/2
[…]

### 2024-08-12 script/old_scripts/GBJ_simulations_3.R

@@ -1,242 +1,0 @@
-## package install
-# BiocManager::install("globaltest")
-# install.packages("coin")
-# install.packages("mboost")
-# install.packages("globalboosttest", repos = c("http://R-Forge.R-project.org"), dep = TRUE)
-
-library(mvtnorm)
-library(GBJ)
-library(survival)
-library(globaltest)
-library(globalboosttest)
-ite_base=as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))+999+999
-# ite_base=1
-ite=(ite_base%%48)+1
-
-if (ite_base<=1920){
-ns=50
-}
-
-if (ite_base>1920 & ite_base<=3840){
-ns=80
-}
-
-
-
-params=read.table("data/simulations_table.txt",header=TRUE)
-# params=read.table("data/simulations_table.txt",header=TRUE)[1:2,]
-if (params[ite,1]==0.1){
-gsp=0.3
-}
-if (params[ite,1]==0.3){
-gsp=0.4
-}
-
-if (params[ite,1]==0.5){
-gsp=0.5
-}
-
-cp=params[ite,3]
-corr=params[ite,2]
-#corr=4
-beta=params[ite,4] #beta type: 1(positively) or 2(random)
-#corr=1 #correlation type: 1(independence), 2(equal), 3(AR), 4(unstructured)
-cenp=1 #censoring type: 1(no), 2(30%)
-sigp=1 #significant gene proportion: 1(10%), 2(30%), 3(50%))
-tau<-0.5 #correlation coefficient between genes
-acp<-c(0.0,0.3) #censoring fraction
-agsp<-c(0.1,0.3,0.5) #significant gene set proportion
-nstat<-3 #test statistics of interest: GSEA with p=0,0.5,1, global test, and Adewale test
-ng<-200 #gene size
-#ns<-80 #sanple size
-B<-1000 #permuation
-# B<-2 #permuation
-iter<-50 #iteration
-# iter<-2 #iteration
-gsize<-50 #gene set size
-#gsp<-agsp[sigp]
-ngs<-gsize*gsp
-ncov<-ngs*(ngs-1)/2
-ncov2<-ng*(ng-1)/2
[…]

### 2024-08-12 script/old_scripts/GBJ_simulations_4.R

@@ -1,242 +1,0 @@
-## package install
-# BiocManager::install("globaltest")
-# install.packages("coin")
-# install.packages("mboost")
-# install.packages("globalboosttest", repos = c("http://R-Forge.R-project.org"), dep = TRUE)
-
-library(mvtnorm)
-library(GBJ)
-library(survival)
-library(globaltest)
-library(globalboosttest)
-ite_base=as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))+999+999+999
-# ite_base=1
-ite=(ite_base%%48)+1
-
-if (ite_base<=1920){
-ns=50
-}
-
-if (ite_base>1920 & ite_base<=3840){
-ns=80
-}
-
-
-
-params=read.table("data/simulations_table.txt",header=TRUE)
-# params=read.table("data/simulations_table.txt",header=TRUE)[1:2,]
-if (params[ite,1]==0.1){
-gsp=0.3
-}
-if (params[ite,1]==0.3){
-gsp=0.4
-}
-
-if (params[ite,1]==0.5){
-gsp=0.5
-}
-
-cp=params[ite,3]
-corr=params[ite,2]
-#corr=4
-beta=params[ite,4] #beta type: 1(positively) or 2(random)
-#corr=1 #correlation type: 1(independence), 2(equal), 3(AR), 4(unstructured)
-cenp=1 #censoring type: 1(no), 2(30%)
-sigp=1 #significant gene proportion: 1(10%), 2(30%), 3(50%))
-tau<-0.5 #correlation coefficient between genes
-acp<-c(0.0,0.3) #censoring fraction
-agsp<-c(0.1,0.3,0.5) #significant gene set proportion
-nstat<-3 #test statistics of interest: GSEA with p=0,0.5,1, global test, and Adewale test
-ng<-200 #gene size
-#ns<-80 #sanple size
-B<-1000 #permuation
-# B<-2 #permuation
-iter<-50 #iteration
-# iter<-2 #iteration
-gsize<-50 #gene set size
-#gsp<-agsp[sigp]
-ngs<-gsize*gsp
-ncov<-ngs*(ngs-1)/2
-ncov2<-ng*(ng-1)/2
[…]

### 2024-08-12 script/old_scripts/comparaisons_sets_bis_corr_beta6_2.R

@@ -1,233 +1,0 @@
-
-library(mvtnorm)
-library(GBJ)
-library(survival)
-library(globaltest)
-library(globalboosttest)
-ite_base=as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))+999+999
-ite=(ite_base%%48)+1
-
-if (ite_base<=1920){
-ns=50
-}
-
-if (ite_base>1920 & ite_base<=3840){
-ns=80
-}
-
-
-
-params=read.table("data/simulations_table.txt",header=TRUE)
-if (params[ite,1]==0.1){
-gsp=0.05
-}
-if (params[ite,1]==0.3){
-gsp=0.1
-}
-
-if (params[ite,1]==0.5){
-gsp=0.2
-}
-
-cp=params[ite,3]
-corr=params[ite,2]
-#corr=4
-beta=params[ite,4] #beta type: 1(positively) or 2(random)
-#corr=1 #correlation type: 1(independence), 2(equal), 3(AR), 4(unstructured)
-cenp=1 #censoring type: 1(no), 2(30%)
-sigp=1 #significant gene proportion: 1(10%), 2(30%), 3(50%))
-tau<-0.5 #correlation coefficient between genes
-acp<-c(0.0,0.3) #censoring fraction
-agsp<-c(0.1,0.3,0.5) #significant gene set proportion
-nstat<-3 #test statistics of interest: GSEA with p=0,0.5,1, global test, and Adewale test
-ng<-200 #gene size
-#ns<-80 #sanple size
-B<-1000 #permuation
-iter<-50 #iteration
-gsize<-50 #gene set size
-#gsp<-agsp[sigp]
-ngs<-gsize*gsp
-ncov<-ngs*(ngs-1)/2
-ncov2<-ng*(ng-1)/2
-p.pvalue<-matrix(0,nrow=nstat,ncol=iter)
-ots<-matrix(0,nrow=nstat,ncol=iter)
-p5<-rep(0,nstat)
-cf<-rep(0,iter)
-
-stats2=function(data,ng,gsize,nstat){
-  time<-data[,1]
-  csg<-data[,2]
-  x<-data[,3:ncol(data)]
-  wtcx<-matrix(0,nrow=2,ncol=ng)
-  ucox<-rep(0,ng)
-  escx0<-0
-  escx1<-0
-  gt<-0
-  aw<-0
-  bst<-0
[…]

### 2024-08-12 script/old_scripts/generation_figure.R

@@ -1,80 +1,0 @@
-### import data
-results=read.table("results/comparaison_survival_sets_bis_corr_beta6_bis_2000.txt",
-                   col.names = c("ite","beta","corr","cens","prop","size","gt","wt","bst","gbj"))
-combinaisons=read.table("data/simulations_table.txt",header=TRUE)
-
-# Je comprends rien à cette section
-###
-lines=seq(1,length(results[,1]),by=2)
-results=results[lines,]
-results_size=results
-results=results_size[results_size[,6]==50,-6]
-###
-
-### change les cp
-# 0.3 -> 0.4
-# 0.1 -> 0.3
-# 0.5 -> 0.5
-for (i in 1:length(combinaisons[,1])){
-  
-  if (combinaisons[i,1]==0.3){
-    combinaisons[i,1]=0.4
-  }
-  if (combinaisons[i,1]==0.1){
-    combinaisons[i,1]=0.3
-  }
-  if (combinaisons[i,1]==0.5){
-    combinaisons[i,1]=0.5
-  }
-  
-}
-gt=numeric(length(combinaisons[,1]))
-wt=numeric(length(combinaisons[,1]))
-gbt=numeric(length(combinaisons[,1]))
-gbj=numeric(length(combinaisons[,1]))
-for (i in 1:length(combinaisons[,1])){
-  results_combinaison=results[results[,2]==combinaisons[i,4]&results[,3]==combinaisons[i,2]&results[,4]==combinaisons[i,3]&results[,5]==combinaisons[i,1],]
-  #results_combinaison=distinct(results_combinaison,ite, .keep_all= TRUE)
-  gt[i]=length(results_combinaison[results_combinaison[,6]<0.05,1])/length(results_combinaison[,1])
-  wt[i]=length(results_combinaison[results_combinaison[,7]<0.05,1])/length(results_combinaison[,1])
-  gbt[i]=length(results_combinaison[results_combinaison[,8]<0.05,1])/length(results_combinaison[,1])
-  gbj[i]=length(results_combinaison[results_combinaison[,9]<0.05,1])/length(results_combinaison[,1])
-}
-
-resultats=cbind(combinaisons,gbj,gt,wt,gbt)
-
-resultats_total=resultats
-library(reshape2)
-library(ggplot2)
[…]

### 2024-11-18 reporting/master_draft/draft_sgbj.qmd
update content aspect

@@
        filename = here::here("reporting/master_draft/figures/plot_simulation_power_alpha.pdf"),
-       height = 10,
-       width = 8.5)
+       height = 7,
+       width = 8)
 ```
@@
   scale_fill_manual(values = c("#F48C06", "#DC2F02", "#9D0208")) +
-  labs(fill = "Grade", color = "Grade")
+  labs(fill = "Grade", color = "Grade", y ="Survival \n Probability")
 ```
@@
          name = factor(name,
-                       labels = c("Raw", "Benjamini-Hockberg"),
+                       labels = c("Raw", "Benjamini- \n Hockberg"),
                        levels = c("nb_sign", "nb_sign_bh"))) |> 
-  ggplot(mapping = aes(x = method, fill = method, y = value, alpha = name)) +
+  ggplot(mapping = aes(x = grade, fill = method, y = value)) +
   geom_bar(stat = "identity", position = position_dodge(), size = .6) +
-  facet_grid(grade ~ ., scales = "free_y") +
-  scale_fill_viridis_d(guide = "none") +
+  facet_grid(name ~ .) +
+  scale_fill_viridis_d() +
   scale_linetype_manual(values = c(1,2)) +
@@
         legend.position = "bottom") +
-  labs(x = "Method",
-       y = "Nb of significant pathways",
-       alpha = "")
+  guides(fill = guide_legend(nrow = 2)) +
+  labs(x = "",
+       y = "Number of significant \n pathways",
+       fill = "")
 ```
@@
 plot_upset_breast <- ggpubr::annotate_figure(plot_upset_breast,
-                                             left = "Nb significant pathways",
+                                             left = "Number of significant \n pathways",
                                              bottom = "Method")
@@
 plot_breast <- ggdraw() +
-  draw_plot(p_type, x = 0, y = .74, width = .4, height = .22) +
-  draw_plot(plot_breast_nb_genesets, x = 0, y = .24, width = .4, height = .5) +
-  draw_plot(plot_upset_breast, x = 0, y = 0, width = 1, height = 0.25) +
-  draw_plot(plot_pval_breast, x = .4, y = .25, width = .6, height = .75) +
+  draw_plot(p_type, x = 0, y = .75, width = .4, height = .22) +
+  draw_plot(plot_breast_nb_genesets, x = 0, y = .30, width = .4, height = .4) +
+  draw_plot(plot_upset_breast, x = 0, y = 0, width = 1, height = 0.28) +
+  draw_plot(plot_pval_breast, x = .4, y = .30, width = .6, height = .70) +
   draw_plot_label(label = c("A", "B", "C", "D"), size = 15,
-                  x = c(0, 0, 0, .4), y = c(1, .75, .26, 1))
+                  x = c(0, 0, 0, .4), y = c(1, .72, .28, 1))
 
@@
   scale_fill_manual(values = c("#2191FB", "#E94F37")) +
-  labs(fill = "", color = "")
+  labs(fill = "", color = "", y = "Survival \n Probability")
 ```
@@
          name = factor(name,
-                       labels = c("Raw", "Benjamini-Hockberg"),
+                       labels = c("Raw", "Benjamini- \n Hockberg"),
                        levels = c("nb_sign", "nb_sign_bh")),
@@
                                       "Astro" = "Astrocytoma")) |> 
-  ggplot(mapping = aes(x = method, fill = method, y = value, alpha = name)) +
+  ggplot(mapping = aes(x = grade, fill = method, y = value)) +
   geom_bar(stat = "identity", position = position_dodge(), size = .6) +
-  facet_grid(grade ~ ., scales = "free_y") +
-  scale_alpha_manual(values = c(0.5, 1)) +
-  scale_fill_viridis_d(guide = "none") +
+  facet_grid(name ~ .) +
+  scale_fill_viridis_d() +
   scale_linetype_manual(values = c(1,2)) +
@@
 plot_upset_rembrandt <- ggpubr::annotate_figure(plot_upset_rembrandt,
-                                                left = "Nb significant pathways",
+                                                left = "Number of significant \n pathways",
                                                 bottom = "Method")
@@
 plot_rembrandt <- ggdraw() +
-  draw_plot(p_type, x = 0, y = .74, width = .4, height = .22) +
-  draw_plot(plot_rembrandt_nb_genesets, x = 0, y = .24, width = .4, height = .5) +
-  draw_plot(plot_upset_rembrandt, x = 0, y = 0, width = 1, height = 0.25) +
-  draw_plot(plot_pval_rembrandt, x = .4, y = .25, width = .6, height = .75) +
+  draw_plot(p_type, x = 0, y = .75, width = .4, height = .22) +
+  draw_plot(plot_breast_nb_genesets, x = 0, y = .30, width = .4, height = .4) +
+  draw_plot(plot_upset_breast, x = 0, y = 0, width = 1, height = 0.28) +
+  draw_plot(plot_pval_breast, x = .4, y = .30, width = .6, height = .70) +
   draw_plot_label(label = c("A", "B", "C", "D"), size = 15,
-                  x = c(0, 0, 0, .4), y = c(1, .75, .26, 1))
+                  x = c(0, 0, 0, .4), y = c(1, .72, .28, 1))
 

### 2024-11-18 script/script_simulation/generate_figures/figures_simu_rembrandt_breast_v2.R

@@ -135,74 +135,87 @@ dfres_draft <- dfres %>%
          case_type = gsub(x = case_type, pattern = "Type : I", replacement = "Type : C"),
          case_type = gsub(x = case_type, pattern = "Case : 4", replacement = "Case : (I)"),
          case_type = gsub(x = case_type, pattern = "Case : 5", replacement = "Case : (II)"),
-         case_type = gsub(x = case_type, pattern = "Case : 6", replacement = "Case : (III)"))
+         case_type = gsub(x = case_type, pattern = "Case : 6", replacement = "Case : (III)"),
+         case = factor(case,
+                       levels = c(4, 5, 6),
+                       labels = c("Case : (I)", "Case : (II)", "Case : (III)")),
+         type = factor(type,
+                       levels = c("G", "H", "I", "Z"),
+                       labels = c("Type: A", "Type: B", "Type: C", "Type: Z")),
+         n_p = as.factor(n_p),
+         n_p = forcats::fct_relevel(n_p,
+                                    "N = 50 ; NG = 10",
+                                    "N = 50 ; NG = 50",
+                                    "N = 100 ; NG = 10",
+                                    "N = 100 ; NG = 50"))
 
 dfres_draft |> 
-  filter(type != "Z") |> 
+  filter(type != "Type: Z") |> 
   group_by(case_type, n_p) |> 
   summarise(power = max(power))
 
 plot_power_simu <- dfres_draft |> 
-  filter(type != "Z") %>%
+  filter(type != "Type: Z") %>%
   group_by(case_type, n_p) |> 
   mutate(max_power = max(power),
          rank = as.factor(dense_rank(-power))) |>
   ungroup() |> 
-  ggplot(mapping = aes(x = method,
-                       yintercept = max_power,
+  ggplot(mapping = aes(x = n_p,
                        y = power,
                        ymin = power_lower,
                        ymax = power_upper,
                        group = method,
-                       fill = method)) +
-  geom_bar(stat = "identity", width = 0.5, linewidth = 1) +
+                       color = method)) +
+  geom_point() +
+  geom_line(linewidth = 1) +
   geom_errorbar(width = .1) +
-  geom_hline(mapping = aes(yintercept = max_power), color = "grey") +
-  scale_fill_viridis_d() +
+  scale_color_viridis_d() +
   # scale_color_manual(values = c("red", "orange", "grey", "black")) +
[…]

### 2024-11-18 reporting/master_draft/draft_sgbj.qmd
fix typo

@@
   draw_plot(p_type, x = 0, y = .75, width = .4, height = .22) +
-  draw_plot(plot_breast_nb_genesets, x = 0, y = .30, width = .4, height = .4) +
-  draw_plot(plot_upset_breast, x = 0, y = 0, width = 1, height = 0.28) +
-  draw_plot(plot_pval_breast, x = .4, y = .30, width = .6, height = .70) +
+  draw_plot(plot_rembrandt_nb_genesets, x = 0, y = .30, width = .4, height = .4) +
+  draw_plot(plot_upset_rembrandt, x = 0, y = 0, width = 1, height = 0.28) +
+  draw_plot(plot_pval_rembrandt, x = .4, y = .30, width = .6, height = .70) +
   draw_plot_label(label = c("A", "B", "C", "D"), size = 15,

### 2024-12-13 reporting/master_draft/draft_sgbj.qmd
update plot boris comment

@@
       theme_minimal() +
-      theme(plot.margin = margin(10, 10, 10, 50)) +
+      theme(plot.margin = margin(0, 0, 0, 0)) +
       labs(title = title_i,
@@
   ggplot(mapping = aes(x = sGBJ_rank, y = p_value)) +
-  geom_point(mapping = aes(color = method)) +
+  geom_point(mapping = aes(color = method), alpha = .4) +
   scale_color_viridis_d(direction = -1) +
@@
        height = 9,
-       width = 7)
+       width = 8)
 
@@
       theme_minimal() +
-      theme(plot.margin = margin(10, 10, 10, 50)) +
+      theme(plot.margin =  margin(0, 0, 0, 0)) +
       labs(title = title_i,

### 2024-12-16 reporting/master_draft/draft_sgbj.qmd
update report

@@
   scale_fill_manual(values = c("#F48C06", "#DC2F02", "#9D0208")) +
-  labs(fill = "Grade", color = "Grade", y ="Survival \n Probability")
+  labs(fill = "Grade", color = "Grade", y ="Metastasis free \n survival probability")
 ```

### 2024-12-19 reporting/master_draft/draft_sgbj.qmd
update report

@@
 
+load(here::here("data/breast_cancer/BreastCancer-Vijver-Clinical.RData"))
+
 ls_pathways <- lsdata$pathways
@@
   gtsummary::tbl_summary()
+
+clin.data |> 
+  mutate(event = factor(evmeta+evdeath>0,
+                        levels = c(T,F),
+                        labels = c("Yes", "No"))) |> 
+  dplyr::select(Age = age,
+                "Composite event" = event,
+                "Follow-up" = tsurv,
+                "Grade" = grade) |> 
+    gtsummary::tbl_summary()
+
 ```
@@
         legend.box="vertical") +
-  labs(x = "SGBJ rank", y = "Raw p-value")
+  labs(x = "sGBJ rank", y = "Raw p-value")
 ```

### 2025-03-12 script/rembrandt/datamanagement/datamanage_rembrandt.R
add info rembrandt

@@ -4,6 +4,8 @@ library(hgu133plus2.db)
 
 ################################################# import data ####################################
 
+# raw data are available at: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE108474
+
 ##### clinical data
 GSE108474_REMBRANDT_Clinical <- read.table(file = "GSE108474_REMBRANDT_clinical.data.txt",
                                                  sep = "\t",

### 2025-03-12 script/breast_cancer/00_get_data_breast_cancer.R
remove useless scripts

@@ -1,0 +1,79 @@
+library(tidyverse)
+library(breastCancerNKI)
+library(msigdbr)
+
+set.seed(1)
+
+data(nki)
+
+##### Phenotype data #####
+phenotype_data <- pData(nki) |> 
+  as_tibble() |> 
+  select("ID" = "samplename",
+         "evdeath" = "e.os",
+         "tsurv" = "t.os",
+         "evmeta" = "e.dmfs",
+         "tmeta" = "t.dmfs",
+         "age" = "age",
+         "grade" = "grade") |> 
+  mutate(event = as.numeric(evdeath == 1 | evmeta == 1),
+         time = tmeta/365.25,
+         time_follow_up = if_else(tsurv > tmeta, tsurv, tmeta)/365.25) %>%
+  select(ID, grade, event, time, time_follow_up, age) |> 
+  na.omit()
+
+##### Pathway data #####
+
+pathway_gene <- msigdbr(
+  species = "Homo sapiens", # For human pathways
+  category = "C2",         # C2 category
+  subcategory = "KEGG"  # Canonical Pathways, KEGG
+) |> 
+  select(gene_symbol, gs_name) |> 
+  as_tibble()
+
+##### Pathway - probe corresponding #####
+
+gene_probe <- fData(nki) |> 
+  select(probe, gene_symbol = NCBI.gene.symbol) |> 
+  na.omit() |> 
+  as_tibble()
+
+##### Expression data #####
+raw_expr <- exprs(nki)
+
+### remove > 10% missing
+
+bool_probes <- raw_expr |> 
+  apply(MARGIN = 1, FUN = function(x) mean(is.na(x))<=0.1) 
+
+### impute by knn similar to https://doi.org/10.1371/journal.pcbi.1004791 
+
+expr_no_missing <- impute::impute.knn(raw_expr[bool_probes,], rng.seed = 123)$data
+
+### to dataframe format
+
+expression_data <- expr_no_missing |> 
+  t() |> 
+  as.data.frame() |> 
+  tibble::rownames_to_column(var = "ID") |> 
+  as_tibble() |> 
+  tidyr::pivot_longer(cols = -ID,
+                      names_to = "probe",
+                      values_to = "expression") |> 
+  inner_join(gene_probe, by = "probe") |> 
+  filter(gene_symbol %in% unique(pathway_gene$gene_symbol)) |> 
+  group_by(ID, gene_symbol) |> 
[…]

### 2025-03-12 script/breast_cancer/01_datamanagement.R

@@ -1,32 +1,0 @@
-##### load packages #####
-library(dplyr)
-
-# add method to grid.draw
-grid.draw.ggsurvplot <- function(x){
-  survminer:::print.ggsurvplot(x, newpage = FALSE)
-}
-
-##### load data #####
-load("data/breast_cancer/BreastCancer-Vijver-Clinical.RData")
-load("data/breast_cancer/BreastCancer-Vijver.RData")
-load("data/breast_cancer/c2.cp.v2.5.GS.RData")
-
-##### data management #####
-
-### clinical data
-df_clinical <- clin.data %>%
-  mutate(event = as.numeric(evdeath == 1 | evmeta == 1),
-         time = if_else(tsurv > tmeta, tmeta, tsurv)) %>%
-  select(ID, grade, event, time, age)
-
-### add counts
-counts=t(llid_norm.mat)
-
-df_all <- df_clinical %>%
-  bind_cols(counts) %>%
-  as_tibble()
-
-### pathways
-saveRDS(object = list(data = df_all,
-                      pathways = GS.llids),
-        file = "data/breast_cancer/datamanaged.rds")

### 2025-03-12 script/breast_cancer/02_survival_analysis.R

@@ -1,40 +1,0 @@
-##### load packages #####
-library(dplyr)
-library(ggplot2)
-library(survival)
-library(survminer)
-
-##### survival analysis #####
-lsdata <- readRDS(file = "data/breast_cancer/datamanaged.rds")
-
-### marginal
-surv_marginal <- survfit(Surv(time = time, event = event) ~ 1,
-                         data = lsdata$data)
-
-plot_marginal_km <- ggsurvplot(surv_marginal,
-                               risk.table = TRUE,
-                               conf.int = TRUE,
-                               ggtheme = theme_bw(),
-                               palette = "black")
-
-ggsave(filename = "images/plot_marginal_km.pdf",
-       plot = plot_marginal_km,
-       device = "pdf",
-       height = 6,
-       width = 6)
-
-### stratified 
-surv_grade <- survfit(Surv(time = time, event = event) ~ grade,
-                      data = lsdata$data)
-
-plot_grade_km <- ggsurvplot(surv_grade,
-                            risk.table = TRUE,
-                            conf.int = TRUE,
-                            palette = c("#FFCD4D", "#F48C06", "#9D0208"),
-                            ggtheme = theme_bw())
-
-ggsave(filename = "images/plot_grade_km.pdf",
-       plot = plot_grade_km,
-       device = "pdf",
-       height = 6,
-       width = 6)

### 2025-03-12 script/breast_cancer/03_pathway_analysis.R

@@ -1,79 +1,0 @@
-##### load packages #####
-library(dplyr)
-library(ggplot2)
-library(survival)
-library(survminer)
-library(parallel)
-
-##### survival analysis #####
-lsdata <- readRDS(file = "data/breast_cancer/datamanaged.rds")
-source("script/breast_cancer/liste70pathways.R")
-
-##### marginal #####
-df_res_marginal <- mclapply(X = liste_70,
-                            mc.cores = 19,
-                            FUN = function(pathway_i){
-                              vec_pathways <- lsdata$pathways[[pathway_i]]
-                              
-                              surv_obj <- survival::Surv(time = lsdata$data$time,
-                                                         event = lsdata$data$event)
-                              
-                              factor_matrix <- lsdata$data %>%
-                                select(any_of(vec_pathways))
-                              
-                              covariates <- lsdata$data %>% select(age)
-                              
-                              test <- sGBJ::sGBJ(surv = surv_obj,
-                                                 factor_matrix = factor_matrix,
-                                                 covariates = covariates)
-                              
-                              res <- data.frame(pathway = pathway_i,
-                                                nb_genes = ncol(factor_matrix),
-                                                p_value = test$sGBJ_pvalue)
-                              
-                              return(res)
-                            }) %>%
-  bind_rows() %>%
-  mutate(p_value_adjusted = p.adjust(p = p_value, method = "hochberg"))
-
-saveRDS(df_res_marginal, file = "data/breast_cancer/df_res_marginal.rds")
-
-##### stratified #####
-df_res_stratified <- mclapply(X = liste_70,
-                              mc.cores = 19,
-                              FUN = function(pathway_i){
-                                vec_pathways <- lsdata$pathways[[pathway_i]]
-                                vec_grade <- unique(lsdata$data$grade)
-                                
-                                res <- lapply(X = vec_grade,
-                                              FUN = function(grade_i){
-                                                dfdata <- lsdata$data %>% filter(grade == grade_i)
-                                                
-                                                surv_obj <- survival::Surv(time = dfdata$time,
-                                                                           event = dfdata$event)
-                                                
-                                                factor_matrix <- dfdata %>%
-                                                  select(any_of(vec_pathways))
-                                                
-                                                covariates <- dfdata %>% select(age)
-                                                
-                                                test <- sGBJ::sGBJ(surv = surv_obj,
-                                                                   factor_matrix = factor_matrix,
-                                                                   covariates = covariates)
-                                                
[…]

### 2025-03-12 script/breast_cancer/04_all_pathway_analysis.R

@@ -1,88 +1,0 @@
-##### load packages #####
-library(dplyr)
-library(ggplot2)
-library(survival)
-library(survminer)
-library(parallel)
-
-##### survival analysis #####
-lsdata <- readRDS(file = "data/breast_cancer/datamanaged.rds")
-ls_pathways <- lsdata$pathways
-vec_grade <- unique(lsdata$data$grade)
-nperm <- 1000
-
-##### stratified #####
-# vec_pathways <- ls_pathways[[1]]
-# grade_i <- vec_grade[1]
-# nperm <- 10
-
-df_res_stratified <- mclapply(X = ls_pathways,
-                              mc.cores = 19,
-                              FUN = function(vec_pathways){
-                                res <- lapply(X = vec_grade,
-                                              FUN = function(grade_i){
-                                                ##### PREP DATA
-                                                dfdata <- lsdata$data %>% filter(grade == grade_i)
-                                                
-                                                surv_obj <- survival::Surv(time = dfdata$time,
-                                                                           event = dfdata$event)
-                                                
-                                                factor_matrix <- dfdata %>%
-                                                  select(any_of(vec_pathways))
-                                                
-                                                covariates <- dfdata %>% select(age)
-                                                
-                                                all_expl_features <- bind_cols(factor_matrix, covariates)
-                                                
-                                                alternative_hyp <- paste0("~ ",
-                                                                          paste0("`", colnames(factor_matrix), "`", collapse = " + "))
-                                                null_hyp <- paste0("~ ",
-                                                                   paste0(colnames(covariates), collapse = " + "))
-                                                
-                                                ##### TESTS
-                                                ### sGBJ
-                                                sGBJ <- sGBJ::sGBJ(surv = surv_obj,
-                                                                   factor_matrix = factor_matrix,
-                                                                   covariates = covariates,
-                                                                   nperm = nperm)$sGBJ_pvalue
-                                                ### GBT
-                                                GBT <- globalboosttest::globalboosttest(
-                                                  X = factor_matrix,
-                                                  Y = surv_obj,
-                                                  Z = covariates,
-                                                  nperm = nperm,
-                                                  mstop = 500,
-                                                  pvalueonly = TRUE
-                                                )$pvalue |> 
-                                                  as.numeric()
-                                                ### WALD
-                                                wald_pval <- Wald_Test_Perm(vectime = dfdata$time,
-                                                                            vecevent = dfdata$event,
-                                                                            x = dfdata %>% select(-time,-event) %>% as.matrix(),
[…]

### 2025-03-12 reporting/breast_cancer/breast_description.Rmd
adapt old scripts

--- 2024-07-01 script/breast_cancer/breast_description.Rmd
@@
 
-lsdata <- readRDS(file = here::here("data/breast_cancer/datamanaged.rds"))
+lsdata <- readRDS(file = here::here("data/breast_cancer/datamanaged2.rds"))
 
@@
 
-ls_pathways <- lsdata$pathways
+df_pathways <- lsdata$pathways
 df_data <- lsdata$data
@@
 ```{r ecdf, fig.cap="Empirical Cumulative Distribution Function of number of genes by pathway. Breast cancer study."}
-lapply(ls_pathways,
-       function(pathway_i) sum(pathway_i %in% colnames(df_data))) |> 
-  bind_rows() |> 
-  tidyr::pivot_longer(cols = everything(),
-                      names_to = "pathway",
-                      values_to = "nb_genes") |> 
+df_pathways |> 
+  group_by(gene_symbol) |> 
+  summarise(nb_genes = n()) |> 
   ggplot(mapping = aes(x = nb_genes)) +
@@
        y = "Empirical Cumulative Distribution")
-
 ```

### 2025-03-12 script/breast_cancer/05_master_analysis_breast.R

@@ -9,16 +9,25 @@ library(survminer)
 library(parallel)
 
 ##### survival analysis #####
-lsdata <- readRDS(file = "data/breast_cancer/datamanaged.rds")
+lsdata <- readRDS(file = "data/breast_cancer/datamanaged2.rds")
 
 lapply(X = list.files("functions/", full.names = T), source)
 
 ls_pathways <- lsdata$pathways
 df_data <- lsdata$data
 
+##### get pathway
+vec_pathway <- unique(lsdata$pathways$gs_name)
+pathway_taskid <- vec_pathway[taskid]
+gene_taskid <- lsdata$pathways |> 
+  filter(gs_name == pathway_taskid) |> 
+  pull(gene_symbol)
+
+ls_pathway_taskid <- setNames(list(gene_taskid), pathway_taskid)
+
 ##### analysis
 df_analysis_breast <- fct_breast_cancer_analysis(df_data = df_data,
-                                                 ls_pathways = ls_pathways[taskid],
+                                                 ls_pathways = ls_pathway_taskid,
                                                  vec_covariates = "age",
                                                  nb_permutation = 1000)
 

### 2025-03-12 script/breast_cancer/liste70pathways.R
reorganise breast files

@@ -1,72 +1,0 @@
-liste_70 = c(
-  "AKTPATHWAY",
-  "ARFPATHWAY",
-  "BADPATHWAY",
-  "ERKPATHWAY",
-  "MPRPATHWAY",
-  "P53PATHWAY",
-  "RASPATHWAY",
-  "MAPKPATHWAY",
-  "MTA3PATHWAY",
-  "MTORPATHWAY",
-  "NFKBPATHWAY",
-  "PTENPATHWAY",
-  "TGFBPATHWAY",
-  "VEGFPATHWAY",
-  "PELP1PATHWAY",
-  "TNFR1PATHWAY",
-  "AKAP13PATHWAY",
-  "STRESSPATHWAY",
-  "WNT_SIGNALING",
-  "APOPTOSIS_KEGG",
-  "ATRBRCAPATHWAY",
-  "EGFR_SMRTEPATHWAY",
-  "HSA03010_RIBOSOME",
-  "HSA03050_PROTEASOME",
-  "HSA03020_RNA_POLYMERASE",
-  "HSA03030_DNA_POLYMERASE",
-  "HSA03060_PROTEIN_EXPORT" ,
-  "HSA00780_BIOTIN_METABOLISM",
-  "HSA00920_SULFUR_METABOLISM",
-  "HSA00830_RETINOL_METABOLISM",
-  "HSA01430_CELL_COMMUNICATION",
-  "HSA00730_THIAMINE_METABOLISM",
-  "HSA00790_FOLATE_BIOSYNTHESIS",
-  "HSA00910_NITROGEN_METABOLISM",
-  "HSA00791_ATRAZINE_DEGRADATION",
-  "HSA04115_P53_SIGNALING_PATHWAY",
-  "HSA04310_WNT_SIGNALING_PATHWAY",
-  "HSA00740_RIBOFLAVIN_METABOLISM",
-  "HSA00750_VITAMIN_B6_METABOLISM",
-  "HSA04010_MAPK_SIGNALING_PATHWAY",
-  "HSA04012_ERBB_SIGNALING_PATHWAY",
-  "HSA04150_MTOR_SIGNALING_PATHWAY" ,
-  "HSA04370_VEGF_SIGNALING_PATHWAY",
-  "HSA00785_LIPOIC_ACID_METABOLISM",
-  "HSA00900_TERPENOID_BIOSYNTHESIS",
[…]

### 2025-03-12 script/breast_cancer/01_master_analysis_breast.R
slight improvement

@@ -1,12 +1,9 @@
-taskid <- Sys.getenv('SLURM_ARRAY_TASK_ID')
+taskid <- as.numeric(Sys.getenv('SLURM_ARRAY_TASK_ID'))
 print(taskid)
 ######## PATWAY ANALYSIS OF BREAST CANCER FILE #########
 ##### load packages #####
 library(dplyr)
-library(ggplot2)
 library(survival)
-library(survminer)
-library(parallel)
 
 ##### survival analysis #####
 lsdata <- readRDS(file = "data/breast_cancer/datamanaged2.rds")

### 2025-03-12 reporting/master_draft/draft_sgbj.qmd
update results with clean online data

@@
 ```{r}
-lsdata <- readRDS(file = here::here("data/breast_cancer/datamanaged.rds"))
-
-load(here::here("data/breast_cancer/BreastCancer-Vijver-Clinical.RData"))
+lsdata <- readRDS(file = here::here("data/breast_cancer/datamanaged2.rds"))
 
@@
                 event,
-                time,
+                time_follow_up,
                 grade) |> 
-  gtsummary::tbl_summary()
-
-clin.data |> 
-  mutate(event = factor(evmeta+evdeath>0,
-                        levels = c(T,F),
-                        labels = c("Yes", "No"))) |> 
-  dplyr::select(Age = age,
-                "Composite event" = event,
-                "Follow-up" = tsurv,
-                "Grade" = grade) |> 
   gtsummary::tbl_summary()

### 2025-12-10 functions/statsTest.R
update simulations for reviewers

@@ -42,31 +42,20 @@ statsTest=function(vectime, vecevent, x){
 #' @param vectime the time
 #' @param vecevent the event
 #' @param x the genes experession
-#' @param covariates a dataframe of covariates (default is NULL)
 #'
 #' @return A vector with the statistical test of global test, adewale test
 #' @export
 #'
-pre_Wald_Test=function(vectime, vecevent, x, covariates = NULL){
+pre_Wald_Test=function(vectime, vecevent, x){
   # baseline
   survObj <- survival::Surv(vectime,vecevent)
-  if(is.null(covariates)){
-    ucox <- lapply(X = 1:ncol(x),
-                   FUN = function(i){
-                     ocx<-survival::coxph(survObj~x[,i])
-                     res<-ocx$coefficients/sqrt(ocx$var)
-                     return(res)
-                   }) %>%
-      unlist()
-  } else {
-    ucox <- lapply(X = 1:ncol(x),
-                   FUN = function(i){
-                     ocx<-survival::coxph(survObj~x[,i]+covariates)
-                     res<-ocx$coefficients[1]/sqrt(ocx$var[1])
-                     return(res)
-                   }) %>%
-      unlist()
-  }
+  ucox <- lapply(X = 1:ncol(x),
+                 FUN = function(i){
+                   ocx<-survival::coxph(survObj~x[,i])
+                   res<-ocx$coefficients/sqrt(ocx$var)
+                   return(res)
+                 }) %>%
+    unlist()
   
   res <- sum(ucox^2)
   
@@ -85,7 +74,7 @@ pre_Wald_Test=function(vectime, vecevent, x, covariates = NULL){
 #' @return A vector with the statistical test of global test, adewale test
 #' @export
 #'
-Wald_Test_Perm=function(vectime, vecevent, x, covariates, nb_permutation){
+Wald_Test_Perm=function(vectime, vecevent, x, nb_permutation){
   # baseline
   baseline_ucox <- pre_Wald_Test(vectime = vectime,
                                  vecevent = vecevent,

### 2025-12-10 script/script_simulation/curta/08_df_rev_resp_round_1.R

@@ -25,13 +25,13 @@ message("start computation")
 res <- pbapply::pbapply(X = dfScenario,
                         MARGIN = 1,
                         FUN = function(row){
+                          print(row[["hp_row"]])
                           
                           if(as.character(row[["method"]]) == "All"){
                             vec_methods = c("sgbj", "gbt", "gt", "wald", "Cauchy", "HM")
-                          } else if (as.character(row[["method"]]) == "SGBJ"){
-                            vec_methods = "sgbj"
                           } else {
-                            stop("method unknown")
+                            vec_methods = strsplit(as.character(row[["method"]]), split = ";") |> 
+                              unlist()
                           }
                           
                           fct_simulation_paper(methods = vec_methods,

### 2025-12-10 script/script_simulation/curta/09_df_rev_resp_round_1_Z.R

@@ -26,12 +26,13 @@ res <- pbapply::pbapply(X = dfScenario,
                         MARGIN = 1,
                         FUN = function(row){
                           
+                          print(row[["hp_row"]])
+                          
                           if(as.character(row[["method"]]) == "All"){
                             vec_methods = c("sgbj", "gbt", "gt", "wald", "Cauchy", "HM")
-                          } else if (as.character(row[["method"]]) == "SGBJ"){
-                            vec_methods = "sgbj"
                           } else {
-                            stop("method unknown")
+                            vec_methods = strsplit(as.character(row[["method"]]), split = ";") |> 
+                              unlist()
                           }
                           
                           fct_simulation_paper(methods = vec_methods,
@@ -42,9 +43,9 @@ res <- pbapply::pbapply(X = dfScenario,
                                                case =  row[["case"]],
                                                type =  row[["type"]],
                                                censoring =  as.numeric(row[["censoring"]]),
-                                               nb_permutation =  as.numeric(row[["nb_permutation"]]),
+                                               nb_permutation = 25,
                                                prop_two_periods =  row[["prop_two_periods"]],
-                                               nperm_sGBJ = as.numeric(row[["nperm_sGBJ"]])) %>%
+                                               nperm_sGBJ = 25) %>%
                             mutate(hp_row = row[["hp_row"]],
                                    iter = slar_taskid) %>%
                             return()

### 2025-12-10 script/script_simulation/generate_scenarios/resp_reviewer_1round.R

@@ -1,46 +1,57 @@
-
+library(dplyr)
 # Power -------------------------------------------------------------------
 
-## launch with 1000 repetitions (500 done before but I wrote 1,000 ...)
-
-df_rev1 <- expand.grid(nb_observations = c(50,100),
-                       nb_genes = c(10, 50, 100),
-                       prop_sig_gene = c(0.2, 0.05),
-                       variance = 0.2,
-                       case = c(4, 5, 6),
-                       type = c("G", "H", "I"),
-                       censoring = c(0.3),
-                       nb_permutation = 1000,
-                       nperm_sGBJ = 1000,
-                       prop_two_periods = FALSE,
-                       method = "All")
-## high gene correlation
-df_rev2 <- expand.grid(nb_observations = c(50,100),
-                       nb_genes = c(10, 50, 100),
-                       prop_sig_gene = c(0.2),
-                       variance = 0.2,
-                       case = c(7),
-                       type = c("G"),
-                       censoring = c(0.3),
-                       nb_permutation = 1000,
-                       nperm_sGBJ = 1000,
-                       prop_two_periods = FALSE,
-                       method = "All")
-
-### vary sGBJ permutations number
-df_rev3 <- expand.grid(nb_observations = c(50,100),
-                       nb_genes = c(10, 50, 100),
-                       prop_sig_gene = c(0.2),
-                       variance = 0.2,
-                       case = c(4),
-                       type = c("G"),
-                       censoring = c(0.3),
-                       nb_permutation = 1000,
-                       nperm_sGBJ = c(500, 2000),
-                       prop_two_periods = FALSE,
-                       method = "SGBJ")
-
-df_rev_resp_round_1 <- dplyr::bind_rows(df_rev1, df_rev2, df_rev3)
+df_rev_resp_round_1 <- list(
+  ## Add Cauchy + HM
+  expand.grid(nb_observations = c(50,100),
+              nb_genes = c(10, 50),
+              prop_sig_gene = c(0.2),
+              variance = 0.2,
+              case = c(4, 5, 6),
+              type = c("G", "H", "I"),
+              censoring = c(0.3),
+              nb_permutation = 1000,
+              nperm_sGBJ = 1000,
+              prop_two_periods = FALSE,
+              method = "Cauchy;HM"),
[…]

### 2025-12-10 script/script_simulation/curta/09_df_rev_resp_round_1_Z.R
bug fix nb permutation

@@ -43,9 +43,9 @@ res <- pbapply::pbapply(X = dfScenario,
                                                case =  row[["case"]],
                                                type =  row[["type"]],
                                                censoring =  as.numeric(row[["censoring"]]),
-                                               nb_permutation = 25,
+                                               nb_permutation =  as.numeric(row[["nb_permutation"]]),
                                                prop_two_periods =  row[["prop_two_periods"]],
-                                               nperm_sGBJ = 25) %>%
+                                               nperm_sGBJ = as.numeric(row[["nperm_sGBJ"]])) %>%
                             mutate(hp_row = row[["hp_row"]],
                                    iter = slar_taskid) %>%
                             return()

### 2025-12-10 script/script_simulation/curta/08_df_rev_resp_round_1.R
fix last typo

@@ -42,7 +42,7 @@ res <- pbapply::pbapply(X = dfScenario,
                                                case =  row[["case"]],
                                                type =  row[["type"]],
                                                censoring =  as.numeric(row[["censoring"]]),
-                                               nb_permutation =  as.numeric(row[["nb_permutation"]]),
+                                               nb_permutation = as.numeric(row[["nb_permutation"]]),
                                                prop_two_periods =  row[["prop_two_periods"]],
                                                nperm_sGBJ = as.numeric(row[["nperm_sGBJ"]])) %>%
                             mutate(hp_row = row[["hp_row"]],

### 2025-12-10 script/script_simulation/curta/09_df_rev_resp_round_1_Z.R

@@ -43,7 +43,7 @@ res <- pbapply::pbapply(X = dfScenario,
                                                case =  row[["case"]],
                                                type =  row[["type"]],
                                                censoring =  as.numeric(row[["censoring"]]),
-                                               nb_permutation =  as.numeric(row[["nb_permutation"]]),
+                                               nb_permutation = as.numeric(row[["nb_permutation"]]),
                                                prop_two_periods =  row[["prop_two_periods"]],
                                                nperm_sGBJ = as.numeric(row[["nperm_sGBJ"]])) %>%
                             mutate(hp_row = row[["hp_row"]],

### 2025-12-10 functions/fct_perform_tests.R
fix covariates wald

@@ -48,7 +48,6 @@ fct_perform_tests <- function(df_data_analysis,
   wald_pval <- Wald_Test_Perm(vectime = df_data_analysis$time,
                               vecevent = df_data_analysis$event,
                               x = as.matrix(factor_matrix),
-                              covariates = as.matrix(covariates),
                               nb_permutation = nb_permutation)
   time_end <- Sys.time()
   time_wald <- difftime(time_end, time_start, units = "secs")

### 2025-12-10 reporting/supplementary_sgbj/simu_rembrandt_breast.Rmd
prepare supplementary material

---
title: "Gene Set Analysis for time-to-event outcome:comparison of a new approach based on theGeneralized Berk–Jones statistic with existingmethods in presence of intra gene-set correlation"
subtitle: "Supplementary"
author: ""
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: True
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(dplyr)
library(ggplot2)

source(file = here::here("functions/fct_report_simu_setting.R"))

dfRembrandt <- readRDS(here::here("data/rembrandt/dfresvarcovar.rds"))
dfBreast <- readRDS(here::here("data/breast_cancer/dfresvarcovar.rds"))
```

# Rembrandt and Breast Cancer Data for Simulation Settings

This section aims to identify realistic simulation settings using the Rembrandt and Breast Cancer datasets.

## Proportion of Significant Genes

As shown in Figure \@ref(fig:propsigngenes), the proportion of significant genes is approximately 45% for the Rembrandt dataset and 20% for the Breast Cancer dataset.

```{r propsigngenes, fig.cap="Proportion of significant genes in each pathway for each data source.", fig.height=4}
dplyr::bind_rows(dfRembrandt$dfCoxVariance |> mutate(data = "Rembrandt"),
                 dfBreast$dfCoxVariance |> mutate(data = "Breast")) |> 
  group_by(pathway, data) |> 
  summarise(nb_genes = n(),
            nb_sign_genes = sum(SIGN),
            prop_sign_genes = nb_sign_genes/nb_genes,
            .groups = "drop") |> 
  ggplot(mapping = aes(y = prop_sign_genes, x = data)) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, 1, by = .2)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "Data source",
       y = "Proportion of significant genes")

```

## Gene expression variance

Gene expression variance was studied in each data source.
It was of approximately $0.3$ and $0.09$ in Rembrandt and Breast cancer data respectively.
For simulations, we use an intermediate variance of 0.2 (Figure \@ref(fig:genevariance)).

```{r genevariance, fig.cap="Gene expression variance for each data source.", fig.height=4}
dplyr::bind_rows(dfRembrandt$dfCoxVariance |> mutate(data = "Rembrandt"),
                 dfBreast$dfCoxVariance |> mutate(data = "Breast")) |> 
  select(data, GENE, VARIANCE) |> 
  distinct() |> 
  ggplot(mapping = aes(y = VARIANCE, x = data)) +
  geom_boxplot() +
  scale_y_log10(guide = "axis_logticks") +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "Data source",
       y = "Gene Expression Variance")

```

## Gene expression correlation

Gene expression correlation was evaluated on each pathway of each data source.
Correlation was evaluated overall and stratified by gene significance.
Figure \@ref(fig:overallgenecorrelation) indicates that an overall gene correlation following a non-standard beta distribution(20, 20, min = -1, max = 1) is realistic for both Rembrandt and Breast cancer data. This is denoted as case I.
When stratified by gene significance, figure \@ref(fig:genecorrelationbysign) outlines that a non-standard beta distribution(10, 10, min = -1, max = 1) for significant genes and non-standard beta distribution(25, 25, min = -1, max = 1) for non-significant genes is realistic. This is denoted as case II.

A simpler simulation setting considering 0.2 correlation between significant genes and 0 correlation between non-significant genes was also considered.

```{r overallgenecorrelation, fig.cap="Gene expression correlation for each pathway in each data source.", fig.height=5}
ls_R <- fct_plot_correlation(dfRembrandt, caption = "Rembrandt : Shapes of non-standard beta distribution by type of gene")
ls_B <- fct_plot_correlation(dfBreast, caption = "Breast : Shapes of non-standard beta distribution by type of gene")

plot_R <- ls_R$plot_unique + ggtitle("Rembrandt")
plot_B <- ls_B$plot_unique + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

```{r genecorrelationbysign, fig.cap="Gene expression correlation for each pathway in each data source stratified by gene significance.", fig.height=6}
plot_R <- ls_R$plot_multiple + ggtitle("Rembrandt")
plot_B <- ls_B$plot_multiple + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

## Beta from Cox model

```{r}
lsbeta_R <- fct_beta_dist(dfRembrandt, caption = "Rembrandt : Proportion of positive, negative and non significant genes")
lsbeta_B <- fct_beta_dist(dfBreast, caption = "Breast cancer : Proportion of positive, negative and non significant genes")
```

Beta coefficient distributions, both overall and stratified by gene significance, were examined in the two datasets.
Figure \@ref(fig:betacoef) indicates that the overall distribution of effect sizes is well approximated by a normal distribution with mean 0 and standard deviation 0.4; we refer to this setting as Type A.
Figure \@ref(fig:betacoefstrat) further shows that, in the Rembrandt dataset, half of the gene effects follow a normal distribution with mean -0.4 and standard deviation 0.2, while the other half follow a normal distribution with mean 0.4 and standard deviation 0.2; we denote this as Type B.
In contrast, the Breast Cancer dataset is characterized by a wider separation, with half of the effects following a normal distribution with mean -0.8 and standard deviation 0.4, and the other half following a normal distribution with mean 0.8 and standard deviation 0.4; this is referred to as Type C.

```{r betacoef, fig.cap="Beta coefficient distribution for each data source.", fig.height=5}
plot_R <- lsbeta_R$plot_unique_beta + ggtitle("Rembrandt")
plot_B <- lsbeta_B$plot_unique_beta + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

```{r betacoefstrat, fig.cap="Beta coefficient distribution for each data source stratified by gene significance.", fig.height=6}
plot_R <- lsbeta_R$plot_multiple_beta + ggtitle("Rembrandt")
plot_B <- lsbeta_B$plot_multiple_beta + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```
